---
title: "Forecasting epidemic trajectories: Time Series Growth Curves package `tsgc`"
author: "Michael Ashby, Andrew Harvey, Paul Kattuman, Edwin Tang, Craig Thamotheram"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Forecasting epidemic trajectories: Time Series Growth Curves package `tsgc`}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 7,
                      fig.height = 7)

library(tsgc)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(zoo)
theme_set(theme_economist_white(gray_bg = FALSE, base_size = 16))
```

# Converting `xts`/`Date`-indexed data to `idx_series`/`idx_calendar`

The package's example datasets (`gauteng`, `england`, `ukitaly`,
`nintendo_sales`, `etrading_apps`, and the weather datasets) still ship
as `xts` objects with a `Date`/`yearqtr`/`yearmon` index, since that is
how they were originally collected. The modelling code in this version
of the package, however, works with `idx_series` (integer-position data)
and an optional `idx_calendar` (the translation layer back to real
calendar time, used only for plotting). This section shows the one-off
conversion step; the same pattern applies to any user's own `xts`,
`zoo`, or `data.frame` data.

The package provides `xts_to_idx()` and `idx_to_pos()` for this: a small
helper that wraps the conversion for the common case of a daily series
with no gaps (every calendar day present, i.e. `pattern = 1`), and its
inverse for translating a calendar date back to an integer position.

## Gauteng daily cases

```{r convert-gauteng}
data(gauteng, package = "tsgc")

conv <- xts_to_idx(gauteng)
gauteng_idx <- conv$series
gauteng_cal <- conv$calendar
```

## England daily cases/hospitalisations and weather

```{r convert-england}
data(england, package = "tsgc")
data(england_weather_2021, package = "tsgc")

conv <- xts_to_idx(england[, 1:2])
england_idx <- conv$series
england_cal <- conv$calendar

conv <- xts_to_idx(england_weather_2021[, 1:4],
                    start.pos = idx_to_pos(england_cal, zoo::index(england_weather_2021)[1]))
england_weather_idx <- conv$series
```

## UK/Italy daily cases

```{r convert-ukitaly}
data(ukitaly, package = "tsgc")

conv <- xts_to_idx(ukitaly)
ukitaly_idx <- conv$series
ukitaly_cal <- conv$calendar
```

## Quarterly Nintendo console sales

Quarterly data is not daily, so the calendar here uses `unit =
"quarters"` and an anchor built from the first date in the series
(coerced to a `Date`). `amount = 1` since every quarter is present
(`pattern = 1`). With `posixct = TRUE`, `idx_to_date()` steps by whole
quarters using calendar-aware arithmetic, so no manual `yearqtr`
lookup is needed.

```{r convert-nintendo}
data(nintendo_sales, package = "tsgc")

y_q_xts <- nintendo_sales[, c("wii", "switch_all")]

nintendo_idx <- idx_series(zoo::coredata(y_q_xts), start = 1L)
nintendo_cal <- idx_calendar(
  anchor = as.Date(zoo::index(y_q_xts)[1]),
  anchor_pos = 1L,
  amount = 1, unit = "quarters",
  posixct = TRUE
)
```

## Monthly e-trading app downloads

```{r convert-etrading}
data(etrading_apps, package = "tsgc")

y_m_xts <- etrading_apps[, c("DEGIRO", "AvaTrade")]

etrading_idx <- idx_series(zoo::coredata(y_m_xts), start = 1L)
etrading_cal <- idx_calendar(
  anchor = as.Date(zoo::index(y_m_xts)[1]),
  anchor_pos = 1L,
  amount = 1, unit = "months",
  posixct = TRUE
)
```

## Working without any calendar dates

Suppose the daily Gauteng series above had no dates attached at all -
just an ordered vector of daily case counts, with day 1 being some
arbitrary reference point (e.g. days since the first recorded case).
`idx_series` positions are already plain integers, so no `idx_calendar`
is required to run the model: `start`/`end` can be passed directly as
integer positions.

```{r no-dates-plain}
gauteng_nodate_idx <- idx_series(zoo::coredata(gauteng), start = 1L)

# start/end are just integer positions - day 45 to day 120 since day 1,
# with no idx_calendar involved at all.
mod_nodate <- SSModelDynamicGompertz(Y = gauteng_nodate_idx, start = 45, end = 120)
res_nodate <- estimate(mod_nodate)
```

With no `calendar` supplied, the x-axis of every plot falls back to raw
integer positions instead of dates - this is what `tsgc` plots look
like for a series with no calendar meaning at all.

```{r fig-nodate-log, fig.cap = "Log forecast with no calendar attached: the x-axis is a plain integer position."}
tsgc::plot_log_forecast(
  res = res_nodate,
  Y = gauteng_nodate_idx,
  n.ahead = 14,
  plt.start = tail(res_nodate$index, 1) - 30
)
```

```{r fig-nodate-level, fig.cap = "Forecast of new cases with no calendar attached."}
tsgc::plot_forecast(
  res = res_nodate,
  n.ahead = 14,
  confidence.level = 0.68,
  plt.start = tail(res_nodate$index, 1) - 30,
  series.name = "cases"
)
```

If you *do* want a lightweight, non-calendar `idx_calendar` anyway - for
example to anchor a "days since reference point" axis, or for a
genuinely non-calendar frequency such as sub-second lab data - use a
plain numeric `anchor` and `posixct = FALSE` (the default). Position
lookups then go through `idx_offset_to_pos()` (the inverse of
`idx_to_date()`'s plain-arithmetic branch) rather than `idx_to_pos()`,
since there is no real date to parse:

```{r no-dates-numeric-anchor}
gauteng_daynum_cal <- idx_calendar(
  anchor = 0, anchor_pos = 1L,
  amount = 1, unit = "days since reference point"
)

# "day 44 since the reference point" -> idx_series position
start.pos <- idx_offset_to_pos(gauteng_daynum_cal, 44)
end.pos   <- idx_offset_to_pos(gauteng_daynum_cal, 119)

mod_daynum <- SSModelDynamicGompertz(
  Y = gauteng_nodate_idx, start = start.pos, end = end.pos,
  calendar = gauteng_daynum_cal
)
res_daynum <- estimate(mod_daynum)
```

With a non-`posixct` `calendar` attached, the x-axis is still plain
integer positions (since `idx_to_date()` only produces real dates when
`calendar$posixct = TRUE`) - but the calendar is carried through to
`idx_to_date()`/`idx_offset_to_pos()` so that estimation windows can be
specified in "days since reference point" rather than raw positions:

```{r fig-daynum-log, fig.cap = "Log forecast using a numeric-anchored (non-POSIXct) calendar: still a plain integer/offset x-axis."}
tsgc::plot_log_forecast(
  res = res_daynum,
  Y = gauteng_nodate_idx,
  n.ahead = 14,
  plt.start = tail(res_daynum$index, 1) - 30
)
```

# Illustration of the `tsgc` package

## New cases and their moving average (Figure 1)

```{r make-fig-1, fig.cap="Figure 1: New Cases and their centered 7-day moving average for Gauteng province in South Africa between 10 March 2020 and 5 January 2022."}
mod_all <- SSModelDynamicGompertz(Y = gauteng_idx, calendar = gauteng_cal)
plot(mod_all, series.name = "daily cases", MA_period = 7)
```

## Setting up the forecasting exercise

`start`/`end` are now integer positions (in the same units as
`gauteng_idx$start`), not dates. We convert the estimation window's
calendar dates to positions using the calendar built above.

```{r set_options}
Y <- gauteng_idx
cal <- gauteng_cal

estimation.pos.start <- idx_to_pos(cal, "2021-02-01")
estimation.pos.end   <- idx_to_pos(cal, "2021-05-03")
n.forecasts <- 14
q <- 0.005
confidence.level <- 0.68
plt.length <- 30
```

## Estimation

```{r model_est_free_q}
model_q <- SSModelDynamicGompertz(Y = Y, sea.period = 7,
                                  start = estimation.pos.start,
                                  end = estimation.pos.end,
                                  calendar = cal)
res_q <- estimate(model_q)
```

```{r model_est_fix_q}
model <- SSModelDynamicGompertz(Y = Y, q = q, sea.period = 7,
                                start = estimation.pos.start,
                                end = estimation.pos.end,
                                calendar = cal)
res <- estimate(model)
```

## Results



```{r fig-2, fig.cap="Figure 2: Fourteen-day forecast of $ln(g_t)$ from 4 May 2021, for Gauteng province in South Africa."}

tsgc::plot_log_forecast(
  res = res,
  Y = Y,
  n.ahead = n.forecasts,
  plt.start = tail(res$index, 1) - plt.length
)
```

```{r fig-3, fig.cap = "Figure 3: Fourteen-day forecast of new cases from 4 May 2021 for Gauteng province in South Africa. The shaded region is a state-implied forecast band obtained by propagating uncertainty in the latent log growth rate."}
tsgc::plot_forecast(
  res = res,
  n.ahead = n.forecasts,
  confidence.level = confidence.level,
  plt.start = tail(res$index, 1) - plt.length,
  series.name = "cases"
)
```

```{r fig-4, fig.cap = "Figure 4: Accuracy of the fourteen-day forecast of new cases from 20 April 2021 for Gauteng province in South Africa."}
# Update estimation end position
estimation.pos.end <- idx_to_pos(cal, "2021-04-19")

# Reestimate the model
model <- SSModelDynamicGompertz$new(Y = Y, q = q, sea.period = 7,
                                    start = estimation.pos.start,
                                    end = estimation.pos.end,
                                    calendar = cal)
res_eval <- estimate(model)

tsgc::plot_holdout(
  res = res_eval,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

## Reproduction number $R_t$ (Figure 5)

```{r calc_rt}
gen_int <- 4
r.t <- estimate_r0(res, gen_int, ndays = 7)
```

```{r fig-5, fig.cap = "Figure 5: Reproduction numbers for the 7-day period to 3 May 2021, for Gauteng province in South Africa."}
r.t.mat <- as.matrix(idx_values(r.t))
r.t.df <- data.frame(
  x = idx_to_date(cal, idx_positions(r.t)),
  fit = r.t.mat[, "fit"],
  lower = r.t.mat[, "lower"],
  upper = r.t.mat[, "upper"]
)

fit <- lower <- upper <- NULL
ggplot(r.t.df, aes(x = x)) +
  geom_line(aes(y = fit, color = "Rt")) +
  geom_point(aes(y = fit), color = "red", size = 3) +
  geom_segment(aes(xend = x, yend = lower, y = fit), color = "blue") +
  geom_segment(aes(xend = x, yend = upper, y = fit), color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = "68%  Interval"), alpha = 0.2) +
  geom_hline(yintercept = 1, linetype = "solid", linewidth = 1.5, color = "black") +
  scale_x_date(date_breaks = "1 day", labels = scales::date_format("%d %b %y")) +
  labs(title = "Reproduction numbers for the 7-day period to 3 May 2021",
      x = "Date", y = expression(R[t])) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_light(base_size = 12) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.2),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(face = "bold")
  )
```

# Reinitialization

```{r change-est-date-end}
estimation.pos.end <- idx_to_pos(cal, "2021-06-25")
```

## Triggering reinitialization (Figure 6)

```{r fig-6, fig.cap = "Figure 6: Trigger for reinitialization: when smoothed slope exceeds twice its standard error above zero, reinitialization is triggered to the date when the smoothed slope crossed zero."}
# Re-estimate model
model <- SSModelDynamicGompertz$new(Y = Y, q = q, sea.period = 7,
                                    start = estimation.pos.start,
                                    end = estimation.pos.end,
                                    calendar = cal)
res <- estimate(model)

# Extract the smoothed slope and its standard deviation as idx_series,
# then convert to a data.frame with real dates for plotting.
smoothed.slope.full <- idx_series(res$output$alphahat[, "slope"], start = res$index[1])
smoothed.P.slope <- idx_series(res$output$P[2, 2, -1], start = res$index[1])

common_pos <- idx_positions(smoothed.slope.full)
d2 <- idx_series(
  cbind(
    smoothed.slope = idx_values(smoothed.slope.full),
    sd.smoothed.slope = sqrt(idx_values(smoothed.P.slope)),
    sd.smoothed.slope.1.5 = 1.5 * sqrt(idx_values(smoothed.P.slope)),
    sd.smoothed.slope.2 = 2 * sqrt(idx_values(smoothed.P.slope))
  ),
  start = common_pos[1]
)

d2.mat <- as.matrix(idx_values(d2))
d2.df <- data.frame(
  Date = idx_to_date(cal, idx_positions(d2)),
  smoothed.slope = d2.mat[, "smoothed.slope"],
  sd.smoothed.slope = d2.mat[, "sd.smoothed.slope"],
  sd.smoothed.slope.1.5 = d2.mat[, "sd.smoothed.slope.1.5"],
  sd.smoothed.slope.2 = d2.mat[, "sd.smoothed.slope.2"]
)

d2.df <- d2.df[d2.df$Date >= as.Date("2021-02-15"), ]

trigger.df <- d2.df %>%
  mutate(prev_smoothed.slope = lag(smoothed.slope)) %>%
  filter((smoothed.slope > sd.smoothed.slope.2 &
            prev_smoothed.slope < sd.smoothed.slope.2))
# Triggered on May 01 when smoothed slope > 2* sd(smoothed slope)

reinit_zero.df <- d2.df %>%
  mutate(prev_smoothed.slope = lag(smoothed.slope)) %>%
  filter(Date < min(trigger.df$Date) &
           (smoothed.slope > 0 & prev_smoothed.slope < 0)) %>%
  arrange(desc(Date)) %>%
  slice(1)
# Reinitialisation on April 21

# Create plot
ggplot(data = d2.df,
      aes(x = Date)) +
  geom_line(aes(y = smoothed.slope,
                color = "smoothed.slope"), linewidth=.5) +
  geom_line(aes(y = sd.smoothed.slope,
                color = "sd.smoothed.slope"),
            linetype = "solid",
            linewidth=.25) +
  geom_line(aes(y = sd.smoothed.slope.1.5,
                color = "sd.smoothed.slope.1.5"),
            linetype = "solid",
            linewidth=.25) +
  geom_line(aes(y = sd.smoothed.slope.2,
                color = "sd.smoothed.slope.2"),
            linetype = "solid",
            linewidth=.5) +
  scale_y_continuous(n.breaks = 10) +
  geom_hline(yintercept = 0,
             linetype = "solid",
             color = "black",linewidth = 1)+
  geom_vline(data = trigger.df,
             aes(xintercept = Date),
             linetype = "solid", size = .5,
             color = "black")+
  ylab("Slope") +
  scale_x_date(date_breaks = "10 days") +
  scale_color_manual(name='',
                     values=c('smoothed.slope'='red',
                              'sd.smoothed.slope'='blue',
                              'sd.smoothed.slope.1.5'='green',
                              'sd.smoothed.slope.2'='black'))+
  theme_light(base_size = 11)+
  theme(legend.title=element_text(size=2),
        legend.text=element_text(size=6))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        plot.title = element_text(face = "bold"))
```

```{r set_reinit_date}
reinit.pos <- idx_to_pos(cal, "2021-04-21")
```

## Estimating the reinitialized model

`reinit.date` becomes `reinit.idx`, an integer position rather than a
`Date`.

```{r reinit_estim}
model <- SSModelDynamicGompertz$new(
  Y = Y, q = q, sea.period = 7,
  start = estimation.pos.start,
  end = estimation.pos.end,
  reinit.idx = reinit.pos,
  calendar = cal
)
res.reinit <- estimate(model)
```

```{r fig-7, fig.cap = "Figure 7: Forecast of $ln(g_t)$ after reinitialization."}
tsgc::plot_log_forecast(
    res = res.reinit,
    Y = Y,
    n.ahead = n.forecasts,
    plt.start = tail(res.reinit$index, 1) - plt.length,
    title='Forecast of log growth rate after reinitialization.'
)
```

```{r fig-8, fig.cap = "Figure 8: Forecast accuracy of the model without reinitialization over the hold-out sample period: 14 days from 25 June 2021."}
tsgc::plot_holdout(
  res = res,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

```{r fig-9, fig.cap = "Figure 9: Forecast accuracy of the model with reinitialization over the hold-out sample period: 14 days from 25 June 2021."}
tsgc::plot_holdout(
  res = res.reinit,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

```{r fig-10, fig.cap = "Figure 10: Comparison of standard (green) and reinitialized (blue) forecasts"}
tsgc::plot_compare_forecast(
  results = list(res.reinit, res),
  n.ahead = 14,
  actual = Y
)
```

# Leading indicator model: England cases and hospitalisations

```{r li-setup}
eng <- england_idx
eng_cal <- england_cal

est.start.eng <- idx_to_pos(eng_cal, "2021-04-30")
est.end.eng   <- idx_to_pos(eng_cal, "2021-07-24")
n.lag.eng     <- 4
n.forecasts.eng <- 7
plt.length.eng  <- 14
```

```{r li-plot, fig.cap = "Figure 11: Daily COVID cases and hospitalisations, England."}
mod.eng.plot <- SSModelLeadingIndicator$new(eng, n.lag = n.lag.eng, calendar = eng_cal)
plot(
  mod.eng.plot,
  title = "Daily COVID cases and hospitalisations (England)",
  series.name.lead = "Cases",
  series.name.target = "Hospitalisations",
  take.log = TRUE
)
```

```{r li-estimate}
mod.eng <- SSModelLeadingIndicator$new(
  Y = eng,
  n.lag = n.lag.eng,
  LeadIndCol = 1,
  sea.period = 7,
  start = est.start.eng,
  end = est.end.eng,
  calendar = eng_cal
)
res.eng <- estimate(mod.eng)
```

```{r fig-li-log, fig.cap = "Figure 12: Forecast of log growth rate of hospitalisations, England."}
tsgc::plot_log_forecast(
  res.eng,
  Y = eng,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  title = "Forecast of log growth rate of hospitalisations (England)"
)
```

```{r fig-li-forecast, fig.cap = "Figure 13: Forecast of hospitalisations, England."}
tsgc::plot_forecast(
  res.eng,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  series.name = "Hospitalisations",
  title = "Forecast of hospitalisations (England)"
)
```

```{r fig-li-holdout, fig.cap = "Figure 14: Accuracy of the seven-day forecast of hospitalisations, England."}
tsgc::plot_holdout(
  res.eng,
  Y = eng,
  n.ahead = n.forecasts.eng,
  series.name = "Hospitalisations",
  title = "Accuracy: forecast of hospitalisations (England)"
)
```

## Adding regressors and future covariate paths

`supply_xpred.new()` was an internal-only helper in the old package
(setting `xpred.new`/`xpred_lead.new`/`xpred_targ.new`); the fields areset directly on the `FilterResultsLI` object instead.

```{r li-xpred}
xpred_lead <- xpred_targ <- england_weather_idx

mod.eng.x <- SSModelLeadingIndicator$new(
  eng,
  n.lag = n.lag.eng,
  xpred_lead = xpred_lead,
  xpred_targ = xpred_targ,
  sea.period = 7,
  start = est.start.eng,
  end = est.end.eng,
  calendar = eng_cal
)
res.eng.x <- estimate(mod.eng.x)
res.eng.x$xpred_lead.new <- england_weather_idx
res.eng.x$xpred_targ.new <- england_weather_idx
```

```{r fig-li-x-forecast, fig.cap = "Figure 15: Forecast of hospitalisations with weather regressors, England."}
tsgc::plot_forecast(
  res.eng.x,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  title = "Forecast of hospitalisations with regressors",
  series.name = "Hospitalisations"
)
```

# Comparing the dynamic Gompertz and leading-indicator models

```{r cv-setup}
n.forecasts.uk <- 14
plt.length.uk  <- 30
q.uk   <- 0.005

est.start.uk <- idx_to_pos(ukitaly_cal, "2020-02-25")
est.end.uk   <- idx_to_pos(ukitaly_cal, "2020-04-01")
Yuk          <- idx_series(idx_values(ukitaly_idx)[, "UK"], start = ukitaly_idx$start)
```

```{r fig-uk-italy, fig.cap = "Figure 16: Daily COVID cases in UK and Italy."}
mod.ukit.plot <- SSModelLeadingIndicator$new(ukitaly_idx, n.lag = 4, calendar = ukitaly_cal)
plot(
  mod.ukit.plot,
  title = "Daily COVID cases in UK and Italy",
  series.name.lead = "Italy",
  series.name.target = "UK",
  take.log = FALSE
)
```

```{r uk-gompertz}
mod.uk.gomp <- SSModelDynamicGompertz$new(
  Y = Yuk, q = q.uk, sea.period = 7,
  start = est.start.uk, end = est.end.uk,
  calendar = ukitaly_cal
)
res.uk.gomp <- estimate(mod.uk.gomp)
```

```{r uk-lead}
n.lag.uk <- 14
mod.uk.lead <- SSModelLeadingIndicator$new(
  Y = ukitaly_idx, n.lag = n.lag.uk, sea.period = 7,
  start = est.start.uk, end = est.end.uk,
  calendar = ukitaly_cal
)
res.uk.lead <- estimate(mod.uk.lead)
```

```{r fig-uk-compare, fig.cap = "Figure 17: Comparison of dynamic Gompertz and leading-indicator forecasts, UK daily cases."}
tsgc::plot_compare_forecast(
  list(res.uk.gomp, res.uk.lead),
  actual = Yuk,
  n.ahead = n.forecasts.uk
)
```

```{r fig-uk-holdout, fig.cap = "Figure 18: Accuracy of the leading-indicator forecast, UK daily cases."}
tsgc::plot_holdout(
  res.uk.lead, Y = ukitaly_idx, n.ahead = n.forecasts.uk,
  title = "Accuracy: leading-indicator forecast of UK cases",
  series.name = "UK cases"
)
```

## Selecting the lag by rolling cross-validation

```{r cv-models}
cv_models <- list()

cv_models[["Vanilla_q"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, q = q.uk, sea.period = 7,
  start = est.start.uk, end = est.end.uk,
  calendar = ukitaly_cal
)

cv_models[["Vanilla_ar1"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, sea.period = 7,
  start = est.start.uk, end = est.end.uk, ar1 = TRUE,
  calendar = ukitaly_cal
)

for (i in c(7, 10, 14, 16, 18, 21)) {
  cv_models[[paste0("Lag", i)]] <- SSModelLeadingIndicator$new(
    Y = ukitaly_idx, sea.period = 7,
    start = est.start.uk, end = est.end.uk, n.lag = i,
    calendar = ukitaly_cal
  )
}
```

```{r cv-run}
cross_val(
  Y = ukitaly_idx,
  model_list = cv_models,
  est.end = est.end.uk,
  criterion = "smape",
  n.ahead = 14,
  n.estimate = 5,
  gap = 2
)
```

# Extensions: non-epidemiological applications and alternative data frequencies

## Quarterly data (Figure 19)

Quarterly positions were converted using `idx_to_pos()`, with quarter
start dates passed as plain `Date` strings.

```{r freq-quarterly}
n.lag.q <- idx_to_pos(nintendo_cal, "2017-01-01") - idx_to_pos(nintendo_cal, "2006-10-01")

mod.q <- SSModelLeadingIndicator$new(
  Y = nintendo_idx,
  sea.period = 4,
  n.lag = n.lag.q,
  start = idx_to_pos(nintendo_cal, "2017-01-01"),
  end = idx_to_pos(nintendo_cal, "2019-10-01"),
  calendar = nintendo_cal
)
res.q <- estimate(mod.q)
```

```{r fig-freq-q, fig.cap = "Figure 19: Log forecasts of quarterly Switch sales, using Wii sales as a leading indicator."}
tsgc::plot_log_forecast(
  res.q, Y = nintendo_idx, n.ahead = 8,
  title = "Log forecasts of Switch sales"
)
```

## Monthly data (Figure 20)

```{r freq-monthly}
n.lag.m <- idx_to_pos(etrading_cal, "2017-07-01") - idx_to_pos(etrading_cal, "2017-01-01")

mod.m <- SSModelLeadingIndicator$new(
  Y = etrading_idx,
  sea.period = 12,
  n.lag = n.lag.m,
  start = idx_to_pos(etrading_cal, "2017-07-01"),
  end = idx_to_pos(etrading_cal, "2021-02-01"),
  calendar = etrading_cal
)
res.m <- estimate(mod.m)
```

```{r fig-freq-m, fig.cap = "Figure 20: Forecasts of monthly AvaTrade downloads, using DEGIRO downloads as a leading indicator."}
tsgc::plot_forecast(
  res.m, n.ahead = 4,
  title = "AvaTrade monthly downloads",
  series.name = "Downloads"
)
```
