---
title: "Tracking the 2026 Bundibugyo Ebola Outbreak with tsgc"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Tracking the 2026 Bundibugyo Ebola Outbreak with tsgc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r libraries}
library(tsgc)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)

theme_set(theme_economist_white(gray_bg = FALSE, base_size = 16))
```

## Introduction

In May 2026 the Democratic Republic of Congo declared an outbreak of
Bundibugyo ebolavirus centred on Ituri and North Kivu provinces, spreading
into neighbouring Uganda. WHO declared it a Public Health Emergency of
International Concern on 16 May 2026. By the most recent national situation
report bundled with this package (10 August 2026), the outbreak had reached
4,449 confirmed cases and 2,061 deaths, making it the second-largest Ebola
outbreak on record after West Africa 2014-16, and the fastest-growing.

This vignette works entirely with the raw INSP SitRep MVE extracts shipped
in `inst/extdata/ebola`, at two levels of aggregation:

* **National daily cumulative totals** - confirmed cases, confirmed deaths,
  suspected cases, suspected deaths, and recoveries, one row per report
  date (`insp_sitrep__national_cumulative_*__daily.csv`).
* **Health-zone daily cumulative totals** - confirmed cases and confirmed
  deaths for 61 named health zones, plus a sparser "new confirmed cases"
  series (`zone_cumulative_confirmed_cases__daily.csv`,
  `zone_cumulative_confirmed_deaths__daily.csv`,
  `zone_new_confirmed_cases__daily.csv`).

Two features of this data need to be handled explicitly before any of
tsgc's state-space machinery is useful:

1. **Reporting is irregular.** Sitreps are not published every calendar
   day - there are gaps of two, three, sometimes several days. `tsgc`'s
   `idx_series`/`idx_calendar` framework can express this in two ways: a
   *patterned* calendar that treats the specific sequence of gaps as a
   repeating cycle, or a genuine *daily* calendar with explicit `NA` on
   days with no report. We use the latter throughout, because it is the
   one that is both accurate (gaps are gaps, not a repeating pattern -
   the actual reporting cadence has no cycle) and legible in plots (dates
   on the x-axis are real calendar dates, not "position 37").
2. **Zone names are inconsistent.** The raw extracts contain both
   `"Miti Murhesa"` and `"Miti-Murhesa"`, both `"Nia Nia"` and
   `"Nia-Nia"`, both `"Makiso Kisangani"` and `"Makiso-Kisangani"`, plus
   the placeholder rows `"NA"` and `"Sans Fiche"` (unassigned cases). We
   harmonise these once, up front.

## Reading and cleaning the data

```{r paths}
data_dir <- system.file("extdata", "ebola", package = "tsgc")

national_files <- c(
  confirmed_cases   = "insp_sitrep__national_cumulative_confirmed_cases__daily.csv",
  confirmed_deaths  = "insp_sitrep__national_cumulative_confirmed_deaths__daily.csv",
  suspected_cases   = "insp_sitrep__national_cumulative_suspected_cases__daily.csv",
  suspected_deaths  = "insp_sitrep__national_cumulative_suspected_deaths__daily.csv",
  recovered_cases   = "insp_sitrep__national_cumulative_recovered_cases__daily.csv"
)
```

### National series

Each national file has one value column named after the metric; we strip
that down to a common `value` column, tag it with the metric name, and
row-bind everything into one long data frame.

```{r read-national}
read_national <- function(metric, file) {
  df <- read.csv(file.path(data_dir, file), stringsAsFactors = FALSE)
  names(df)[3] <- "value"
  df$date <- as.Date(df$date)
  df$metric <- metric
  df[, c("date", "metric", "value")]
}

national_long <- do.call(
  rbind,
  Map(read_national, names(national_files), national_files)
)

national_long <- national_long[order(national_long$metric, national_long$date), ]
head(national_long)
```

### Health-zone series

```{r read-zone}
zone_cases <- read.csv(file.path(data_dir, "zone_cumulative_confirmed_cases__daily.csv"),
                       stringsAsFactors = FALSE)
zone_deaths <- read.csv(file.path(data_dir, "zone_cumulative_confirmed_deaths__daily.csv"),
                        stringsAsFactors = FALSE)
zone_new <- read.csv(file.path(data_dir, "zone_new_confirmed_cases__daily.csv"),
                     stringsAsFactors = FALSE)

names(zone_cases)[3]  <- "cum_cases"
names(zone_deaths)[3] <- "cum_deaths"
names(zone_new)[3]    <- "new_cases"

# Value columns are read as character because a handful of cells contain
# "ND" (not reported, per SOURCE_metadata.yaml) or stray transcription
# artefacts (e.g. "17-"); coerce to numeric so these become explicit NA.
zone_cases$cum_cases   <- suppressWarnings(as.numeric(zone_cases$cum_cases))
zone_deaths$cum_deaths <- suppressWarnings(as.numeric(zone_deaths$cum_deaths))
zone_new$new_cases     <- suppressWarnings(as.numeric(zone_new$new_cases))

zone_cases$date  <- as.Date(zone_cases$date)
zone_deaths$date <- as.Date(zone_deaths$date)
zone_new$date    <- as.Date(zone_new$date)
```

Harmonise the duplicated zone spellings and drop the two placeholder
labels (`"NA"`, unassigned cases pending zone confirmation, and any
`"Sans Fiche"` rows, i.e. cases without a case-report form yet - neither
represents a real health zone):

```{r harmonise-zones}
harmonise_zone <- function(x) {
  x <- trimws(x)
  x <- gsub("Miti Murhesa", "Miti-Murhesa", x, fixed = TRUE)
  x <- gsub("Nia Nia", "Nia-Nia", x, fixed = TRUE)
  x <- gsub("Makiso Kisangani", "Makiso-Kisangani", x, fixed = TRUE)
  x
}

placeholder <- function(x) x %in% c("NA", "Sans Fiche", "")

zone_cases  <- zone_cases  %>% mutate(nom = harmonise_zone(nom)) %>% filter(!placeholder(nom))
zone_deaths <- zone_deaths %>% mutate(nom = harmonise_zone(nom)) %>% filter(!placeholder(nom))
zone_new    <- zone_new    %>% mutate(nom = harmonise_zone(nom)) %>% filter(!placeholder(nom))

length(unique(zone_cases$nom))
sort(unique(zone_cases$nom))
```

61 raw labels collapse to `r length(unique(zone_cases$nom))` canonical
health zones.

### Building a daily, `NA`-padded `idx_series`

The core conversion we need, repeatedly, is: given a data frame of
irregular `(date, value)` rows for one series, produce an `idx_series`
whose integer positions correspond to *every calendar day* between the
first and last report, with `NA` on days that were not reported and (per
the README) taking the **latest-vintage** value whenever a date is
duplicated within a file. We anchor every calendar at the same date, so
that all resulting `idx_series` objects share one common integer axis and
can be freely combined, lagged and compared with `idx_lag`/`idx_diff`
without any further date bookkeeping.

```{r build-daily-series}
outbreak_start <- as.Date("2026-05-14")   # first sitrep date in the data
outbreak_calendar <- idx_calendar(
  anchor = as.POSIXct(outbreak_start, tz = "UTC"),
  anchor_pos = 1L,
  amount = 1,
  unit = "days",
  posixct = TRUE,
  anchor_name = "outbreak start (2026-05-14)"
)

#' Turn an irregular (date, value) data frame into a daily idx_series,
#' anchored at outbreak_calendar, with NA on unreported days.
to_daily_idx <- function(df, date_col = "date", value_col = "value",
                         calendar = outbreak_calendar,
                         from = outbreak_start, to = max(df[[date_col]])) {
  full_dates <- seq(from, to, by = "day")
  df <- df[order(df[[date_col]]), ]
  # latest-vintage: if a date appears more than once, keep the last row
  df <- df[!duplicated(df[[date_col]], fromLast = TRUE), ]
  vals <- setNames(df[[value_col]], as.character(df[[date_col]]))
  daily <- unname(vals[as.character(full_dates)])   # NA where unmatched
  start_pos <- idx_to_pos(calendar, as.POSIXct(from, tz = "UTC"))
  idx_series(as.numeric(daily), start = start_pos)
}
```

Applying this to the five national series:

```{r national-idx-series}
national_series <- lapply(names(national_files), function(m) {
  to_daily_idx(subset(national_long, metric == m), value_col = "value",
               to = max(national_long$date[national_long$metric == m]))
})
names(national_series) <- names(national_files)

sapply(national_series, length)
```

Each series covers the same 89-day span (14 May - 10 Aug 2026) but the
number of *reported* (non-`NA`) days differs slightly by metric, since
not every sitrep restates every table:

```{r national-coverage}
sapply(national_series, function(s) sum(!is.na(idx_values(s))))
```

## National picture

### Confirmed cases and deaths

The five national series do not all end on the same date (not every
sitrep restates every table, and the most recent report(s) may omit some
metrics), so before combining them into one data frame we pad every
series out to the widest common span, anchored at the same
`outbreak_start`.

```{r national-plot-cumulative}
national_end <- max(vapply(national_series, function(s) idx_range(s)[2], integer(1)))

pad_to <- function(s, end_pos) {
  if (idx_range(s)[2] >= end_pos) return(s)
  n_pad <- end_pos - idx_range(s)[2]
  idx_series(c(idx_values(s), rep(NA_real_, n_pad)), start = s$start)
}

national_series_aligned <- lapply(national_series, pad_to, end_pos = national_end)
sapply(national_series_aligned, length)

df_nat <- data.frame(
  date = seq(outbreak_start, by = "day", length.out = length(national_series_aligned$confirmed_cases)),
  confirmed_cases  = idx_values(national_series_aligned$confirmed_cases),
  confirmed_deaths = idx_values(national_series_aligned$confirmed_deaths),
  suspected_cases  = idx_values(national_series_aligned$suspected_cases),
  recovered_cases  = idx_values(national_series_aligned$recovered_cases)
)

df_nat_long <- pivot_longer(df_nat, -date, names_to = "series", values_to = "value")

ggplot(df_nat_long, aes(x = date, y = value, color = series)) +
  geom_line(na.rm = TRUE) +
  labs(title = "DRC national cumulative totals, Bundibugyo ebolavirus outbreak 2026",
       x = NULL, y = "Cumulative count", color = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The gaps between markers are the unreported days made explicit - most
visibly around late June and mid-to-late July, exactly the periods the
README flags as irregular reporting.

### Setting up the dynamic Gompertz model

`SSModelDynamicGompertz` models the log growth rate of a cumulative series
as a smooth, evolving trend (with optional day-of-week seasonality). Daily
sitrep publication has no real weekly pattern (it is driven by field data
collection and PDF production, not a weekly rhythm) but reporting is
sparse enough that we turn seasonality off (`sea.period = 0`) and let the
`NA` gaps be handled by the state-space filter, which naturally skips
missing observations. Constructing the model object (before calling
`estimate()`) is enough to use the package's own `plot()` method for a
first look at daily new confirmed cases, with a centered 7-day moving
average:

```{r dyngompertz-national-setup}
Y_confirmed <- national_series$confirmed_cases
n.forecasts.nat <- 14
estimation.pos.end.nat <- idx_range(Y_confirmed)[2] - n.forecasts.nat

model_nat <- SSModelDynamicGompertz$new(
  Y = Y_confirmed, q = NULL, sea.period = 0,
  start = idx_range(Y_confirmed)[1], end = estimation.pos.end.nat,
  calendar = outbreak_calendar
)
plot(model_nat, series.name = "confirmed cases", MA_period = 7,
     title = "Daily new confirmed cases (national), with 7-day centered MA")
```

The last `r n.forecasts.nat` days of the national confirmed-cases series
are held back as a genuine out-of-sample holdout, so the forecast plots
below can be compared against what was actually reported.

The **case fatality ratio implied by confirmed counts** is a simple but
informative derived series (no dedicated package plotting function covers
a ratio of two series, so this one chart also uses `ggplot2` directly):

```{r cfr}
cfr <- idx_series(
  idx_values(national_series_aligned$confirmed_deaths) / idx_values(national_series_aligned$confirmed_cases),
  start = national_series_aligned$confirmed_cases$start
)
df_cfr <- data.frame(date = df_nat$date, cfr = idx_values(cfr))
ggplot(df_cfr, aes(x = date, y = cfr)) +
  geom_line(na.rm = TRUE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Implied case fatality ratio (confirmed deaths / confirmed cases)",
       x = NULL, y = "CFR") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Estimating the model

```{r dyngompertz-national, message = FALSE, warning = FALSE}
res_nat <- estimate(model_nat)
```

`summary()` reports the estimation window, model states and their
standard errors; `tsgc::print_model_diagnostics()` supplements this with
the fitted log-likelihood and a check for estimated variances sitting at
or near a degenerate boundary.

```{r dyngompertz-national-diagnostics}
summary(res_nat)
tsgc::print_model_diagnostics(res_nat)
```

```{r dyngompertz-national-log-forecast}
plot_log_forecast(res_nat, Y = Y_confirmed, n.ahead = n.forecasts.nat,
                  title = "Forecast of log growth rate of confirmed cases (national)")
```

```{r dyngompertz-national-forecast}
plot_forecast(res_nat, n.ahead = n.forecasts.nat, series.name = "confirmed cases",
              title = paste0(n.forecasts.nat, "-day forecast of new confirmed cases (national)"))
```

```{r dyngompertz-national-holdout}
plot_holdout(res_nat, Y = Y_confirmed, n.ahead = n.forecasts.nat,
            series.name = "confirmed cases",
            title = "Accuracy: forecast of new confirmed cases (national)")
```

```{r dyngompertz-national-plot}
plot_gy_ci(res_nat, series.name = "confirmed cases",
           title = "Smoothed growth rate of national confirmed cases, with 68% CI")
```

The growth rate has hovered close to, and at times above, zero for most
of June-August - consistent with the outbreak still expanding rather than
being brought under control as of the last sitrep in this dataset.

## Health-zone geography

Ituri province (Bunia, Mongbwalu, Rwampara, Nizi, Komanda, Nyankunde,
Mahagi, Djugu-area zones) has carried the great majority of cases; North
Kivu (Beni, Butembo, Katwa, Oicha, Kalunguta) is the outbreak's second
cluster; a scatter of zones in Tshopo/Haut-Uele (Kisangani, Isiro, Rungu,
Wamba) and a handful of border zones account for the rest.

```{r zone-totals}
zone_peak <- zone_cases %>%
  group_by(nom) %>%
  summarise(peak_cases = max(cum_cases, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(peak_cases))

head(zone_peak, 15)
```

```{r zone-totals-plot}
ggplot(head(zone_peak, 15), aes(x = reorder(nom, peak_cases), y = peak_cases)) +
  geom_col() +
  coord_flip() +
  labs(title = "15 health zones with the most confirmed cases to date",
       x = NULL, y = "Cumulative confirmed cases (latest report)")
```

Four zones dominate: **Bunia** (the provincial capital of Ituri, a major
urban/transport hub), **Rwampara**, **Mongbwalu** (a gold-mining town
with high population mobility), and **Nizi**. All four are in Ituri and
within roughly 30-60km of each other along the same road corridor
running north from Bunia.

```{r zone-trajectories}
top_zones <- head(zone_peak$nom, 6)

zone_cases %>%
  filter(nom %in% top_zones) %>%
  ggplot(aes(x = date, y = cum_cases, color = nom)) +
  geom_line(na.rm = TRUE) +
  labs(title = "Cumulative confirmed cases, top 6 health zones",
       x = NULL, y = "Cumulative confirmed cases", color = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Are some zones leading indicators for others?

The `SSModelLeadingIndicator` class in `tsgc` is designed for exactly this
question: given two cumulative series sharing one position axis, it
estimates a common-trend/idiosyncratic-trend state-space model in which
one series (the leading indicator, lagged by `n.lag` positions) helps
predict the other (the target). We use it here to ask whether the
**earliest-affected zone** in the Ituri cluster, Mongbwalu (first report
2026-05-14, one of only two zones reporting on day one), leads the
subsequent growth of **Bunia**, the largest and most urban zone, which
would be epidemiologically plausible: Mongbwalu-Bunia is a known transport
corridor and Bunia's role as a provincial hub means cases could arrive
there somewhat after first appearing in a smaller, earlier-affected
outlying zone.

### Preparing the paired series

Both `df2ldl` (used internally by the leading-indicator model) and the
Gompertz model require **strictly increasing, gap-free** cumulative
series over the estimation window - the model cannot itself handle `NA`
gaps or duplicate values in the training window. We therefore restrict
estimation to the sub-range where both zones have a complete run of
distinct daily reports, rather than passing the full `NA`-padded daily
series straight in. This is a deliberate trade-off: the daily/`NA`
calendar built above is what makes *presentation* (plots, dates,
irregular-gap visibility) accurate; the leading-indicator *estimation*
step needs a clean, monotone sub-window, obtained by restricting to dates
both zones actually reported on, in order.

```{r li-prep}
build_zone_series <- function(zone_name, calendar = outbreak_calendar) {
  df <- zone_cases %>% filter(nom == zone_name) %>% arrange(date)
  df <- df %>% filter(!is.na(cum_cases))
  # keep only strictly increasing points (drop any flat/duplicate-date rows)
  keep <- c(TRUE, diff(df$cum_cases) > 0)
  df <- df[keep, ]
  to_daily_idx(df, value_col = "cum_cases", calendar = calendar,
               from = min(df$date), to = max(df$date))
}

mongbwalu <- build_zone_series("Mongbwalu")
bunia     <- build_zone_series("Bunia")

#' Given a sorted vector of integer positions, return the longest run of
#' consecutive integers it contains, as c(start, end).
longest_consecutive_run <- function(pos) {
  pos <- sort(unique(pos))
  breaks <- which(diff(pos) != 1L)
  run_starts <- c(pos[1], pos[breaks + 1L])
  run_ends   <- c(pos[breaks], pos[length(pos)])
  run_lengths <- run_ends - run_starts + 1L
  best <- which.max(run_lengths)
  c(run_starts[best], run_ends[best])
}

common_pos <- intersect(
  idx_positions(mongbwalu)[!is.na(idx_values(mongbwalu))],
  idx_positions(bunia)[!is.na(idx_values(bunia))]
)
run <- longest_consecutive_run(common_pos)
run_start <- run[1]
run_end   <- run[2]

Y_li <- idx_series(
  cbind(lead = idx_values(mongbwalu[run_start:run_end]),
        targ = idx_values(bunia[run_start:run_end])),
  start = run_start
)
idx_range(Y_li)
```

### Estimating the leading-indicator model

`n.lag = 3` (three days) is used below as a plausible reporting-to-
confirmation delay for a same-province, short road-distance pair. As in
the England cases-hospitalisations example, we look at the two series on
the log scale before formal estimation, and hold back the last few days
as an out-of-sample check.

```{r li-setup}
n.lag.li <- 3
n.forecasts.li <- 7
estimation.pos.end.li <- idx_range(Y_li)[2] - n.forecasts.li

li_model_plot <- SSModelLeadingIndicator$new(Y_li, n.lag = n.lag.li, calendar = outbreak_calendar)
plot(li_model_plot, series.name.lead = "Mongbwalu (leading)",
     series.name.target = "Bunia (target)", take.log = TRUE,
     title = "Log daily new confirmed cases: Mongbwalu vs Bunia")
```

```{r li-estimate, message = FALSE, warning = FALSE}
li_model <- SSModelLeadingIndicator(
  Y = Y_li, n.lag = n.lag.li, sea.period = 0, LeadIndCol = 1,
  start = idx_range(Y_li)[1], end = estimation.pos.end.li,
  calendar = outbreak_calendar
)
li_res <- estimate(li_model)
```

```{r li-diagnostics}
summary(li_res)
tsgc::print_model_diagnostics(li_res)
```

```{r li-log-forecast}
plot_log_forecast(li_res, Y = Y_li, n.ahead = n.forecasts.li,
                  title = "Forecast of log growth rate of Bunia cases, led by Mongbwalu")
```

```{r li-forecast}
plot_forecast(li_res, n.ahead = n.forecasts.li, series.name = "Bunia confirmed cases",
             title = paste0(n.forecasts.li, "-day forecast of new confirmed cases in Bunia"))
```

```{r li-holdout}
plot_holdout(li_res, Y = Y_li, n.ahead = n.forecasts.li,
            series.name = "Bunia confirmed cases",
            title = "Accuracy: forecast of new confirmed cases in Bunia")
```

```{r li-plot-gy}
plot_gy_ci(li_res, series.name = "Bunia (led by Mongbwalu, lag = 3 days)",
           title = "Smoothed growth rate of Bunia confirmed cases, informed by Mongbwalu")
```

### Does the geography make sense?

Worth being explicit about what this analysis can and cannot claim.
Mongbwalu and Bunia sit on the same road corridor in Ituri, and Mongbwalu
was among the first two zones to report any cases at all, so a short lead
time (days, not weeks) from Mongbwalu into the province's largest urban
centre is at least directionally consistent with a mining-town-to-hub
transmission or population-movement pathway. However:

* With daily case counts still in the single-to-low-double digits per
  zone for much of the outbreak, `n.lag` is hard to pin down precisely
  from `r diff(idx_range(Y_li)) + 1` overlapping days of data; the model
  above should be read as illustrative rather than a confirmed causal
  lead-lag relationship.
* A genuinely careful version of this analysis would fit the model over a
  grid of `n.lag` values and compare out-of-sample forecast accuracy
  (e.g. via `plot_holdout()` or an equivalent of the `mapes()` diagnostic
  available on the non-leading-indicator `FilterResults` class) rather
  than picking one lag by inspection, and would also test the *reverse*
  direction (Bunia leading Mongbwalu) and a genuinely distant pairing
  (e.g. Bunia vs a North Kivu zone such as Butembo, which has no obvious
  road-adjacency reason to be linked) as a negative control.
* Sparse, gap-heavy daily counts for individual health zones are noisy;
  the smoothed growth-rate confidence bands above are correspondingly
  wide, and the model's seasonal component was switched off rather than
  estimated, which is a simplification.

### A same-province comparison: Rwampara and Nizi vs Bunia

As a second, less mining-specific example, we compare how the model's
short-run forecast for Bunia looks when built from two different, more
distant Ituri zones as the "leading" series, using `plot_compare_forecast`
to put multiple fitted models side by side.

```{r li-compare, message = FALSE, warning = FALSE}
build_li <- function(lead_zone, targ_zone = "Bunia", n.lag = 3) {
  lead_s <- build_zone_series(lead_zone)
  targ_s <- build_zone_series(targ_zone)
  pos <- intersect(idx_positions(lead_s)[!is.na(idx_values(lead_s))],
                   idx_positions(targ_s)[!is.na(idx_values(targ_s))])
  run <- longest_consecutive_run(pos)
  Y <- idx_series(cbind(lead = idx_values(lead_s[run[1]:run[2]]),
                        targ = idx_values(targ_s[run[1]:run[2]])),
                  start = run[1])
  mod <- SSModelLeadingIndicator(Y = Y, n.lag = n.lag, sea.period = 0,
                                 LeadIndCol = 1, start = idx_range(Y)[1],
                                 end = idx_range(Y)[2], calendar = outbreak_calendar)
  estimate(mod)
}

res_mongbwalu <- li_res
res_rwampara  <- build_li("Rwampara")
res_nizi      <- build_li("Nizi")

plot_compare_forecast(
  list(Mongbwalu = res_mongbwalu, Rwampara = res_rwampara, Nizi = res_nizi),
  n.ahead = 10, sea.on = FALSE,
  title = "Bunia forecasts using different Ituri zones as leading indicator"
)
```

If the three forecasts diverge noticeably, that is itself informative -
it would suggest the choice of leading zone matters, i.e. that not all
nearby zones carry equally useful early signal about Bunia, which is
plausible given that Mongbwalu, Rwampara and Nizi differ in size,
road-distance from Bunia, and the timing of their own outbreaks.

## Health-zone confirmed deaths

For completeness, the zone-level deaths file lets us check whether the
case-heavy zones are also carrying a disproportionate share of deaths, or
whether fatality is more evenly spread:

```{r zone-deaths}
zone_deaths_peak <- zone_deaths %>%
  group_by(nom) %>%
  summarise(peak_deaths = max(cum_deaths, na.rm = TRUE), .groups = "drop")

zone_combined <- zone_peak %>%
  left_join(zone_deaths_peak, by = "nom") %>%
  mutate(cfr = peak_deaths / peak_cases) %>%
  filter(peak_cases >= 10) %>%
  arrange(desc(peak_cases))

head(zone_combined, 15)
```

```{r zone-deaths-plot}
ggplot(head(zone_combined, 15), aes(x = peak_cases, y = cfr, label = nom)) +
  geom_point() +
  geom_text(vjust = -0.6, size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Implied CFR vs case burden, zones with 10+ confirmed cases",
       x = "Cumulative confirmed cases", y = "Cumulative deaths / cases")
```

## Summary

* National confirmed cases and deaths were reconstructed as daily,
  `NA`-padded `idx_series` objects on a shared `POSIXct` calendar anchored
  at the first sitrep date, making the many unreported days visible as
  genuine gaps rather than an artificial repeating pattern.
* Growth-rate estimation via `SSModelDynamicGompertz` shows the national
  outbreak's growth rate remaining near or above zero through most of the
  available window, consistent with the sustained rise in cumulative
  totals reported in the README.
* At health-zone level, four Ituri zones (Bunia, Rwampara, Mongbwalu,
  Nizi) account for the large majority of confirmed cases; harmonising
  duplicated zone-name spellings was necessary before any of this
  aggregation was reliable.
* A `SSModelLeadingIndicator` fit using Mongbwalu (an early-affected,
  smaller zone) as a 3-day-lagged leading indicator for Bunia (the
  provincial hub) is geographically plausible given their road-corridor
  proximity, but the short, sparse overlap window means this should be
  read as a demonstration of the method rather than a settled finding;
  comparing several candidate leading zones and lags, and testing
  negative-control pairings, would be the natural next step.
