---
title: "Forecasting the 2026 Bundibugyo Ebola Outbreak with tsgc"
author: "Guglielmo Secchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = TRUE,
  warning    = TRUE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

INSP SitRep MVE data on the 2026 DRC/Uganda Bundibugyo ebolavirus outbreak
(PHEIC declared 16 May 2026). National cumulative confirmed cases and
deaths are clean, strictly increasing series -- a direct fit for
`SSModelDynamicGompertz`, no surge-window reconstruction needed (unlike
the PM2.5 vignette). Reporting dates are irregular (gaps of 1-4 days
between sitreps), so the daily calendar is built explicitly with `NA` on
non-report days rather than relying on `xts_to_idx`'s pattern detection,
which requires a regular step. Confirmed cases leading confirmed deaths
is tried as a `SSModelLeadingIndicator` pair. Zone-level data (61 health
zones, national total not a simple sum -- some zones only report from
2026-05-15/18/20) is used for one zone (Bunia) as a single-zone
illustration; a full 61-zone sweep is future work.

## Data

```{r libraries}
suppressPackageStartupMessages({
  library(tsgc)
  library(dplyr)
  library(ggplot2)
  library(ggthemes)
  library(zoo)
  library(xts)
  library(tidyr)
})

theme_set(ggthemes::theme_economist_white(gray_bg = FALSE, base_size = 14))
options(scipen = 7)

CONF_LEVEL <- 0.68
```

```{r load-data}
cases_raw  <- read.csv("../inst/extdata/ebola/insp_sitrep__national_cumulative_confirmed_cases__daily.csv",
                       stringsAsFactors = FALSE)
deaths_raw <- read.csv("../inst/extdata/ebola/insp_sitrep__national_cumulative_confirmed_deaths__daily.csv",
                       stringsAsFactors = FALSE)

cases_raw$date  <- as.Date(cases_raw$date)
deaths_raw$date <- as.Date(deaths_raw$date)
cases_raw  <- cases_raw[order(cases_raw$date), ]
deaths_raw <- deaths_raw[order(deaths_raw$date), ]

str(cases_raw)
range(cases_raw$date)
range(deaths_raw$date)
```

`r nrow(cases_raw)` sitreps for confirmed cases, `r nrow(deaths_raw)` for
confirmed deaths, `r format(min(cases_raw$date), "%d %b %Y")` to
`r format(max(cases_raw$date), "%d %b %Y")`. Both strictly increasing at
report dates (checked below) -- no downward revisions, unlike the
suspected-case/suspected-death files (not used further here).

```{r check-monotonic}
stopifnot(!is.unsorted(cases_raw$national_cumulative_confirmed_cases, strictly = FALSE))
stopifnot(!is.unsorted(deaths_raw$national_cumulative_confirmed_deaths, strictly = FALSE))

# Zero-increment (flat) report-to-report steps: estimate() tolerates NA
# gaps but not non-positive increments once regridded to a daily
# calendar, since consecutive report dates always differ, so no
# same-day flats are possible here regardless.
diff(cases_raw$national_cumulative_confirmed_cases)  |> (\(x) sum(x == 0))()
diff(deaths_raw$national_cumulative_confirmed_deaths) |> (\(x) sum(x == 0))()
```

```{r raw-series-plot, fig.cap="National cumulative confirmed cases and deaths, DRC Bundibugyo outbreak 2026."}
combined_raw <- dplyr::full_join(
  cases_raw[, c("date", "national_cumulative_confirmed_cases")],
  deaths_raw[, c("date", "national_cumulative_confirmed_deaths")],
  by = "date"
) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value") %>%
  dplyr::mutate(series = dplyr::recode(series,
    national_cumulative_confirmed_cases  = "Confirmed cases",
    national_cumulative_confirmed_deaths = "Confirmed deaths"))

ggplot(combined_raw, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  geom_point(size = 1.2, na.rm = TRUE) +
  labs(title = "Cumulative confirmed cases and deaths",
       subtitle = "INSP SitRep MVE, DRC, 2026",
       x = NULL, y = "Cumulative count", color = NULL) +
  theme(plot.title = element_text(face = "bold"))
```

Cases show the classic epidemic S-shape building since mid-May, no sign
yet of a peak as of the final sitrep (10 Aug); deaths track cases with a
lag and lower level, consistent with a case-fatality delay rather than
elevated CFR change over time (checked more directly via the CFR ratio
below).

```{r cfr-plot, fig.cap="Implied cumulative case-fatality ratio over time."}
cfr_df <- dplyr::inner_join(
  cases_raw[, c("date", "national_cumulative_confirmed_cases")],
  deaths_raw[, c("date", "national_cumulative_confirmed_deaths")],
  by = "date"
)
cfr_df$cfr <- cfr_df$national_cumulative_confirmed_deaths / cfr_df$national_cumulative_confirmed_cases

ggplot(cfr_df, aes(x = date, y = cfr)) +
  geom_line(color = "black", linewidth = 0.7) +
  geom_point(size = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cumulative case-fatality ratio",
       x = NULL, y = "Deaths / cases") +
  theme(plot.title = element_text(face = "bold"))
```

CFR falls from an early high (small-numbers noise, first few sitreps) and
stabilises; current national estimate `r round(100*tail(cfr_df$cfr,1),1)`%
(2,061/4,449 as of 10 Aug per the README).

## Building a daily calendar with explicit gaps

Sitrep gaps are irregular (1-4 days between consecutive reports), not a
detectable single-step pattern, so `xts_to_idx(detect = TRUE)` will fail
here -- it needs a regular step or repeating gap cycle. Instead, both
series are regridded onto an explicit daily sequence spanning their full
date range, with `NA` on non-report days; this daily grid *is* regular,
so `xts_to_idx` then detects `amount = 1, unit = "days"` on it directly.

```{r daily-grid}
build_daily_idx <- function(raw, value_col, anchor_date = NULL) {
  full_range <- seq(min(raw$date), max(raw$date), by = "day")
  daily <- data.frame(date = full_range) %>%
    dplyr::left_join(raw[, c("date", value_col)], by = "date")
  x <- xts::xts(daily[[value_col]], order.by = daily$date)
  conv <- xts_to_idx(x, start.pos = if (is.null(anchor_date)) 1L else
                       idx_to_pos(NULL, anchor_date))
  list(series = conv$series, calendar = conv$calendar, daily = daily)
}

cases_conv  <- build_daily_idx(cases_raw,  "national_cumulative_confirmed_cases")
Y_cases     <- cases_conv$series
cal_cases   <- cases_conv$calendar

length(Y_cases)
sum(is.na(idx_values(Y_cases)))
idx_range(Y_cases)
```

`r length(Y_cases)` daily positions spanning the sitrep range,
`r sum(is.na(idx_values(Y_cases)))` `NA` (non-report) days. `KFAS`
propagates the state prediction across these without an update step, as
in the PM2.5 vignette.

```{r deaths-grid}
deaths_conv <- build_daily_idx(deaths_raw, "national_cumulative_confirmed_deaths")
Y_deaths    <- deaths_conv$series
cal_deaths  <- deaths_conv$calendar

length(Y_deaths)
sum(is.na(idx_values(Y_deaths)))
```

Cases and deaths cover different date ranges to the day
(`r format(min(cases_raw$date), "%d %b")`-`r format(max(cases_raw$date), "%d %b %Y")`
vs `r format(min(deaths_raw$date), "%d %b")`-`r format(max(deaths_raw$date), "%d %b %Y")`),
so each gets its own `xts_to_idx` call and its own calendar; they are put
on a single shared calendar below only when building the leading-indicator
pair, where `SSModelLeadingIndicator` requires one combined `idx_series`.

# Model 1: dynamic Gompertz on national confirmed cases

No surge-window transform needed here -- national confirmed cases is
already a single strictly increasing cumulative series over the full
outbreak to date, unlike PM2.5's need for within-season resets. Fitting
on the full available history and holding out the final 14 days to check
forecast accuracy.

```{r model1-setup}
n.forecasts <- 14
q.default   <- 0.005

rng_cases <- idx_range(Y_cases)
fit_end_pos <- rng_cases[2] - n.forecasts
stopifnot(!is.na(idx_values(Y_cases[fit_end_pos])))
stopifnot(!is.na(idx_values(Y_cases[rng_cases[2]])))
```

```{r model1-inspect, fig.cap="Cumulative confirmed cases, full outbreak to date."}
mod_check <- tsgc::SSModelDynamicGompertz(Y = Y_cases, calendar = cal_cases)
p <- plot(mod_check, title = "Cumulative confirmed Ebola cases, DRC 2026",
         series.name = "Cum. confirmed cases")
print(p)
```

## Estimation, holding out the final 14 days

```{r model1-fit}
model_cases <- tsgc::SSModelDynamicGompertz(
  Y        = Y_cases,
  q        = q.default,
  sea.period = 0,
  start    = rng_cases[1],
  end      = fit_end_pos,
  calendar = cal_cases
)
res_cases <- tsgc::estimate(model_cases)
summary(res_cases)
tsgc::print_model_diagnostics(res_cases)
```

`sea.period = 0` here rather than the PM2.5 vignette's `7` -- sitreps are
irregular, not day-of-week patterned, so a weekly seasonal component has
no obvious mechanism and would be fit against mostly-`NA` day-of-week
cells anyway.

## Forecast and holdout accuracy

```{r model1-forecast-log, fig.cap="Log growth rate forecast, 14 days held out."}
p <- tsgc::plot_log_forecast(
  res_cases, Y = Y_cases, n.ahead = n.forecasts,
  plt.start = fit_end_pos - 30,
  title = "Log growth rate of cumulative confirmed cases\n14-day holdout"
)
print(p)
```

```{r model1-forecast-level, fig.cap="New (daily) confirmed cases forecast, 14-day holdout."}
p <- tsgc::plot_forecast(
  res_cases, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = fit_end_pos - 30,
  title = "Forecast of daily new confirmed cases\n14-day holdout",
  series.name = "Confirmed cases"
)
print(p)
```

```{r model1-holdout, fig.cap="Holdout accuracy, 14 days."}
p <- tsgc::plot_holdout(
  res_cases, Y = Y_cases, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily confirmed cases forecast, 14-day holdout",
  series.name = "Confirmed cases"
)
print(p)
```

```{r model1-mapes}
res_cases$mapes(n.ahead = n.forecasts, Y = Y_cases)
```

## Reproduction number $R_t$

`plot_r0()` requires a `gen_int` argument (generation-interval
distribution) that this vignette does not yet supply -- Ebola virus
disease serial interval estimates in the outbreak literature cluster
around 12-15 days, but the correct object to pass (a discretised
gamma vector, or whatever form `tsgc` expects) needs deciding in
consultation with the epi team rather than assumed here. Left as
`eval=FALSE` pending that.

```{r r0-plot, fig.cap="Estimated reproduction number Rt, national confirmed cases.", eval=FALSE}
gen_int <- NULL  # TODO: real generation-interval object for Bundibugyo ebolavirus
p <- tsgc::plot_r0(res_cases, gen_int = gen_int, n.ahead = 7)
print(p)
```

# Rolling forecast origins through the outbreak

Refitting at successive origins and checking 14-day-ahead accuracy across
the outbreak's course, as in the PM2.5 vignette's rolling-origin design.
Origins at roughly fortnightly intervals from the first date with 14 full
days of subsequent holdout available.

```{r rolling-origins}
origin_dates <- seq(min(cases_raw$date) + 21, max(cases_raw$date) - n.forecasts, by = 14)

fit_rolling_origin <- function(Y, calendar, origin_date, q = q.default,
                               n.ahead = 14, sea.period = 0) {
  origin_pos <- idx_to_pos(calendar, origin_date)
  rng <- idx_range(Y)
  if (origin_pos + n.ahead > rng[2]) {
    return(NULL)
  }
  if (is.na(idx_values(Y[origin_pos]))) {
    return(NULL)  # origin falls on a non-report day
  }
  model <- tsgc::SSModelDynamicGompertz(
    Y = Y, q = q, sea.period = sea.period,
    start = rng[1], end = origin_pos, calendar = calendar
  )
  res <- tryCatch(tsgc::estimate(model), error = function(e) NULL)
  if (is.null(res)) return(NULL)
  acc <- tryCatch(res$mapes(n.ahead = n.ahead, Y = Y), error = function(e) NULL)
  list(origin = origin_date, res = res, accuracy = acc)
}

rolling_cases <- lapply(origin_dates, function(d) fit_rolling_origin(Y_cases, cal_cases, d))
names(rolling_cases) <- format(origin_dates)
rolling_cases <- Filter(Negate(is.null), rolling_cases)
length(rolling_cases)
```

```{r rolling-table}
rolling_accuracy <- do.call(rbind, lapply(names(rolling_cases), function(nm) {
  a <- rolling_cases[[nm]]$accuracy
  data.frame(origin = nm, mape = a$mape, smape = a$smape,
            mae = a$mae, rmse = a$rmse, coverage = a$coverage)
}))
rolling_accuracy
```

```{r rolling-plot, fig.cap="14-day-ahead forecast accuracy by origin, national confirmed cases."}
rolling_long <- tidyr::pivot_longer(
  rolling_accuracy, cols = c(smape, coverage),
  names_to = "metric", values_to = "value"
)
rolling_long$origin <- factor(rolling_long$origin, levels = rolling_accuracy$origin)

ggplot(rolling_long, aes(x = origin, y = value, group = metric, color = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(title = "14-day-ahead forecast accuracy across the outbreak",
       x = "Forecast origin", y = NULL, color = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))
```

Sample is short (outbreak began mid-May, data to 10 Aug -- under three
months), so few rolling origins are available; treat as illustrative of
the method rather than a stable accuracy estimate. Revisit once the
outbreak has run longer or reaches its peak.

# Leading indicator model: confirmed cases leading confirmed deaths

Death reporting typically lags case confirmation (time from confirmation
to death, plus reporting delay), so cases are a natural candidate leading
indicator for deaths. `SSModelLeadingIndicator` needs both series as a
single 2-column `idx_series` on one shared calendar -- cases and deaths
were converted on separate calendars above, so they are re-combined here.

```{r deflatten-helper}
# estimate() (both SSModelDynamicGompertz and SSModelLeadingIndicator)
# requires strictly positive non-missing increments. NA gaps (non-report
# days) are fine -- KFAS skips them -- but a same-to-next-report flat
# step, once regridded to daily positions, is a zero increment between
# two adjacent non-NA positions, which is rejected. Per the package's
# own guidance, add a small increment at each such flat point (only
# among the non-NA observations, preserving the NA gaps as NA) so the
# non-missing values are strictly increasing.
deflatten_idx_series <- function(Y, bump = 0.01) {
  vals <- idx_values(Y)
  obs_idx <- which(!is.na(vals))
  if (length(obs_idx) < 2) return(Y)
  obs_vals <- vals[obs_idx]
  flat_at <- which(diff(obs_vals) == 0)
  if (length(flat_at) > 0) {
    adj <- rep(0, length(obs_vals))
    for (i in flat_at) adj[(i + 1):length(obs_vals)] <- adj[(i + 1):length(obs_vals)] + bump
    obs_vals <- obs_vals + adj
  }
  vals[obs_idx] <- obs_vals
  idx_series(vals, start = Y$start)
}
```

```{r li-setup}
common_start <- max(idx_range(Y_cases)[1], idx_range(Y_deaths)[1])
common_end   <- min(idx_range(Y_cases)[2], idx_range(Y_deaths)[2])

Y_cases_common  <- get_timeframe(Y_cases,  common_start, common_end)
Y_deaths_common <- get_timeframe(Y_deaths, common_start, common_end)

# De-flatten both series before combining -- df2ldl_lead()'s increments
# (newLead/newTarg) must be strictly positive wherever non-missing, and
# both cumulative series contain flat (zero-increment) steps once
# regridded to daily positions.
Y_cases_li  <- deflatten_idx_series(Y_cases_common)
Y_deaths_li <- deflatten_idx_series(Y_deaths_common)
stopifnot(all(na.omit(diff(idx_values(Y_cases_li)))  > 0))
stopifnot(all(na.omit(diff(idx_values(Y_deaths_li))) > 0))

Y_li <- idx_cbind(Y_cases_li, Y_deaths_li)
colnames(Y_li$data) <- c("cases", "deaths")
cal_li <- cal_cases  # cases and deaths share the same daily grid/anchor unit;
                     # cal_cases and cal_deaths only differ in anchor date/pos,
                     # both already realigned onto common_start via get_timeframe
idx_range(Y_li)
```

```{r li-lag-selection}
# Candidate lags (days from confirmation to reported death): trying a
# short range around typical Ebola case-to-death intervals rather than
# a single assumed value.
lag_candidates <- 3:14
li_fit_end <- common_end - n.forecasts

fit_li <- function(n.lag) {
  tryCatch({
    model <- tsgc::SSModelLeadingIndicator(
      Y = Y_li, n.lag = n.lag, sea.period = 0, q = NULL,
      LeadIndCol = 1, start = common_start, end = li_fit_end,
      calendar = cal_li
    )
    res <- tsgc::estimate(model)
    acc <- res$mapes(n.ahead = n.forecasts, Y = Y_deaths_common)
    list(n.lag = n.lag, res = res, accuracy = acc, error = NULL)
  }, error = function(e) list(n.lag = n.lag, res = NULL, accuracy = NULL, error = conditionMessage(e)))
}

li_by_lag <- lapply(lag_candidates, fit_li)
names(li_by_lag) <- lag_candidates

# Surface any per-lag fit failures rather than silently dropping them --
# an empty li_by_lag after filtering means every lag failed, which is a
# data/setup problem worth seeing, not something to plot past.
li_errors <- Filter(function(x) !is.null(x$error), li_by_lag)
if (length(li_errors) > 0) {
  cat("Lag candidates that failed to fit:\n")
  invisible(lapply(li_errors, function(x) cat(" n.lag =", x$n.lag, "-", x$error, "\n")))
}

li_by_lag <- Filter(function(x) is.null(x$error), li_by_lag)
stopifnot("No lag candidate produced a usable SSModelLeadingIndicator fit -- see errors printed above." =
            length(li_by_lag) > 0)

li_lag_table <- do.call(rbind, lapply(names(li_by_lag), function(nm) {
  a <- li_by_lag[[nm]]$accuracy
  data.frame(n.lag = as.integer(nm), mape = a$mape, smape = a$smape,
            mae = a$mae, rmse = a$rmse)
}))
li_lag_table
```

```{r li-lag-plot, fig.cap="Leading-indicator forecast accuracy (sMAPE) by assumed cases-to-deaths lag."}
ggplot(li_lag_table, aes(x = n.lag, y = smape)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "14-day-ahead deaths forecast accuracy vs assumed lag",
       x = "Lag (days, cases leading deaths)", y = "sMAPE") +
  theme(plot.title = element_text(face = "bold"))
```

```{r li-best-lag}
best_lag <- li_lag_table$n.lag[which.min(li_lag_table$smape)]
best_lag
res_li <- li_by_lag[[as.character(best_lag)]]$res
summary(res_li)
tsgc::print_model_diagnostics(res_li)
```

```{r li-forecast-plot, fig.cap="Deaths forecast from the leading-indicator model, best-fit lag."}
p <- tsgc::plot_forecast(
  res_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = li_fit_end - 30,
  title = paste0("Confirmed deaths forecast, cases leading by ", best_lag, " days"),
  series.name = "Confirmed deaths"
)
print(p)
```

```{r li-holdout-plot, fig.cap="Holdout accuracy, leading-indicator deaths forecast."}
p <- tsgc::plot_holdout(
  res_li, Y = Y_deaths_common, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = paste0("Accuracy: deaths forecast, lag = ", best_lag, " days"),
  series.name = "Confirmed deaths"
)
print(p)
```

## Comparison: leading-indicator vs univariate deaths model

```{r li-vs-univariate}
model_deaths_uni <- tsgc::SSModelDynamicGompertz(
  Y = Y_deaths_li, q = q.default, sea.period = 0,
  start = common_start, end = li_fit_end, calendar = cal_li
)
res_deaths_uni <- tsgc::estimate(model_deaths_uni)

tsgc::plot_compare_forecast(
  list(univariate_deaths = res_deaths_uni, leading_indicator = res_li),
  n.ahead = n.forecasts, actual = Y_deaths_common,
  title = paste0("Univariate deaths model vs cases-leading-deaths (lag = ", best_lag, ")")
)
```

Comparison is illustrative given `r length(li_by_lag)` lag candidates and
one outbreak's data; a proper lag selection would follow the PM2.5
vignette's rolling cross-validation over multiple forecast origins, not
just one holdout window, before treating any particular lag as
established.

# Zone-level illustration: Bunia

61 health zones report from varying start dates; Bunia has among the
longest, cleanest series (monotonic throughout, only flat -- never
decreasing -- steps) and is one of the outbreak's largest zone-level
case counts, so it is used here as a single-zone worked example rather
than attempting all 61 zones.

```{r zone-load}
zone_cases_raw <- read.csv("../inst/extdata/ebola/zone_cumulative_confirmed_cases__daily.csv",
                           stringsAsFactors = FALSE)
zone_cases_raw$date <- as.Date(zone_cases_raw$date)

bunia_raw <- zone_cases_raw[zone_cases_raw$nom == "Bunia", ]
bunia_raw <- bunia_raw[order(bunia_raw$date), ]

nrow(bunia_raw)
range(bunia_raw$date)
stopifnot(!is.unsorted(bunia_raw$cumulative_confirmed_cases, strictly = FALSE))
sum(diff(bunia_raw$cumulative_confirmed_cases) == 0)
```

`r sum(diff(bunia_raw$cumulative_confirmed_cases) == 0)` flat (zero
same-report-to-report-increment) steps in Bunia's series -- these are
between *report* dates, not necessarily adjacent calendar days once
regridded, so whether they produce a zero daily increment (which
`estimate()` rejects) depends on the regridding; checked directly below
once the daily calendar is built.

```{r zone-daily-grid}
bunia_conv <- build_daily_idx(bunia_raw, "cumulative_confirmed_cases")
Y_bunia    <- bunia_conv$series
cal_bunia  <- bunia_conv$calendar

# Zero increments once regridded to daily (calendar days between reports
# with no growth, not just report-to-report flats).
zero_incr <- diff(na.omit(idx_values(Y_bunia))) == 0
sum(zero_incr)
```

```{r zone-fix-flats}
# estimate() requires strictly positive non-missing increments; per the
# package's own guidance (SSModelDynamicGompertz's error message),
# eliminate flat segments with small increments rather than dropping
# those days -- an approximation, not a claim about the true
# sub-report-interval case trajectory. Same helper used for the
# national cases/deaths series above (deflatten-helper chunk).
Y_bunia_adj <- deflatten_idx_series(Y_bunia)
stopifnot(all(na.omit(diff(idx_values(Y_bunia_adj))) > 0))
```

```{r zone-fit}
rng_bunia <- idx_range(Y_bunia_adj)
bunia_fit_end <- rng_bunia[2] - n.forecasts
stopifnot(!is.na(idx_values(Y_bunia_adj[bunia_fit_end])))

model_bunia <- tsgc::SSModelDynamicGompertz(
  Y = Y_bunia_adj, q = q.default, sea.period = 0,
  start = rng_bunia[1], end = bunia_fit_end, calendar = cal_bunia
)
res_bunia <- tsgc::estimate(model_bunia)
summary(res_bunia)
tsgc::print_model_diagnostics(res_bunia)
```

```{r zone-forecast-plot, fig.cap="Bunia confirmed cases forecast, 14-day holdout."}
p <- tsgc::plot_forecast(
  res_bunia, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = bunia_fit_end - 30,
  title = "Bunia: forecast of daily new confirmed cases\n14-day holdout",
  series.name = "Bunia confirmed cases"
)
print(p)
```

```{r zone-holdout-plot, fig.cap="Bunia holdout accuracy, 14 days."}
p <- tsgc::plot_holdout(
  res_bunia, Y = Y_bunia_adj, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: Bunia confirmed cases, 14-day holdout",
  series.name = "Bunia confirmed cases"
)
print(p)
```

## Comparison: Bunia vs national

```{r zone-vs-national-compare, fig.cap="Bunia (single zone) vs national forecast, 14-day holdout, comparable origins."}
tsgc::plot_compare_forecast(
  list(national = res_cases, bunia = res_bunia),
  n.ahead = n.forecasts,
  actual = list(national = Y_cases, bunia = Y_bunia_adj),
  title = "National vs Bunia zone-level forecast (independent origins)"
)
```

National and Bunia forecast origins are not the same calendar date (each
model's `end` is its own series' last-14-days-back position), so this
overlays two forecasts on different windows rather than a like-for-like
comparison; useful for a first look at whether zone-level noise is much
higher than the national aggregate, not for concluding one model is more
accurate than the other.

# Data caveats for further work

* **Revisions**: suspected-case and suspected-death series have frequent
  downward revisions (sitrep corrections) and are not used here;
  confirmed cases/deaths show none in this data, but future sitreps may
  still revise past confirmed counts -- worth re-checking monotonicity
  whenever the data is refreshed rather than assuming it persists.
* **Zone coverage**: 61 zones, varying start dates (14-20 May), zone-level
  reporting ends 4 Aug while national reporting continues to 10 Aug --
  zone data is roughly a week stale relative to the national series as of
  this data pull.
* **`new_confirmed_cases` (zone-level)**: sparsely reported (212 rows vs
  2,261 for cumulative) -- not used here; would need its own missingness
  handling if pursued (distinct from cumulative-series `NA` gaps, since a
  missing daily-increment reading doesn't imply zero cases that day).
* **Generation interval for $R_t$**: `plot_r0()` needs a generation-interval
  distribution as an argument; the placeholder above is not fit to this
  outbreak and the `eval=FALSE` chunk should be resolved (with the
  correct package helper, if one exists, for constructing `gen_int`)
  before this section is usable.
* **Stubble-burning-style covariates**: no weather/behavioural `xpred`
  candidates attempted here (unlike PM2.5's weather regressors) --
  contact-tracing counts or hospitalisation data, if available in a
  compatible cumulative form, would be the natural equivalent.
