---
title: "Forecasting Delhi PM2.5 Pollution Surges with tsgc"
author: "Guglielmo Secchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = TRUE,
  warning    = TRUE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

`tsgc`'s dynamic Gompertz model is built for a cumulative, saturating
series: $Y_t$ strictly increasing, growth rate $\ln g_t = \ln y_t - \ln
Y_{t-1}$. Raw PM2.5 is a level series with no saturation point, so it
can't be fed to `SSModelDynamicGompertz` directly. Instead, each pollution
season is treated as its own bounded episode: cumulative PM2.5 *within* a
surge window (1 May-30 April), reset to zero at each new window, mirroring
how the package treats one wave of an epidemic. Forecasts of $y_t$
(daily PM2.5) are recovered from the fitted cumulative series via
`predict_level(..., return.diff = TRUE)`.

## Data

```{r libraries}
suppressPackageStartupMessages({
  library(tsgc)
  library(dplyr)
  library(ggplot2)
  library(ggthemes)
  library(zoo)
  library(xts)
  library(tidyr)
})

theme_set(ggthemes::theme_economist_white(gray_bg = FALSE, base_size = 14))
options(scipen = 7)

CONF_LEVEL <- 0.68
```

```{r load-data}
pm <- read.csv("../inst/Alipur_Delhi_PM_daily_2023_2025.csv", stringsAsFactors = FALSE)
pm$date <- as.Date(pm$date, format = "%d/%m/%Y")
pm <- pm[order(pm$date), ]

str(pm)
range(pm$date)
sum(is.na(pm$pm25_mean))
```

`r nrow(pm)` daily observations, `r format(min(pm$date), "%d %b %Y")` to
`r format(max(pm$date), "%d %b %Y")`, spanning four winter pollution
seasons (2023, 2024, 2025, and the still-incomplete 2026 wave).
`r sum(is.na(pm$pm25_mean))` missing PM2.5 readings
(`r round(100*sum(is.na(pm$pm25_mean))/nrow(pm),1)`%); weather covariates
complete. Stubble-burning intensity and a Diwali dummy are natural
`xpred` candidates -- Diwali is built below; stubble-burning still needs
a data source.

```{r raw-series-plot, fig.cap="Daily PM2.5 at Alipur, Delhi, 2023-2026."}
ggplot(pm, aes(x = date, y = pm25_mean)) +
  geom_line(color = "firebrick", linewidth = 0.3, na.rm = TRUE) +
  labs(title = "Daily PM2.5, Alipur station, Delhi",
       x = NULL, y = expression(paste("PM2.5 (", mu, "g/", m^3, ")"))) +
  theme(plot.title = element_text(face = "bold"))
```

Trough around July-September (roughly 25-45 $\mu g/m^3$), sharp rise
from October, winter peak typically November-December (200-275
$\mu g/m^3$), decline back to baseline by March-April.

# Constructing surge-window cumulative series

Surge window: 1 May-30 April, so the full rise-peak-decline cycle sits
inside one window, starting from the summer trough. Three complete
windows (2023-24, 2024-25, 2025-26); data ends July 2026, partway through
what would be the 2026-27 trough, so that window isn't built.

```{r surge-windows}
surge_windows <- data.frame(
  label = c("2023-24", "2024-25", "2025-26"),
  start = as.Date(c("2023-05-01", "2024-05-01", "2025-05-01")),
  end   = as.Date(c("2024-04-30", "2025-04-30", "2026-04-30"))
)
surge_windows
```

## Missing days

`estimate()` requires non-missing increments to be strictly positive
(checked via `na.omit()`), so `NA` gaps are tolerated -- `KFAS` handles
missing observations natively, skipping the update step and propagating
the state prediction across the gap. Missing days are left as `NA`
(no interpolation).

```{r check-na-clusters}
na_run_lengths <- rle(is.na(pm$pm25_mean))
table(na_run_lengths$lengths[na_run_lengths$values])
```

Isolated single-day gaps only.

```{r build-cumulative-surges}
cumsum_na <- function(x) {
  out <- rep(NA_real_, length(x))
  running <- 0
  for (i in seq_along(x)) {
    if (!is.na(x[i])) {
      running <- running + x[i]
      out[i] <- running
    }
  }
  out
}

build_surge_cumulative <- function(pm, start_date, end_date) {
  w <- pm[pm$date >= start_date & pm$date <= end_date, ]
  w$cum_pm25 <- cumsum_na(w$pm25_mean)
  w
}

surge_data <- lapply(seq_len(nrow(surge_windows)), function(i) {
  build_surge_cumulative(pm, surge_windows$start[i], surge_windows$end[i])
})
names(surge_data) <- surge_windows$label

sapply(surge_data, nrow)
sapply(surge_data, function(d) sum(is.na(d$cum_pm25)))
```

```{r surge-cumulative-plot, fig.cap="Cumulative within-season PM2.5 exposure, by surge window."}
surge_plot_df <- do.call(rbind, lapply(names(surge_data), function(nm) {
  d <- surge_data[[nm]]
  data.frame(day = seq_len(nrow(d)), cum_pm25 = d$cum_pm25, surge = nm)
}))

ggplot(surge_plot_df, aes(x = day, y = cum_pm25, color = surge)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(title = "Cumulative PM2.5 exposure within each surge window",
       subtitle = "Window: 1 May to 30 April; day 1 = 1 May. Gaps mark missing days.",
       x = "Day of surge window", y = "Cumulative PM2.5",
       color = "Surge") +
  theme(plot.title = element_text(face = "bold"))
```

Clearly S-shaped: shallow slope through summer, steepening sharply
around day 150-180 (October), flattening again by day 330-360
(March-April).

```{r idx-conversion}
make_surge_idx <- function(d) {
  x <- xts::xts(d$cum_pm25, order.by = d$date)
  conv <- xts_to_idx(x)
  list(series = conv$series, calendar = conv$calendar, raw = d)
}

surge_idx <- lapply(surge_data, make_surge_idx)
```

# Model 1: dynamic Gompertz, single surge (2024-25), no regressors

Most recent complete surge as the primary worked example: fit on data up
to a forecast origin, forecast forward, check against the actual
continuation.

```{r model1-setup}
s2425 <- surge_idx[["2024-25"]]
cal2425 <- s2425$calendar
Y2425  <- s2425$series

oct1_pos  <- idx_to_pos(cal2425, as.Date("2024-10-01"))
n.forecasts <- 14
q.default   <- 0.005
stopifnot(!is.na(idx_values(Y2425[oct1_pos])))
```

```{r model1-inspect, fig.cap="2024-25 surge window: cumulative PM2.5 exposure."}
mod_check <- tsgc::SSModelDynamicGompertz(Y = Y2425, calendar = cal2425)
p <- plot(mod_check, title = "2024-25 surge: cumulative PM2.5", series.name = "Cum. PM2.5")
print(p)
```

## Estimation from a 1 October forecast origin

```{r model1-fit}
model_oct1 <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_oct1 <- tsgc::estimate(model_oct1)
summary(res_oct1)
tsgc::print_model_diagnostics(res_oct1)
```

## Forecast and holdout accuracy

```{r model1-forecast, fig.cap="Log growth rate forecast, 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_oct1, Y = Y2425, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5 exposure\nforecast from 1 Oct 2024"
)
print(p)
```

```{r model1-forecast-level, fig.cap="New (daily) PM2.5 forecast, 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_oct1, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r model1-holdout, fig.cap="Holdout accuracy, 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_oct1, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r model1-mapes}
res_oct1$mapes(n.ahead = n.forecasts, Y = Y2425)
```

# Rolling forecast origins through a surge (2024-25)

Forecast origins across the surge: 1 Oct, 15 Oct, 30 Oct, 15 Nov, 30 Nov,
15 Dec, 31 Dec, 15 Jan, 31 Jan, 15 Feb -- refit at each, 14-day-ahead
accuracy. 31 Jan 2025 is itself a missing reading, so that origin drops
out below.

```{r rolling-origins-2425}
origin_dates_2425 <- as.Date(c(
  "2024-10-01", "2024-10-15", "2024-10-30",
  "2024-11-15", "2024-11-30",
  "2024-12-15", "2024-12-31",
  "2025-01-15", "2025-01-31",
  "2025-02-15"
))

fit_rolling_origin <- function(surge_idx_obj, origin_date, q = q.default,
                               n.ahead = 14, sea.period = 7) {
  cal <- surge_idx_obj$calendar
  Y   <- surge_idx_obj$series
  origin_pos <- idx_to_pos(cal, origin_date)
  rng <- idx_range(Y)
  if (origin_pos + n.ahead > rng[2]) {
    return(NULL)  # not enough holdout left to evaluate
  }
  if (is.na(idx_values(Y[origin_pos]))) {
    return(NULL)  # origin falls on a missing PM2.5 day; predict_level()
                  # requires a non-NA cumulative value at the fit's end
  }
  model <- tsgc::SSModelDynamicGompertz(
    Y = Y, q = q, sea.period = sea.period,
    start = rng[1], end = origin_pos, calendar = cal
  )
  res <- tryCatch(tsgc::estimate(model), error = function(e) NULL)
  if (is.null(res)) return(NULL)
  acc <- tryCatch(res$mapes(n.ahead = n.ahead, Y = Y), error = function(e) NULL)
  list(origin = origin_date, res = res, accuracy = acc)
}

rolling_2425 <- lapply(origin_dates_2425, function(d) {
  fit_rolling_origin(s2425, d)
})
names(rolling_2425) <- format(origin_dates_2425)

# Drop any origins too close to the end of the window to evaluate a full
# 14-day holdout.
rolling_2425 <- Filter(Negate(is.null), rolling_2425)
length(rolling_2425)
```

```{r rolling-2425-table}
rolling_accuracy_2425 <- do.call(rbind, lapply(names(rolling_2425), function(nm) {
  a <- rolling_2425[[nm]]$accuracy
  data.frame(origin = nm, mape = a$mape, smape = a$smape,
            mae = a$mae, rmse = a$rmse, coverage = a$coverage)
}))
rolling_accuracy_2425
```

```{r rolling-2425-plot, fig.cap="Forecast accuracy (sMAPE, coverage) by forecast origin, 2024-25 surge."}
rolling_long <- tidyr::pivot_longer(
  rolling_accuracy_2425, cols = c(smape, coverage),
  names_to = "metric", values_to = "value"
)
rolling_long$origin <- factor(rolling_long$origin, levels = rolling_accuracy_2425$origin)

ggplot(rolling_long, aes(x = origin, y = value, group = metric, color = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(title = "14-day-ahead forecast accuracy across the 2024-25 surge",
       x = "Forecast origin", y = NULL, color = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))
```

Accuracy is generally worse for origins around the peak (mid/late
November), where the growth rate turns from strongly positive to
strongly negative in a short window -- hard to forecast 14 days ahead
regardless of model. Coverage should be read loosely given the small
number of holdout days per origin.

# Applying the same rolling-origin design to all three surges

Same offsets relative to each surge's own 1 October, for 2023-24 and
2025-26 too.

```{r rolling-all-surges}
rolling_offsets <- as.numeric(origin_dates_2425 - as.Date("2024-10-01"))

origin_dates_for <- function(anchor_oct1) anchor_oct1 + rolling_offsets

all_surge_origins <- list(
  "2023-24" = origin_dates_for(as.Date("2023-10-01")),
  "2024-25" = origin_dates_for(as.Date("2024-10-01")),
  "2025-26" = origin_dates_for(as.Date("2025-10-01"))
)

rolling_all <- lapply(names(surge_idx), function(surge_nm) {
  origins <- all_surge_origins[[surge_nm]]
  res_list <- lapply(origins, function(d) fit_rolling_origin(surge_idx[[surge_nm]], d))
  names(res_list) <- format(origins)
  Filter(Negate(is.null), res_list)
})
names(rolling_all) <- names(surge_idx)

sapply(rolling_all, length)
```

```{r rolling-all-table}
rolling_accuracy_all <- do.call(rbind, lapply(names(rolling_all), function(surge_nm) {
  do.call(rbind, lapply(names(rolling_all[[surge_nm]]), function(nm) {
    a <- rolling_all[[surge_nm]][[nm]]$accuracy
    data.frame(surge = surge_nm, origin = nm, mape = a$mape, smape = a$smape,
              mae = a$mae, rmse = a$rmse, coverage = a$coverage)
  }))
}))
rolling_accuracy_all
```

```{r rolling-all-plot, fig.cap="14-day-ahead sMAPE by forecast origin, all three surges."}
rolling_accuracy_all$origin_offset <- rep(rolling_offsets, length.out = nrow(rolling_accuracy_all))

ggplot(rolling_accuracy_all, aes(x = origin_offset, y = smape, color = surge)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "14-day-ahead sMAPE across forecast origins, by surge",
       subtitle = "x-axis: days after 1 October of that surge's year",
       x = "Days after 1 October", y = "sMAPE", color = "Surge") +
  theme(plot.title = element_text(face = "bold"))
```

The degraded accuracy around the November peak recurs across all three
surges -- a structural feature of forecasting through a turning point 14
days ahead, not an artefact of one year.

# Weather regressors (`xpred`)

Four weather covariates with a clear mechanistic link to pollutant
accumulation: temperature, relative humidity, wind speed, boundary-layer
height (inversions and low BLH trap pollutants; wind disperses them).
Starting with these four of the nine available, added to the 2024-25,
1 October model.

```{r xpred-setup}
weather_cols <- c("at_c_mean", "rh_percent_mean", "ws_m_s_mean",
                  "boundary_layer_height_m_mean")

w2425 <- surge_data[["2024-25"]]
weather_xts <- xts::xts(w2425[, weather_cols], order.by = w2425$date)
weather_idx_full <- xts_to_idx(
  weather_xts,
  start.pos = idx_to_pos(cal2425, w2425$date[1])
)$series

weather_est <- get_timeframe(weather_idx_full, 1, oct1_pos)
head(weather_est)
```

```{r xpred-fit}
model_weather <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  xpred    = weather_est,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_weather <- tsgc::estimate(model_weather)
summary(res_weather)
tsgc::print_model_diagnostics(res_weather)
```

## Realised weather vs a naive forecast

Realised future weather answers "what if we'd known the weather
perfectly" -- not an operational forecast, since an archived weather
forecast (not the realised outcome) is what's actually available at the
origin. Two forecasts from the same 1 October 2024 origin:

* **Realised weather** -- actual observed weather over the 14-day
  horizon, an upper bound on what perfect foresight buys.
* **Naive persistence** -- each weather variable held at its last
  observed value for the full horizon; uses only information available
  at the origin.

```{r xpred-future}
weather_future_realised <- get_timeframe(weather_idx_full, oct1_pos + 1, oct1_pos + n.forecasts)

# Naive persistence: repeat the last observed (1 Oct) weather values for
# every day of the forecast horizon, using only information available at
# the forecast origin.
last_obs_weather <- idx_values(get_timeframe(weather_idx_full, oct1_pos, oct1_pos))
persistence_mat <- matrix(
  rep(last_obs_weather, each = n.forecasts),
  nrow = n.forecasts,
  dimnames = list(NULL, weather_cols)
)
weather_future_naive <- idx_series(persistence_mat, start = oct1_pos + 1)

# FilterResults is a reference class (setRefClass): `res_weather_realised
# <- res_weather` would alias the same underlying object, so setting
# $xpred.new on one would silently overwrite the other. Re-estimating
# independently for each weather path avoids this.
res_weather_realised <- tsgc::estimate(model_weather)
res_weather_realised$xpred.new <- weather_future_realised

res_weather_naive <- tsgc::estimate(model_weather)
res_weather_naive$xpred.new <- weather_future_naive
```

```{r xpred-forecast-plot, fig.cap="PM2.5 forecast, 1 Oct 2024 origin: realised vs naive-persistence future weather."}
p <- tsgc::plot_forecast(
  res_weather_realised, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\nwith realised future weather (upper-bound comparison)",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_forecast(
  res_weather_naive, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\nwith naive-persistence future weather (operationally realistic)",
  series.name = "PM2.5"
)
print(p)
```

```{r xpred-holdout, fig.cap="Holdout accuracy: realised vs naive-persistence future weather."}
p <- tsgc::plot_holdout(
  res_weather_realised, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: PM2.5 forecast, realised future weather",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_holdout(
  res_weather_naive, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: PM2.5 forecast, naive-persistence future weather",
  series.name = "PM2.5"
)
print(p)

weather_accuracy_compare <- rbind(
  data.frame(weather_path = "realised", as.data.frame(res_weather_realised$mapes(n.ahead = n.forecasts, Y = Y2425))),
  data.frame(weather_path = "naive_persistence", as.data.frame(res_weather_naive$mapes(n.ahead = n.forecasts, Y = Y2425)))
)
weather_accuracy_compare
```

```{r xpred-compare-paths, fig.cap="Weather-augmented forecast: realised vs naive-persistence future weather, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(realised_weather = res_weather_realised, naive_persistence_weather = res_weather_naive),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Realised vs naive-persistence future weather (2024-25, 1 Oct origin)"
)
```

Naive persistence uses only origin-available information and is the
fairer comparison against baseline below; realised weather stays as an
upper bound, not an operational result.

## Comparison: baseline vs weather-augmented

```{r xpred-compare, fig.cap="Baseline vs weather-augmented forecast, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_oct1, weather_realised = res_weather_realised,
      weather_naive_persistence = res_weather_naive),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Baseline vs weather-augmented forecast (2024-25, 1 Oct origin)"
)
```

# Leading-indicator model: boundary-layer height and PM2.5

Boundary-layer height (BLH) is a mechanistic lead for pollutant
accumulation: a shallow boundary layer traps pollutants near the
surface, and BLH contraction in autumn typically precedes the sharpest
rise in measured PM2.5. `SSModelLeadingIndicator` requires both series
in `Y` to be strictly increasing cumulative series on a shared position
axis, so BLH itself (not cumulative, not monotone) can't be used
directly -- it is first transformed to a positive, cumulable series by
inverting it (low BLH -> high value) and cumulating that within the
2024-25 surge window, exactly as PM2.5 itself was cumulated above.

`SSModelLeadingIndicator$estimate()` additionally requires the lead and
target series, after lagging, to leave a single *contiguous* block of
finite rows -- unlike `SSModelDynamicGompertz`, it does not fall back to
KFAS's native handling of interior missing observations. The isolated
single-day gaps identified earlier are therefore linearly interpolated
here, but only for the purposes of constructing the lead/target pair
`Y_li` used in this section; `Y2425` (used by the baseline Gompertz and
weather models elsewhere in this vignette) is left with its NAs intact.

The inverse-BLH series is scaled by 1000 (rather than using `1/BLH`
directly) so its magnitude is closer to that of cumulative PM2.5; this
is good practice for numerical stability but is a separate issue from
the very wide forecast intervals seen further below -- see the
diagnostic chunk after model estimation for the actual source of that.

```{r li-blh-transform}
blh_raw  <- zoo::na.approx(w2425$boundary_layer_height_m_mean, na.rm = FALSE)
pm25_raw <- zoo::na.approx(w2425$pm25_mean, na.rm = FALSE)

inv_blh <- 1000 / blh_raw
cum_inv_blh <- cumsum_na(inv_blh)
cum_pm25_li <- cumsum_na(pm25_raw)

li_df <- data.frame(
  date = w2425$date,
  lead = cum_inv_blh,
  targ = cum_pm25_li
)

li_xts <- xts::xts(li_df[, c("lead", "targ")], order.by = li_df$date)
li_conv <- xts_to_idx(li_xts)
Y_li <- li_conv$series
cal_li <- li_conv$calendar
```

## Preview: comovement of inverse-BLH and PM2.5

The built-in `plot.SSModelLeadingIndicator` overlays both series on a
single (log) axis. Because inverse-BLH and cumulative PM2.5 live on
very different scales, a dual-axis plot is more informative here.

```{r li-preview-plot, fig.cap="Cumulative inverse boundary-layer height (lead) vs cumulative PM2.5 (target), 2024-25 surge."}
n.lag.li <- 5

li_plot_df <- data.frame(
  date        = li_df$date,
  lead_series = li_df$lead,
  targ_series = li_df$targ
)

lead_range <- range(li_plot_df$lead_series, na.rm = TRUE)
targ_range <- range(li_plot_df$targ_series, na.rm = TRUE)
scale_factor <- diff(targ_range) / diff(lead_range)
offset <- targ_range[1] - lead_range[1] * scale_factor

p <- ggplot(li_plot_df, aes(x = date)) +
  geom_line(aes(y = lead_series * scale_factor + offset, color = "Cum. inverse BLH"),
            linewidth = 0.7, na.rm = TRUE) +
  geom_line(aes(y = targ_series, color = "Cum. PM2.5"),
            linewidth = 0.7, na.rm = TRUE) +
  scale_y_continuous(
    name = "Cumulative PM2.5",
    sec.axis = sec_axis(~ (. - offset) / scale_factor, name = "Cumulative inverse BLH")
  ) +
  scale_color_manual(values = c("Cum. inverse BLH" = "steelblue", "Cum. PM2.5" = "firebrick")) +
  labs(title = "Cumulative inverse BLH (lead) vs cumulative PM2.5 (target)",
       x = NULL, color = NULL) +
  theme(plot.title = element_text(face = "bold"),
        axis.title.y.left = element_text(color = "firebrick"),
        axis.title.y.right = element_text(color = "steelblue"))
print(p)
```

## Estimation from the 1 October forecast origin

```{r li-fit}
model_li <- SSModelLeadingIndicator$new(
  Y          = Y_li,
  n.lag      = n.lag.li,
  LeadIndCol = 1,
  sea.period = 7,
  start      = 1,
  end        = oct1_pos,
  calendar   = cal_li
)
res_li <- estimate(model_li)
summary(res_li)
tsgc::print_model_diagnostics(res_li)
```

Left free, `q` is estimated at 0.3144 here -- roughly 60x
`q.default = 0.005`, the value fixed throughout the rest of this
vignette (including the vanilla Gompertz comparisons). A state
variance that large lets the log-growth-rate trend wander freely
between observations, and a random-walk state's forecast-interval
width grows with horizon; compounded through `cumprod()` on the
exponentiated scale, a linear widening in log-space becomes an
explosive one in level-space. The forecast and holdout plots below,
from this freely-estimated fit, show that explosion directly rather
than describing it.

```{r li-forecast-log, fig.cap="Log growth rate forecast, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_li, Y = Y_li, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5\nleading-indicator forecast from 1 Oct 2024 (q estimated freely)"
)
print(p)
```

```{r li-forecast-level, fig.cap="Daily PM2.5 forecast, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment (leading-indicator model, q estimated freely)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-holdout, fig.cap="Holdout accuracy, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_li, Y = Y_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast (leading-indicator model, q estimated freely)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-mapes}
res_li$mapes(n.ahead = n.forecasts, Y = Y_li)
```

## Fixing q at the package default

The freely-estimated `q` above is itself informative: with only
~150 days of estimation data, this particular leading-indicator
specification weakly identifies its own state variance, and left
unconstrained it produces forecasts that are technically correct but
practically useless past about day 6-7. Fixing `q = q.default`, as
already done for the vanilla Gompertz model throughout this vignette,
keeps the leading-indicator forecast on a comparable, usable scale.

```{r li-fit-qfixed}
model_li_qfixed <- SSModelLeadingIndicator$new(
  Y          = Y_li,
  n.lag      = n.lag.li,
  LeadIndCol = 1,
  sea.period = 7,
  q          = q.default,
  start      = 1,
  end        = oct1_pos,
  calendar   = cal_li
)
res_li_qfixed <- estimate(model_li_qfixed)
summary(res_li_qfixed)
```

```{r li-forecast-log-qfixed, fig.cap="Log growth rate forecast, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_li_qfixed, Y = Y_li, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5\nleading-indicator forecast from 1 Oct 2024 (q fixed at default)"
)
print(p)
```

```{r li-forecast-level-qfixed, fig.cap="Daily PM2.5 forecast, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_li_qfixed, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment (leading-indicator model, q fixed at default)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-holdout-qfixed, fig.cap="Holdout accuracy, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_li_qfixed, Y = Y_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast (leading-indicator model, q fixed at default)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-mapes-qfixed}
res_li_qfixed$mapes(n.ahead = n.forecasts, Y = Y_li)
```

The remainder of this section (and the lag/q sensitivity analysis
further below) uses `res_li_qfixed`, so its plots stay on a scale
comparable to the vanilla Gompertz baseline.

## Comparison: baseline vs leading-indicator forecast

`plot_compare_forecast()`'s `actual` overlay assumes a `FilterResults`
object (it looks for an `index` field that `FilterResultsLI` doesn't
have), so it can only be passed when every model in the list is a plain
`SSModelDynamicGompertz` fit. Comparing a `FilterResults` and a
`FilterResultsLI` side by side therefore drops `actual` here; the actual
values are already visible in the individual forecast/holdout plots
above.

```{r li-compare, fig.cap="Baseline dynamic Gompertz vs BLH leading-indicator forecast (q fixed at default), 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_oct1, leading_indicator_blh = res_li_qfixed),
  n.ahead = n.forecasts,
  title = "Baseline vs leading-indicator (BLH) forecast (2024-25, 1 Oct origin)"
)
```

## Lag selection and `q` sensitivity via rolling cross-validation

The lag of 5 days above was chosen by inspection. This section sweeps
lag and `q` jointly on a grid, so the effect of each can be read off
independently as well as in combination, using `cross_val()`'s
disjoint selection and reporting folds so the reported accuracy of the
chosen specification isn't inflated by having been used to pick it.

```{r li-cv-setup}
n.ahead.cv.li  <- 5
gap.cv.li      <- n.ahead.cv.li
n.select.cv.li <- 2
n.report.cv.li <- 1

report.end.cv.li <- oct1_pos - (n.report.cv.li - 1) * gap.cv.li - n.ahead.cv.li
select.end.cv.li <- report.end.cv.li - n.select.cv.li * gap.cv.li - n.ahead.cv.li
stopifnot(select.end.cv.li > 1)

max_lag_cv_li <- 10
stopifnot(select.end.cv.li - 1 > max_lag_cv_li + 1)
```

### Building the lag × q grid

```{r li-cv-models}
q_candidates <- c(0.001, 0.005, 0.01, 0.02, 0.05)
lag_candidates <- c(3, 5, 7, 10)

cv_models_li <- list()

cv_models_li[["Vanilla_q_default"]] <- SSModelDynamicGompertz$new(
  Y = Y2425, q = q.default, sea.period = 7,
  start = 1, end = select.end.cv.li, calendar = cal2425
)

for (qv in q_candidates) {
  cv_models_li[[paste0("Vanilla_q", qv)]] <- SSModelDynamicGompertz$new(
    Y = Y2425, q = qv, sea.period = 7,
    start = 1, end = select.end.cv.li, calendar = cal2425
  )
}

grid_spec_li <- expand.grid(lag = lag_candidates, q = q_candidates)

for (i in seq_len(nrow(grid_spec_li))) {
  lg <- grid_spec_li$lag[i]
  qv <- grid_spec_li$q[i]
  nm <- paste0("LI_lag", lg, "_q", qv)
  cv_models_li[[nm]] <- SSModelLeadingIndicator$new(
    Y = Y_li, n.lag = lg, LeadIndCol = 1, sea.period = 7,
    q = qv, start = 1, end = select.end.cv.li, calendar = cal_li
  )
}
```

### Selection-fold results: full grid

```{r li-cv-select}
cv_selection_li <- cross_val(
  Y = Y_li,
  model_list = cv_models_li,
  est.end = select.end.cv.li,
  criterion = "smape",
  n.ahead = n.ahead.cv.li,
  n.estimate = n.select.cv.li,
  gap = gap.cv.li
)
cv_selection_li$mean_smape <- rowMeans(cv_selection_li[, -1, drop = FALSE])
cv_selection_li[order(cv_selection_li$mean_smape), c("Model", "mean_smape")]
```

### Lag × q grid, reshaped

The table above is easiest to read reshaped so that lag runs down the
rows and `q` runs across the columns, restricted to the leading-indicator
models (the vanilla Gompertz rows are kept separately as a benchmark
below).

```{r li-cv-grid-table}
li_only <- cv_selection_li[grepl("^LI_lag", cv_selection_li$Model), c("Model", "mean_smape")]
li_only$lag <- as.integer(sub("^LI_lag(\\d+)_q.*$", "\\1", li_only$Model))
li_only$q   <- as.numeric(sub("^LI_lag\\d+_q", "", li_only$Model))

grid_table_li <- reshape(
  li_only[, c("lag", "q", "mean_smape")],
  idvar = "lag", timevar = "q", direction = "wide"
)
names(grid_table_li) <- sub("mean_smape\\.", "q=", names(grid_table_li))
grid_table_li <- grid_table_li[order(grid_table_li$lag), ]
grid_table_li
```

### Isolating the effect of lag (at the package default `q`)

```{r li-cv-lag-effect}
lag_effect_li <- li_only[li_only$q == q.default, c("lag", "mean_smape")]
lag_effect_li <- lag_effect_li[order(lag_effect_li$lag), ]
lag_effect_li
```

```{r li-lag-effect-plot, fig.cap="Selection-fold SMAPE by lag, leading-indicator model at q = 0.005."}
p <- ggplot(lag_effect_li, aes(x = lag, y = mean_smape)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Selection-fold SMAPE by lag (q = 0.005)",
       x = "Lag (days)", y = "Mean SMAPE") +
  theme(plot.title = element_text(face = "bold"))
print(p)
```

### Isolating the effect of `q` (at the best lag found above)

```{r li-cv-q-effect}
best_lag_li <- lag_effect_li$lag[which.min(lag_effect_li$mean_smape)]
best_lag_li

q_effect_li <- li_only[li_only$lag == best_lag_li, c("q", "mean_smape")]
q_effect_li <- q_effect_li[order(q_effect_li$q), ]
q_effect_li

vanilla_rows_li <- cv_selection_li[grepl("^Vanilla_q", cv_selection_li$Model), c("Model", "mean_smape")]
vanilla_rows_li$q <- ifelse(
  vanilla_rows_li$Model == "Vanilla_q_default", q.default,
  as.numeric(sub("^Vanilla_q", "", vanilla_rows_li$Model))
)
vanilla_rows_li <- vanilla_rows_li[order(vanilla_rows_li$q), c("q", "mean_smape")]
vanilla_rows_li
```

```{r li-q-effect-plot, fig.cap="Selection-fold SMAPE across q, leading-indicator model vs vanilla Gompertz."}
q_effect_li$model <- paste0("LI_lag", best_lag_li)
vanilla_rows_li$model <- "Vanilla"
q_compare_li <- rbind(q_effect_li, vanilla_rows_li)

p <- ggplot(q_compare_li, aes(x = q, y = mean_smape, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(title = paste0("Selection-fold SMAPE across q (LI at lag = ", best_lag_li, ")"),
       x = "q (log scale)", y = "Mean SMAPE", color = NULL) +
  theme(plot.title = element_text(face = "bold"))
print(p)
```

### Reporting-fold accuracy of the selected specification

Having picked `best_lag_li` and the `q` that minimised selection-fold
SMAPE from the tables above, the reporting fold gives a final,
held-out accuracy check against the vanilla Gompertz benchmark.

```{r li-cv-report}
best_q_li <- q_effect_li$q[which.min(q_effect_li$mean_smape)]
selected_spec_li <- paste0("LI_lag", best_lag_li, "_q", best_q_li)
selected_spec_li

cv_report_models_li <- cv_models_li[c("Vanilla_q_default", selected_spec_li)]

cv_reporting_li <- cross_val(
  Y = Y_li,
  model_list = cv_report_models_li,
  est.end = report.end.cv.li,
  criterion = "smape",
  n.ahead = n.ahead.cv.li,
  n.estimate = n.report.cv.li,
  gap = gap.cv.li
)
cv_reporting_li
```

## Forecast overlay across `q`

```{r li-q-overlay-fit}
res_li_q_list <- lapply(q_candidates, function(qv) {
  m <- SSModelLeadingIndicator$new(
    Y = Y_li, n.lag = best_lag_li, LeadIndCol = 1, sea.period = 7,
    q = qv, start = 1, end = oct1_pos, calendar = cal_li
  )
  estimate(m)
})
names(res_li_q_list) <- paste0("q_", q_candidates)
```

As above, `actual` is omitted from this overlay since every model here
is a `FilterResultsLI` object and `plot_compare_forecast()`'s `actual`
argument does not support that class.

```{r li-q-overlay-plot, fig.cap="Leading-indicator forecast of daily PM2.5 across candidate q values, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  res_li_q_list,
  n.ahead = n.forecasts,
  title = paste0("Leading-indicator (BLH) forecast across q, 2024-25, 1 Oct origin, lag = ", best_lag_li)
)
```

# Reinitialisation for a single new surge

`reinit.idx` restarts the cumulative series at zero at a chosen position,
optionally carrying forward the prior state via `use.presample.info`.
Treating the 1 October transition into the 2024-25 surge as a
reinitialisation point: fit a longer window starting in the preceding
(2023-24) surge's tail, reinitialising at 2024-25's own 1 October start.

```{r reinit-single-setup}
# Build one continuous cumulative PM2.5 series spanning the tail of the
# 2023-24 surge through the 2024-25 surge, so a single reinit.idx can mark
# the transition between them.
span_start <- as.Date("2024-07-01")
span_end   <- surge_windows$end[surge_windows$label == "2024-25"]
w_span <- pm[pm$date >= span_start & pm$date <= span_end, ]
w_span$cum_pm25 <- cumsum_na(w_span$pm25_mean)

span_xts <- xts::xts(w_span$cum_pm25, order.by = w_span$date)
span_conv <- xts_to_idx(span_xts)
Y_span <- span_conv$series
cal_span <- span_conv$calendar

reinit_pos <- idx_to_pos(cal_span, as.Date("2024-10-01"))
span_end_pos <- idx_range(Y_span)[2]
long_end     <- span_end_pos - n.forecasts  # leave a holdout for plot_compare_forecast
```

```{r reinit-single-fit}
model_span_base <- tsgc::SSModelDynamicGompertz(
  Y = Y_span, q = q.default, start = 1, end = long_end, calendar = cal_span
)
res_span_base <- tsgc::estimate(model_span_base)

model_span_reinit <- tsgc::SSModelDynamicGompertz(
  Y = Y_span, q = q.default, start = 1, end = long_end,
  reinit.idx = reinit_pos, calendar = cal_span
)
res_span_reinit <- tsgc::estimate(model_span_reinit)
summary(res_span_reinit)
tsgc::print_model_diagnostics(res_span_reinit)
```

```{r reinit-single-compare, fig.cap="Reinitialised vs non-reinitialised forecast at the 2024-25 surge onset."}
tsgc::plot_compare_forecast(
  list(no_reinit = res_span_base, reinit = res_span_reinit),
  n.ahead = n.forecasts, actual = Y_span,
  title = "Effect of reinitialising at 1 Oct 2024 (surge onset)"
)
```

# Continual (chained) reinitialisation across all observed surges

Chaining reinitialisation at every surge onset, rather than fitting each
surge independently as above.

`reinit.idx` only takes one restart position per model, so this is a
manual chain: 1 October of each subsequent year passed as `reinit.idx`
in turn, re-basing the growth-rate target to zero each time, while
`start` stays at the series start throughout -- `use.presample.info =
TRUE` internally re-estimates a presample fit over `[start, reinit.idx]`
to seed the post-reinit state, which needs real history before
`reinit.idx`, not `start == reinit.idx`. `end` advances through the
boundaries (Oct 2023, Oct 2024, final origin below); what changes
segment to segment is the rebasing point and how far the fit extends,
not the span of history informing it. (`original.results` isn't reused
across segments -- valid only at the exact `reinit.idx` it was estimated
for; each segment lets `use.presample.info` re-derive its own presample
fit instead.)

```{r chained-reinit-setup}
full_xts <- xts::xts(pm$pm25_mean, order.by = pm$date)
full_conv <- xts_to_idx(full_xts)
Y_full <- full_conv$series
cal_full <- full_conv$calendar
Y_full_cum <- idx_series(cumsum_na(idx_values(Y_full)), start = Y_full$start)

reinit_dates <- as.Date(c("2023-10-01", "2024-10-01", "2025-10-01"))
reinit_positions <- sapply(reinit_dates, function(d) idx_to_pos(cal_full, d))
reinit_positions
```

```{r chained-reinit-fit}
chain_reinit_forward <- function(Y, calendar, reinit_positions, q = q.default,
                                 sea.period = 7, end = idx_range(Y)[2]) {
  series_start <- idx_range(Y)[1]
  results <- list()

  boundaries <- c(reinit_positions, end)

  for (i in seq_along(boundaries)) {
    seg_end <- boundaries[i]
    if (i == 1) {
      # First segment: no reinitialisation, plain fit from the series
      # start to the first reinit point.
      model <- tsgc::SSModelDynamicGompertz(
        Y = Y, q = q, sea.period = sea.period,
        start = series_start, end = seg_end, calendar = calendar
      )
      res <- tsgc::estimate(model)
    } else {
      # reinit.idx restarts the growth-rate target (Y is re-based to zero
      # from this point on), but SSModelDynamicGompertz's own
      # use.presample.info logic internally re-estimates a presample fit
      # over [start, reinit.idx] to seed the post-reinit state -- this
      # requires start to be well before reinit.idx, not equal to it, so
      # start is kept at series_start for every segment (matching the
      # package's own reinit.idx examples) rather than advanced to the
      # previous boundary.
      reinit_pos_i <- reinit_positions[i - 1]
      model <- tsgc::SSModelDynamicGompertz(
        Y = Y, q = q, sea.period = sea.period,
        start = series_start, end = seg_end,
        reinit.idx = reinit_pos_i,
        use.presample.info = TRUE,
        calendar = calendar
      )
      res <- tsgc::estimate(model)
    }
    results[[i]] <- res
  }
  results
}

final_segment_end <- idx_to_pos(cal_full, as.Date("2025-10-01"))  # match the independent 2025-26 model's own origin
reinit_positions_for_chain <- reinit_positions[reinit_positions < final_segment_end]

chained_results <- chain_reinit_forward(
  Y_full_cum, cal_full, reinit_positions_for_chain,
  end = final_segment_end
)
length(chained_results)
```

```{r chained-reinit-summary}
invisible(lapply(seq_along(chained_results), function(i) {
  cat("--- Segment", i, "---\n")
  print(summary(chained_results[[i]]))
}))
```

## Comparing continual reinitialisation to independent per-surge models

```{r chained-vs-independent-forecast, fig.cap="14-day forecast from 1 Oct 2025: chained (multi-surge reinitialisation) vs a model fitted only on the 2025-26 surge."}
# "Independent" here means: re-cumulate PM2.5 from scratch at the 2025-26
# surge's own onset (1 May 2025), using none of the prior surges' data or
# reinitialisation chain -- but still positioned on cal_full so its
# forecast lines up on the same axis as the chained model and Y_full_cum.
surge_2025_26_start <- surge_windows$start[surge_windows$label == "2025-26"]
start_pos_2526_full <- idx_to_pos(cal_full, surge_2025_26_start)
oct1_2526_pos_full  <- idx_to_pos(cal_full, as.Date("2025-10-01"))

w2526 <- pm[pm$date >= surge_2025_26_start & pm$date <= as.Date("2025-10-01"), ]
Y_2526_only <- idx_series(cumsum_na(w2526$pm25_mean), start = start_pos_2526_full)

model_2526_indep <- tsgc::SSModelDynamicGompertz(
  Y = Y_2526_only, q = q.default, sea.period = 7,
  start = start_pos_2526_full, end = oct1_2526_pos_full, calendar = cal_full
)
res_2526_indep <- tsgc::estimate(model_2526_indep)

final_chained <- chained_results[[length(chained_results)]]

tsgc::plot_compare_forecast(
  list(chained = final_chained, independent = res_2526_indep),
  n.ahead = n.forecasts, actual = Y_full_cum,
  title = "Chained (continually reinitialised) vs independently fitted 2025-26 model"
)
```

Not a difference in data span -- every chained segment's presample still
draws on the full history -- but in where the growth-rate series is
rebased and what prior seeds the post-reinit state. Independent models
start from a diffuse prior on that surge's own data; the chained model
carries forward the terminal slope/level state from a presample fit over
all preceding history. Whether that improves accuracy is worth checking
again once more surges accumulate; illustrative rather than conclusive
with three.

# Extensions: a Diwali festival regressor

Firecracker use around Diwali falls within the October-November window
where forecast accuracy has been weakest above. Diwali dummy from public
festival dates, added as `xpred` alongside the weather covariates.
Stubble-burning is a natural companion (residue burning in Punjab
overlaps the Diwali period) but no proxy is built yet -- a next step,
e.g. from satellite fire-count products.

## Constructing the Diwali dummy

Diwali dates for the years spanned by this dataset are fixed calendar
facts (India's National Informatics Centre / standard public references):

```{r diwali-dates}
diwali_dates <- as.Date(c("2023-11-12", "2024-11-01", "2025-10-20", "2026-11-08"))
names(diwali_dates) <- c("2023", "2024", "2025", "2026")
diwali_dates
```

Coded as 1 over a short window centred on each Diwali date: two days
before through two days after (five days total), to catch both build-up
and immediate aftermath rather than a single-day spike.

```{r diwali-dummy}
diwali_window <- 2L  # days either side of Diwali itself

pm$diwali_dummy <- 0L
for (d in diwali_dates) {
  in_window <- pm$date >= (d - diwali_window) & pm$date <= (d + diwali_window)
  pm$diwali_dummy[in_window] <- 1L
}

table(pm$diwali_dummy)
```

```{r diwali-plot, fig.cap="Diwali windows against daily PM2.5, 2023-2026."}
ggplot(pm, aes(x = date, y = pm25_mean)) +
  geom_line(color = "grey40", linewidth = 0.3, na.rm = TRUE) +
  geom_point(data = pm[pm$diwali_dummy == 1L, ], color = "darkorange", size = 1.2, na.rm = TRUE) +
  labs(title = "Diwali windows (orange) against daily PM2.5",
       x = NULL, y = expression(paste("PM2.5 (", mu, "g/", m^3, ")"))) +
  theme(plot.title = element_text(face = "bold"))
```

## Estimating with the Diwali regressor

```{r diwali-xpred-setup}
w2425 <- surge_data[["2024-25"]]
w2425$diwali_dummy <- pm$diwali_dummy[match(w2425$date, pm$date)]

diwali_weather_cols <- c(weather_cols, "diwali_dummy")
diwali_xreg_xts <- xts::xts(w2425[, diwali_weather_cols], order.by = w2425$date)
diwali_xreg_idx_full <- xts_to_idx(
  diwali_xreg_xts,
  start.pos = idx_to_pos(cal2425, w2425$date[1])
)$series

diwali_xreg_est <- get_timeframe(diwali_xreg_idx_full, 1, oct1_pos)
```

```{r diwali-xpred-fit}
model_diwali <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  xpred    = diwali_xreg_est,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_diwali <- tsgc::estimate(model_diwali)
summary(res_diwali)
tsgc::print_model_diagnostics(res_diwali)
```

Diwali 2024-25 fell on 1 November. Estimating from a 1-14 October origin
leaves the dummy at zero throughout the estimation window, since Diwali
falls entirely in the forecast horizon rather than the estimation
sample; the dummy's coefficient is then not identified by the data at
all and is carried forward purely from KFAS's diffuse initialisation.
This is visible above as `xregdiwali_dummy` estimating at exactly 0
with a 0 standard error in `res_diwali`, and produces the "model is
degenerate" warnings in `KFS()` -- a constant (all-zero) regressor
column gives a degenerate diffuse prior for that state, not a fitting
failure to be silenced.

To let the model actually learn the Diwali effect from data, the
estimation window is moved to 10 November -- a few days after the
1 November +/-2-day Diwali window (which runs 30 Oct-3 Nov) closes, so
those non-zero dummy rows sit inside the estimation sample rather than
the forecast horizon. The 14-day forecast horizon from this origin
(11-24 Nov) then tests the post-Diwali decay rather than asking the
model to predict the festival spike itself from an unidentified
coefficient.

```{r diwali-forecast-origin}
nov10_pos <- idx_to_pos(cal2425, as.Date("2024-11-10"))
stopifnot(!is.na(idx_values(Y2425[nov10_pos])))

model_baseline_nov10 <- tsgc::SSModelDynamicGompertz(
  Y = Y2425, q = q.default, sea.period = 7,
  start = 1, end = nov10_pos, calendar = cal2425
)
res_baseline_nov10 <- tsgc::estimate(model_baseline_nov10)

diwali_xreg_est_nov10 <- get_timeframe(diwali_xreg_idx_full, 1, nov10_pos)
table(diwali_xreg_est_nov10$data[, "diwali_dummy"])

model_diwali_nov10 <- tsgc::SSModelDynamicGompertz(
  Y = Y2425, xpred = diwali_xreg_est_nov10, q = q.default, sea.period = 7,
  start = 1, end = nov10_pos, calendar = cal2425
)
res_diwali_nov10 <- tsgc::estimate(model_diwali_nov10)
summary(res_diwali_nov10)

diwali_xreg_future <- get_timeframe(diwali_xreg_idx_full, nov10_pos + 1, nov10_pos + n.forecasts)
res_diwali_nov10$xpred.new <- diwali_xreg_future
```

```{r diwali-forecast-plot, fig.cap="14-day forecast from 10 Nov 2024 (post-Diwali decay): baseline vs weather+Diwali."}
p <- tsgc::plot_forecast(
  res_diwali_nov10, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = nov10_pos - 30,
  title = "Forecast of daily PM2.5 increment\nweather + Diwali dummy, from 10 Nov 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r diwali-holdout-compare, fig.cap="Holdout accuracy, 10 Nov 2024 origin: baseline vs weather+Diwali."}
p <- tsgc::plot_holdout(
  res_baseline_nov10, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: baseline, 10 Nov 2024 origin (Diwali in estimation window)",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_holdout(
  res_diwali_nov10, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: weather + Diwali dummy, 10 Nov 2024 origin",
  series.name = "PM2.5"
)
print(p)

diwali_accuracy_compare <- rbind(
  data.frame(model = "baseline", as.data.frame(res_baseline_nov10$mapes(n.ahead = n.forecasts, Y = Y2425))),
  data.frame(model = "weather_plus_diwali", as.data.frame(res_diwali_nov10$mapes(n.ahead = n.forecasts, Y = Y2425)))
)
diwali_accuracy_compare
```

```{r diwali-compare-plot, fig.cap="Baseline vs weather+Diwali forecast, 10 Nov 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_baseline_nov10, weather_plus_diwali = res_diwali_nov10),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Baseline vs weather + Diwali dummy (2024-25, 10 Nov origin, post-Diwali)"
)
```

The `r 2*diwali_window + 1`-day window is a starting specification, not a
validated one -- worth testing sensitivity to width, and once a
stubble-burning proxy exists, comparing Diwali-only against both
regressors together. Punjab's stubble-burning season runs roughly
mid-October to end of November, overlapping all three observed Diwali
dates (12 Nov 2023, 1 Nov 2024, 20 Oct 2025), so the two effects may be
hard to separate with only three occurrences.