---
title: "Forecasting Delhi PM2.5 Pollution Surges with tsgc"
author: "Guglielmo Secchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = TRUE,
  warning    = TRUE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

We want a strictly increasing cumulative, saturating time series, so we 
consider cumulative daily PM2.5 levels as a measurement of "total pollution
experienced". Because there is a repeating pattern of pollution spikes, we
initially isolate three pollution seasons in the data (1 May - 30 April) 
and consider cumulative pollution from 1 May each time. This is similar to 
analysing a single wave of an epidemic, though we later look at chaining 
reinitialisations (by defalt `SSModelLeadingIndicator` supports a single 
reinitialisation) and the effect of using cumulative pollution across the 
entire dataset in a single model rather than specifying surges.

## Data

```{r libraries}
suppressPackageStartupMessages({
  library(tsgc)
  library(dplyr)
  library(ggplot2)
  library(ggthemes)
  library(zoo)
  library(xts)
  library(tidyr)
})

theme_set(ggthemes::theme_economist_white(gray_bg = FALSE, base_size = 14))
options(scipen = 7)

CONF_LEVEL <- 0.68
```

```{r load-data}
pm <- read.csv("../inst/Alipur_Delhi_PM_daily_2023_2025.csv", stringsAsFactors = FALSE)
pm$date <- as.Date(pm$date, format = "%d/%m/%Y")
pm <- pm[order(pm$date), ]

str(pm)
range(pm$date)
sum(is.na(pm$pm25_mean))
```

The data has `r nrow(pm)` daily observations, spanning four winter pollution
seasons (2023, 2024, 2025, and the still-incomplete 2026 wave).
However, it has `r sum(is.na(pm$pm25_mean))` missing PM2.5 readings
(`r round(100*sum(is.na(pm$pm25_mean))/nrow(pm),1)`%), though weather covariates
complete. Stubble-burning intensity and a Diwali dummy are natural
`xpred` candidates -- Diwali is built below, whilst stubble-burning still needs
a data source.

```{r raw-series-plot, fig.cap="Daily PM2.5 at Alipur, Delhi, 2023-2026."}
ggplot(pm, aes(x = date, y = pm25_mean)) +
  geom_line(color = "firebrick", linewidth = 0.3, na.rm = TRUE) +
  labs(title = "Daily PM2.5, Alipur station, Delhi",
       x = NULL, y = expression(paste("PM2.5 (", mu, "g/", m^3, ")"))) +
  theme(plot.title = element_text(face = "bold"))
```

We see a pollution trough around July-September (roughly 25-45 $\mu g/m^3$), 
a sharp rise from October, and a winter peak typically November-December 
(200-275$\mu g/m^3$), with a decline back to baseline levels by March-April.

# Constructing surge-window cumulative series

The surge window is qualitatively chosen as 1 May-30 April, so the full 
rise-peak-decline cycle sits inside one window, starting from the summer 
trough. The data ends in July 2026, so the 2026-2027 window is not used and 
we are left with three complete sets of data (2023-24 to 2025-26).

```{r surge-windows}
surge_windows <- data.frame(
  label = c("2023-24", "2024-25", "2025-26"),
  start = as.Date(c("2023-05-01", "2024-05-01", "2025-05-01")),
  end   = as.Date(c("2024-04-30", "2025-04-30", "2026-04-30"))
)
surge_windows
```

## Missing days

A brief check shows that most sets of missing days are relatively short (1 or 
2 days, with an irrelevant number of larger missing sets (the longest is 8 
days, occuring only once).

```{r check-na-clusters}
na_run_lengths <- rle(is.na(pm$pm25_mean))
table(na_run_lengths$lengths[na_run_lengths$values])
```

```{r build-cumulative-surges}
cumsum_na <- function(x) {
  out <- rep(NA_real_, length(x))
  running <- 0
  for (i in seq_along(x)) {
    if (!is.na(x[i])) {
      running <- running + x[i]
      out[i] <- running
    }
  }
  out
}

build_surge_cumulative <- function(pm, start_date, end_date) {
  w <- pm[pm$date >= start_date & pm$date <= end_date, ]
  w$cum_pm25 <- cumsum_na(w$pm25_mean)
  w
}

surge_data <- lapply(seq_len(nrow(surge_windows)), function(i) {
  build_surge_cumulative(pm, surge_windows$start[i], surge_windows$end[i])
})
names(surge_data) <- surge_windows$label

sapply(surge_data, nrow)
sapply(surge_data, function(d) sum(is.na(d$cum_pm25)))
```

```{r surge-cumulative-plot, fig.cap="Cumulative within-season PM2.5 exposure, by surge window."}
surge_plot_df <- do.call(rbind, lapply(names(surge_data), function(nm) {
  d <- surge_data[[nm]]
  data.frame(day = seq_len(nrow(d)), cum_pm25 = d$cum_pm25, surge = nm)
}))

ggplot(surge_plot_df, aes(x = day, y = cum_pm25, color = surge)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(title = "Cumulative PM2.5 exposure within each surge window",
       subtitle = "Window: 1 May to 30 April; day 1 = 1 May. Gaps mark missing days.",
       x = "Day of surge window", y = "Cumulative PM2.5",
       color = "Surge") +
  theme(plot.title = element_text(face = "bold"))
```

A qualitative view of cumulative pollution in the surges supports the selection 
of the start day for the year-long windows, with the surge relatively in the 
middle of the period. We can thus begin to construct models.

```{r idx-conversion}
make_surge_idx <- function(d) {
  x <- xts::xts(d$cum_pm25, order.by = d$date)
  conv <- xts_to_idx(x)
  list(series = conv$series, calendar = conv$calendar, raw = d)
}

surge_idx <- lapply(surge_data, make_surge_idx)
```

# Model 1: dynamic Gompertz, single surge (2024-25), no regressors

For the most basic model, we use the 2024-25 surge with no additional 
regressors. We can then look at how it performs against the actual data.

```{r model1-setup}
s2425 <- surge_idx[["2024-25"]]
cal2425 <- s2425$calendar
Y2425  <- s2425$series

oct1_pos  <- idx_to_pos(cal2425, as.Date("2024-10-01"))
n.forecasts <- 14
q.default   <- 0.005
stopifnot(!is.na(idx_values(Y2425[oct1_pos])))
```

```{r model1-inspect, fig.cap="2024-25 surge window: cumulative PM2.5 exposure."}
mod_check <- tsgc::SSModelDynamicGompertz(Y = Y2425, calendar = cal2425)
p <- plot(mod_check, title = "2024-25 surge: cumulative PM2.5", series.name = "Cum. PM2.5")
print(p)
```

## Estimation from a 1 October forecast origin

```{r model1-fit}
model_oct1 <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_oct1 <- tsgc::estimate(model_oct1)
summary(res_oct1)
tsgc::print_model_diagnostics(res_oct1)
```

## Forecast and holdout accuracy

```{r model1-forecast, fig.cap="Log growth rate forecast, 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_oct1, Y = Y2425, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5 exposure\nforecast from 1 Oct 2024"
)
print(p)
```

```{r model1-forecast-level, fig.cap="New (daily) PM2.5 forecast, 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_oct1, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r model1-holdout, fig.cap="Holdout accuracy, 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_oct1, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r model1-mapes}
res_oct1$mapes(n.ahead = n.forecasts, Y = Y2425)
```

# Rolling forecast origins through a surge (2024-25)

We can then refit the model throughout the surge (1 Oct, 15 Oct, 30 Oct, 15 
Nov, 30 Nov, 15 Dec, 31 Dec, 15 Jan, 31 Jan although it is missing a reading, 
15 Feb) and observe 14-day-ahead accuracy. We find that for origins around 
the peak, accuracy tends to be worse, likely because of the sharp drop 
in growth rate. Coverage is harder to interpret because of the small holdout 
period.

```{r rolling-origins-2425}
origin_dates_2425 <- as.Date(c(
  "2024-10-01", "2024-10-15", "2024-10-30",
  "2024-11-15", "2024-11-30",
  "2024-12-15", "2024-12-31",
  "2025-01-15", "2025-01-31",
  "2025-02-15"
))

fit_rolling_origin <- function(surge_idx_obj, origin_date, q = q.default,
                               n.ahead = 14, sea.period = 7) {
  cal <- surge_idx_obj$calendar
  Y   <- surge_idx_obj$series
  origin_pos <- idx_to_pos(cal, origin_date)
  rng <- idx_range(Y)
  if (origin_pos + n.ahead > rng[2]) {
    return(NULL)  # not enough holdout left to evaluate
  }
  if (is.na(idx_values(Y[origin_pos]))) {
    return(NULL)  # origin falls on a missing PM2.5 day; predict_level()
                  # requires a non-NA cumulative value at the fit's end
  }
  model <- tsgc::SSModelDynamicGompertz(
    Y = Y, q = q, sea.period = sea.period,
    start = rng[1], end = origin_pos, calendar = cal
  )
  res <- tryCatch(tsgc::estimate(model), error = function(e) NULL)
  if (is.null(res)) return(NULL)
  acc <- tryCatch(res$mapes(n.ahead = n.ahead, Y = Y), error = function(e) NULL)
  list(origin = origin_date, res = res, accuracy = acc)
}

rolling_2425 <- lapply(origin_dates_2425, function(d) {
  fit_rolling_origin(s2425, d)
})
names(rolling_2425) <- format(origin_dates_2425)

# Drop any origins too close to the end of the window to evaluate a full
# 14-day holdout.
rolling_2425 <- Filter(Negate(is.null), rolling_2425)
length(rolling_2425)
```

```{r rolling-2425-table}
rolling_accuracy_2425 <- do.call(rbind, lapply(names(rolling_2425), function(nm) {
  a <- rolling_2425[[nm]]$accuracy
  data.frame(origin = nm, mape = a$mape, smape = a$smape,
            mae = a$mae, rmse = a$rmse, coverage = a$coverage)
}))
rolling_accuracy_2425
```

```{r rolling-2425-plot, fig.cap="Forecast accuracy (sMAPE, coverage) by forecast origin, 2024-25 surge."}
rolling_long <- tidyr::pivot_longer(
  rolling_accuracy_2425, cols = c(smape, coverage),
  names_to = "metric", values_to = "value"
)
rolling_long$origin <- factor(rolling_long$origin, levels = rolling_accuracy_2425$origin)

ggplot(rolling_long, aes(x = origin, y = value, group = metric, color = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(title = "14-day-ahead forecast accuracy across the 2024-25 surge",
       x = "Forecast origin", y = NULL, color = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))
```

# Applying the same rolling-origin design to all three surges

We can also apply this rolling-origin analysis to the other surges, though 
here we only look at sMAPE, and compare how the model's accuracy evolves in 
different surges.

```{r rolling-all-surges}
rolling_offsets <- as.numeric(origin_dates_2425 - as.Date("2024-10-01"))

origin_dates_for <- function(anchor_oct1) anchor_oct1 + rolling_offsets

all_surge_origins <- list(
  "2023-24" = origin_dates_for(as.Date("2023-10-01")),
  "2024-25" = origin_dates_for(as.Date("2024-10-01")),
  "2025-26" = origin_dates_for(as.Date("2025-10-01"))
)

rolling_all <- lapply(names(surge_idx), function(surge_nm) {
  origins <- all_surge_origins[[surge_nm]]
  res_list <- lapply(origins, function(d) fit_rolling_origin(surge_idx[[surge_nm]], d))
  names(res_list) <- format(origins)
  Filter(Negate(is.null), res_list)
})
names(rolling_all) <- names(surge_idx)

sapply(rolling_all, length)
```

```{r rolling-all-table}
rolling_accuracy_all <- do.call(rbind, lapply(names(rolling_all), function(surge_nm) {
  do.call(rbind, lapply(names(rolling_all[[surge_nm]]), function(nm) {
    a <- rolling_all[[surge_nm]][[nm]]$accuracy
    data.frame(surge = surge_nm, origin = nm, mape = a$mape, smape = a$smape,
              mae = a$mae, rmse = a$rmse, coverage = a$coverage)
  }))
}))
rolling_accuracy_all
```

```{r rolling-all-plot, fig.cap="14-day-ahead sMAPE by forecast origin, all three surges."}
rolling_accuracy_all$origin_offset <- rep(rolling_offsets, length.out = nrow(rolling_accuracy_all))

ggplot(rolling_accuracy_all, aes(x = origin_offset, y = smape, color = surge)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "14-day-ahead sMAPE across forecast origins, by surge",
       subtitle = "x-axis: days after 1 October of that surge's year",
       x = "Days after 1 October", y = "sMAPE", color = "Surge") +
  theme(plot.title = element_text(face = "bold"))
```

# Weather regressors (`xpred`)

Because pollution is significantly affected by weather, we can add relevant 
weather covariates (temperature, relative humidity, wind speed, and boundary-layer 
height/BLH) to the model. We consider these in the 2024-25 model.

```{r xpred-setup}
weather_cols <- c("at_c_mean", "rh_percent_mean", "ws_m_s_mean",
                  "boundary_layer_height_m_mean")

w2425 <- surge_data[["2024-25"]]
weather_xts <- xts::xts(w2425[, weather_cols], order.by = w2425$date)
weather_idx_full <- xts_to_idx(
  weather_xts,
  start.pos = idx_to_pos(cal2425, w2425$date[1])
)$series

weather_est <- get_timeframe(weather_idx_full, 1, oct1_pos)
head(weather_est)
```

```{r xpred-fit}
model_weather <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  xpred    = weather_est,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_weather <- tsgc::estimate(model_weather)
summary(res_weather)
tsgc::print_model_diagnostics(res_weather)
```

## Realised weather vs a naive forecast

Whilst using realised weather is a fine initial exercise, it would be more 
relevant to use weather forecasts to predict the model. We consider two such 
forecasts from the 1 Oct 2024 origin:

* **Realised weather** -- actual observed weather over the 14-day
  horizon, an upper bound on what perfect foresight buys.
* **Naive persistence** -- each weather variable held at its last
  observed value for the full horizon; uses only information available
  at the origin. While imperfect, this is a reasonable proxy for a forecast 
  when looking at a 14-day-ahead estimation.

```{r xpred-future}
weather_future_realised <- get_timeframe(weather_idx_full, oct1_pos + 1, oct1_pos + n.forecasts)

# Naive persistence: repeat the last observed (1 Oct) weather values for
# every day of the forecast horizon, using only information available at
# the forecast origin.
last_obs_weather <- idx_values(get_timeframe(weather_idx_full, oct1_pos, oct1_pos))
persistence_mat <- matrix(
  rep(last_obs_weather, each = n.forecasts),
  nrow = n.forecasts,
  dimnames = list(NULL, weather_cols)
)
weather_future_naive <- idx_series(persistence_mat, start = oct1_pos + 1)

res_weather_realised <- tsgc::estimate(model_weather)
res_weather_realised$xpred.new <- weather_future_realised

res_weather_naive <- tsgc::estimate(model_weather)
res_weather_naive$xpred.new <- weather_future_naive
```

```{r xpred-forecast-plot, fig.cap="PM2.5 forecast, 1 Oct 2024 origin: realised vs naive-persistence future weather."}
p <- tsgc::plot_forecast(
  res_weather_realised, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\nwith realised future weather (upper-bound comparison)",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_forecast(
  res_weather_naive, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment\nwith naive-persistence future weather (operationally realistic)",
  series.name = "PM2.5"
)
print(p)
```

```{r xpred-holdout, fig.cap="Holdout accuracy: realised vs naive-persistence future weather."}
p <- tsgc::plot_holdout(
  res_weather_realised, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: PM2.5 forecast, realised future weather",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_holdout(
  res_weather_naive, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: PM2.5 forecast, naive-persistence future weather",
  series.name = "PM2.5"
)
print(p)

weather_accuracy_compare <- rbind(
  data.frame(weather_path = "realised", as.data.frame(res_weather_realised$mapes(n.ahead = n.forecasts, Y = Y2425))),
  data.frame(weather_path = "naive_persistence", as.data.frame(res_weather_naive$mapes(n.ahead = n.forecasts, Y = Y2425)))
)
weather_accuracy_compare
```

```{r xpred-compare-paths, fig.cap="Weather-augmented forecast: realised vs naive-persistence future weather, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(realised_weather = res_weather_realised, naive_persistence_weather = res_weather_naive),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Realised vs naive-persistence future weather (2024-25, 1 Oct origin)"
)
```

## Comparison: baseline vs weather-augmented

We can then compare the two forecasts, and note that they behave similarly, 
with more variation in the realised-weather forecast likely due to the use 
of variable weather covariates. 

```{r xpred-compare, fig.cap="Baseline vs weather-augmented forecast, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_oct1, weather_realised = res_weather_realised,
      weather_naive_persistence = res_weather_naive),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Baseline vs weather-augmented forecast (2024-25, 1 Oct origin)"
)
```

# Leading-indicator model: boundary-layer height and PM2.5

We may consider boundary-layer height (BLH) as a mechanistic lead for pollutant
accumulation: a shallow boundary layer traps pollutants near the
surface, but noting this in readings would likely take time from the BLH drop. `SSModelLeadingIndicator` requires both series in `Y` to be strictly increasing 
cumulative series on a shared position axis, so BLH itself (not cumulative, 
not monotone) can't be used directly -- it is first transformed to a positive, 
cumulable series by inverting it (low BLH -> high value) and cumulating that 
within the 2024-25 surge window, exactly as PM2.5 itself was cumulated above.

`SSModelLeadingIndicator$estimate()` additionally requires the lead and
target series, after lagging, to leave a single *contiguous* block of
finite rows -- unlike `SSModelDynamicGompertz`, it does not fall back to
KFAS's native handling of interior missing observations. The isolated
single-day gaps identified earlier are therefore linearly interpolated
here, but only for the purposes of constructing the lead/target pair
`Y_li` used in this section; `Y2425` (used by the baseline Gompertz and
weather models elsewhere in this vignette) is left with its NAs intact.

```{r li-blh-transform}
blh_raw  <- zoo::na.approx(w2425$boundary_layer_height_m_mean, na.rm = FALSE)
pm25_raw <- zoo::na.approx(w2425$pm25_mean, na.rm = FALSE)

inv_blh <- 1 / blh_raw
cum_inv_blh <- cumsum_na(inv_blh)
cum_pm25_li <- cumsum_na(pm25_raw)

li_df <- data.frame(
  date = w2425$date,
  lead = cum_inv_blh,
  targ = cum_pm25_li
)

li_xts <- xts::xts(li_df[, c("lead", "targ")], order.by = li_df$date)
li_conv <- xts_to_idx(li_xts)
Y_li <- li_conv$series
cal_li <- li_conv$calendar
```

## Preview: comovement of inverse-BLH and PM2.5

To compare the series visually, we use non-native plotting as `tsgc`'s own 
plotting functionality does not allow the use of multiple y-axes. 

```{r li-preview-plot, fig.cap="Cumulative inverse boundary-layer height (lead) vs cumulative PM2.5 (target), 2024-25 surge."}
n.lag.li <- 5

li_plot_df <- data.frame(
  date        = li_df$date,
  lead_series = li_df$lead,
  targ_series = li_df$targ
)

lead_range <- range(li_plot_df$lead_series, na.rm = TRUE)
targ_range <- range(li_plot_df$targ_series, na.rm = TRUE)
scale_factor <- diff(targ_range) / diff(lead_range)
offset <- targ_range[1] - lead_range[1] * scale_factor

p <- ggplot(li_plot_df, aes(x = date)) +
  geom_line(aes(y = lead_series * scale_factor + offset, color = "Cum. inverse BLH"),
            linewidth = 0.7, na.rm = TRUE) +
  geom_line(aes(y = targ_series, color = "Cum. PM2.5"),
            linewidth = 0.7, na.rm = TRUE) +
  scale_y_continuous(
    name = "Cumulative PM2.5",
    sec.axis = sec_axis(~ (. - offset) / scale_factor, name = "Cumulative inverse BLH")
  ) +
  scale_color_manual(values = c("Cum. inverse BLH" = "steelblue", "Cum. PM2.5" = "firebrick")) +
  labs(title = "Cumulative inverse BLH (lead) vs cumulative PM2.5 (target)",
       x = NULL, color = NULL) +
  theme(plot.title = element_text(face = "bold"),
        axis.title.y.left = element_text(color = "firebrick"),
        axis.title.y.right = element_text(color = "steelblue"))
print(p)
```

## Estimation from the 1 October forecast origin

```{r li-fit}
model_li <- SSModelLeadingIndicator$new(
  Y          = Y_li,
  n.lag      = n.lag.li,
  LeadIndCol = 1,
  sea.period = 7,
  start      = 1,
  end        = oct1_pos,
  calendar   = cal_li
)
res_li <- estimate(model_li)
summary(res_li)
tsgc::print_model_diagnostics(res_li)
```

Left free, `q` is estimated at 0.3144 here. Thsi lets the log-growth-rate trend 
wander freely between observations, and a random-walk state's forecast-interval
width grows explosively with horizon. This is seen in the forecast and 
holdout plots below.

```{r li-forecast-log, fig.cap="Log growth rate forecast, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_li, Y = Y_li, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5\nleading-indicator forecast from 1 Oct 2024 (q estimated freely)"
)
print(p)
```

```{r li-forecast-level, fig.cap="Daily PM2.5 forecast, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment (leading-indicator model, q estimated freely)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-holdout, fig.cap="Holdout accuracy, leading-indicator model (q estimated freely), 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_li, Y = Y_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast (leading-indicator model, q estimated freely)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-mapes}
res_li$mapes(n.ahead = n.forecasts, Y = Y_li)
```

## Fixing q at the package default

The freely-estimated `q` above is itself informative: with only
~150 days of estimation data, this particular leading-indicator
specification weakly identifies its own state variance, and left
unconstrained it produces forecasts that are technically correct but
practically useless. Fixing `q = q.default`, as
already done for the vanilla Gompertz model throughout this vignette,
keeps the leading-indicator forecast on a comparable, usable scale.

```{r li-fit-qfixed}
model_li_qfixed <- SSModelLeadingIndicator$new(
  Y          = Y_li,
  n.lag      = n.lag.li,
  LeadIndCol = 1,
  sea.period = 7,
  q          = q.default,
  start      = 1,
  end        = oct1_pos,
  calendar   = cal_li
)
res_li_qfixed <- estimate(model_li_qfixed)
summary(res_li_qfixed)
```

```{r li-forecast-log-qfixed, fig.cap="Log growth rate forecast, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_log_forecast(
  res_li_qfixed, Y = Y_li, n.ahead = n.forecasts,
  plt.start = oct1_pos - 30,
  title = "Log growth rate of cumulative PM2.5\nleading-indicator forecast from 1 Oct 2024 (q fixed at default)"
)
print(p)
```

```{r li-forecast-level-qfixed, fig.cap="Daily PM2.5 forecast, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_forecast(
  res_li_qfixed, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_pos - 30,
  title = "Forecast of daily PM2.5 increment (leading-indicator model, q fixed at default)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-holdout-qfixed, fig.cap="Holdout accuracy, leading-indicator model (q fixed at default), 14 days from 1 Oct 2024."}
p <- tsgc::plot_holdout(
  res_li_qfixed, Y = Y_li, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: daily PM2.5 forecast (leading-indicator model, q fixed at default)\n14 days from 1 Oct 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r li-mapes-qfixed}
res_li_qfixed$mapes(n.ahead = n.forecasts, Y = Y_li)
```

The remainder of this section (and the lag/q sensitivity analysis) uses 
`res_li_qfixed`, so its plots stay on a scale comparable to the vanilla 
Gompertz baseline.

## Comparison: baseline vs leading-indicator forecast

We now compare the baseline and leading-indicator models. 

```{r li-compare, fig.cap="Baseline dynamic Gompertz vs BLH leading-indicator forecast (q fixed at default), 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_oct1, leading_indicator_blh = res_li_qfixed),
  n.ahead = n.forecasts,
  title = "Baseline vs leading-indicator (BLH) forecast (2024-25, 1 Oct origin)"
)
```

## Lag selection and `q` sensitivity via rolling cross-validation

While the default value for `q` seems to perform well for epidemics, there is 
no reason this should be a universally optimal value. This is also the case 
for the 5-day lag, which was chosen by inspection. To illustrate a 
potential selection mechanism for `q` and the lag, we jointly sweep these 
so as to select the most accurate compbination.

```{r li-cv-setup}
n.ahead.cv.li  <- 5
gap.cv.li      <- n.ahead.cv.li
n.select.cv.li <- 2
n.report.cv.li <- 1

report.end.cv.li <- oct1_pos - (n.report.cv.li - 1) * gap.cv.li - n.ahead.cv.li
select.end.cv.li <- report.end.cv.li - n.select.cv.li * gap.cv.li - n.ahead.cv.li
stopifnot(select.end.cv.li > 1)

max_lag_cv_li <- 10
stopifnot(select.end.cv.li - 1 > max_lag_cv_li + 1)
```

### Building the lag × q grid

```{r li-cv-models}
q_candidates <- c(0.001, 0.005, 0.01, 0.02, 0.05)
lag_candidates <- c(3, 5, 7, 10)

cv_models_li <- list()

cv_models_li[["Vanilla_q_default"]] <- SSModelDynamicGompertz$new(
  Y = Y2425, q = q.default, sea.period = 7,
  start = 1, end = select.end.cv.li, calendar = cal2425
)

for (qv in q_candidates) {
  cv_models_li[[paste0("Vanilla_q", qv)]] <- SSModelDynamicGompertz$new(
    Y = Y2425, q = qv, sea.period = 7,
    start = 1, end = select.end.cv.li, calendar = cal2425
  )
}

grid_spec_li <- expand.grid(lag = lag_candidates, q = q_candidates)

for (i in seq_len(nrow(grid_spec_li))) {
  lg <- grid_spec_li$lag[i]
  qv <- grid_spec_li$q[i]
  nm <- paste0("LI_lag", lg, "_q", qv)
  cv_models_li[[nm]] <- SSModelLeadingIndicator$new(
    Y = Y_li, n.lag = lg, LeadIndCol = 1, sea.period = 7,
    q = qv, start = 1, end = select.end.cv.li, calendar = cal_li
  )
}
```

### Selection-fold results: full grid

```{r li-cv-select}
cv_selection_li <- cross_val(
  Y = Y_li,
  model_list = cv_models_li,
  est.end = select.end.cv.li,
  criterion = "smape",
  n.ahead = n.ahead.cv.li,
  n.estimate = n.select.cv.li,
  gap = gap.cv.li
)
cv_selection_li$mean_smape <- rowMeans(cv_selection_li[, -1, drop = FALSE])
cv_selection_li[order(cv_selection_li$mean_smape), c("Model", "mean_smape")]
```

### Lag × q grid, reshaped

```{r li-cv-grid-table}
li_only <- cv_selection_li[grepl("^LI_lag", cv_selection_li$Model), c("Model", "mean_smape")]
li_only$lag <- as.integer(sub("^LI_lag(\\d+)_q.*$", "\\1", li_only$Model))
li_only$q   <- as.numeric(sub("^LI_lag\\d+_q", "", li_only$Model))

grid_table_li <- reshape(
  li_only[, c("lag", "q", "mean_smape")],
  idvar = "lag", timevar = "q", direction = "wide"
)
names(grid_table_li) <- sub("mean_smape\\.", "q=", names(grid_table_li))
grid_table_li <- grid_table_li[order(grid_table_li$lag), ]
grid_table_li
```

### Isolating the effect of lag (at the package default `q`)

```{r li-cv-lag-effect}
lag_effect_li <- li_only[li_only$q == q.default, c("lag", "mean_smape")]
lag_effect_li <- lag_effect_li[order(lag_effect_li$lag), ]
lag_effect_li
```

```{r li-lag-effect-plot, fig.cap="Selection-fold SMAPE by lag, leading-indicator model at q = 0.005."}
p <- ggplot(lag_effect_li, aes(x = lag, y = mean_smape)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Selection-fold SMAPE by lag (q = 0.005)",
       x = "Lag (days)", y = "Mean SMAPE") +
  theme(plot.title = element_text(face = "bold"))
print(p)
```

### Isolating the effect of `q` (at the best lag found above)

```{r li-cv-q-effect}
best_lag_li <- lag_effect_li$lag[which.min(lag_effect_li$mean_smape)]
best_lag_li

q_effect_li <- li_only[li_only$lag == best_lag_li, c("q", "mean_smape")]
q_effect_li <- q_effect_li[order(q_effect_li$q), ]
q_effect_li

vanilla_rows_li <- cv_selection_li[grepl("^Vanilla_q", cv_selection_li$Model), c("Model", "mean_smape")]
vanilla_rows_li$q <- ifelse(
  vanilla_rows_li$Model == "Vanilla_q_default", q.default,
  as.numeric(sub("^Vanilla_q", "", vanilla_rows_li$Model))
)
vanilla_rows_li <- vanilla_rows_li[order(vanilla_rows_li$q), c("q", "mean_smape")]
vanilla_rows_li
```

```{r li-q-effect-plot, fig.cap="Selection-fold SMAPE across q, leading-indicator model vs vanilla Gompertz."}
q_effect_li$model <- paste0("LI_lag", best_lag_li)
vanilla_rows_li$model <- "Vanilla"
q_compare_li <- rbind(q_effect_li, vanilla_rows_li)

p <- ggplot(q_compare_li, aes(x = q, y = mean_smape, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(title = paste0("Selection-fold SMAPE across q (LI at lag = ", best_lag_li, ")"),
       x = "q (log scale)", y = "Mean SMAPE", color = NULL) +
  theme(plot.title = element_text(face = "bold"))
print(p)
```

### Reporting-fold accuracy of the selected specification

Having picked `best_lag_li` and the `q` that minimised selection-fold
SMAPE from the tables above, the reporting fold gives a final,
held-out accuracy check against the vanilla Gompertz benchmark.

```{r li-cv-report}
best_q_li <- q_effect_li$q[which.min(q_effect_li$mean_smape)]
selected_spec_li <- paste0("LI_lag", best_lag_li, "_q", best_q_li)
selected_spec_li

cv_report_models_li <- cv_models_li[c("Vanilla_q_default", selected_spec_li)]

cv_reporting_li <- cross_val(
  Y = Y_li,
  model_list = cv_report_models_li,
  est.end = report.end.cv.li,
  criterion = "smape",
  n.ahead = n.ahead.cv.li,
  n.estimate = n.report.cv.li,
  gap = gap.cv.li
)
cv_reporting_li
```

## Forecast overlay across `q`

```{r li-q-overlay-fit}
res_li_q_list <- lapply(q_candidates, function(qv) {
  m <- SSModelLeadingIndicator$new(
    Y = Y_li, n.lag = best_lag_li, LeadIndCol = 1, sea.period = 7,
    q = qv, start = 1, end = oct1_pos, calendar = cal_li
  )
  estimate(m)
})
names(res_li_q_list) <- paste0("q_", q_candidates)
```

```{r li-q-overlay-plot, fig.cap="Leading-indicator forecast of daily PM2.5 across candidate q values, 1 Oct 2024 origin."}
tsgc::plot_compare_forecast(
  res_li_q_list,
  n.ahead = n.forecasts,
  title = paste0("Leading-indicator (BLH) forecast across q, 2024-25, 1 Oct origin, lag = ", best_lag_li)
)
```

# Reinitialisation for a single new surge

We can also reinitialise the model on 1 October as the series transitions 
into the 2024-25 surge, and fit on a longer window including the tail from 
the 2023-2024 surge..

```{r reinit-single-setup}
span_start <- as.Date("2024-07-01")
span_end   <- surge_windows$end[surge_windows$label == "2024-25"]
w_span <- pm[pm$date >= span_start & pm$date <= span_end, ]
w_span$cum_pm25 <- cumsum_na(w_span$pm25_mean)

span_xts <- xts::xts(w_span$cum_pm25, order.by = w_span$date)
span_conv <- xts_to_idx(span_xts)
Y_span <- span_conv$series
cal_span <- span_conv$calendar

reinit_pos <- idx_to_pos(cal_span, as.Date("2024-10-01"))
span_end_pos <- idx_range(Y_span)[2]
long_end     <- span_end_pos - n.forecasts  # leave a holdout for plot_compare_forecast
```

```{r reinit-single-fit}
model_span_base <- tsgc::SSModelDynamicGompertz(
  Y = Y_span, q = q.default, start = 1, end = long_end, calendar = cal_span
)
res_span_base <- tsgc::estimate(model_span_base)

model_span_reinit <- tsgc::SSModelDynamicGompertz(
  Y = Y_span, q = q.default, start = 1, end = long_end,
  reinit.idx = reinit_pos, calendar = cal_span
)
res_span_reinit <- tsgc::estimate(model_span_reinit)
summary(res_span_reinit)
tsgc::print_model_diagnostics(res_span_reinit)
```

```{r reinit-single-compare, fig.cap="Reinitialised vs non-reinitialised forecast at the 2024-25 surge onset."}
tsgc::plot_compare_forecast(
  list(no_reinit = res_span_base, reinit = res_span_reinit),
  n.ahead = n.forecasts, actual = Y_span,
  title = "Effect of reinitialising at 1 Oct 2024 (surge onset)"
)
```

# Continual (chained) reinitialisation across all observed surges

Another potential exercise is to chain reinitialisations at the start of every 
surge, rather than fitting three individual models.`reinit.idx` only takes one 
restart position per model, so this is a
manual chain: 1 October of each subsequent year passed as `reinit.idx`
in turn, re-basing the growth-rate target to zero each time, while
`start` stays at the series start throughout -- `use.presample.info =
TRUE` internally re-estimates a presample fit over `[start, reinit.idx]`
to seed the post-reinit state. `end` advances through the
boundaries (Oct 2023, Oct 2024, final origin below). What changes
segment to segment is the rebasing point and how far the fit extends,
not the span of history informing it. Note that `original.results` is 
only valid at the exact `reinit.idx` it was estimated for, so it isn't reused 
across segments. Each segment lets `use.presample.info` re-derive its presample 
fit instead.

```{r chained-reinit-setup}
full_xts <- xts::xts(pm$pm25_mean, order.by = pm$date)
full_conv <- xts_to_idx(full_xts)
Y_full <- full_conv$series
cal_full <- full_conv$calendar
Y_full_cum <- idx_series(cumsum_na(idx_values(Y_full)), start = Y_full$start)

reinit_dates <- as.Date(c("2023-10-01", "2024-10-01", "2025-10-01"))
reinit_positions <- sapply(reinit_dates, function(d) idx_to_pos(cal_full, d))
reinit_positions
```

```{r chained-reinit-fit}
chain_reinit_forward <- function(Y, calendar, reinit_positions, q = q.default,
                                 sea.period = 7, end = idx_range(Y)[2]) {
  series_start <- idx_range(Y)[1]
  results <- list()

  boundaries <- c(reinit_positions, end)

  for (i in seq_along(boundaries)) {
    seg_end <- boundaries[i]
    if (i == 1) {
      # First segment: no reinitialisation, plain fit from the series
      model <- tsgc::SSModelDynamicGompertz(
        Y = Y, q = q, sea.period = sea.period,
        start = series_start, end = seg_end, calendar = calendar
      )
      res <- tsgc::estimate(model)
    } else {
      # Reinitialise
      reinit_pos_i <- reinit_positions[i - 1]
      model <- tsgc::SSModelDynamicGompertz(
        Y = Y, q = q, sea.period = sea.period,
        start = series_start, end = seg_end,
        reinit.idx = reinit_pos_i,
        use.presample.info = TRUE,
        calendar = calendar
      )
      res <- tsgc::estimate(model)
    }
    results[[i]] <- res
  }
  results
}

final_segment_end <- idx_to_pos(cal_full, as.Date("2025-10-01"))
reinit_positions_for_chain <- reinit_positions[reinit_positions < final_segment_end]

chained_results <- chain_reinit_forward(
  Y_full_cum, cal_full, reinit_positions_for_chain,
  end = final_segment_end
)
length(chained_results)
```

```{r chained-reinit-summary}
invisible(lapply(seq_along(chained_results), function(i) {
  cat("--- Segment", i, "---\n")
  print(summary(chained_results[[i]]))
}))
```

## Comparing continual reinitialisation to independent per-surge models

We can then compare how this chained model performs against the individual 
surges models.

```{r chained-vs-independent-forecast, fig.cap="14-day forecast from 1 Oct 2025: chained (multi-surge reinitialisation) vs a model fitted only on the 2025-26 surge."}
surge_2025_26_start <- surge_windows$start[surge_windows$label == "2025-26"]
start_pos_2526_full <- idx_to_pos(cal_full, surge_2025_26_start)
oct1_2526_pos_full  <- idx_to_pos(cal_full, as.Date("2025-10-01"))

w2526 <- pm[pm$date >= surge_2025_26_start & pm$date <= as.Date("2025-10-01"), ]
Y_2526_only <- idx_series(cumsum_na(w2526$pm25_mean), start = start_pos_2526_full)

model_2526_indep <- tsgc::SSModelDynamicGompertz(
  Y = Y_2526_only, q = q.default, sea.period = 7,
  start = start_pos_2526_full, end = oct1_2526_pos_full, calendar = cal_full
)
res_2526_indep <- tsgc::estimate(model_2526_indep)

final_chained <- chained_results[[length(chained_results)]]

tsgc::plot_compare_forecast(
  list(chained = final_chained, independent = res_2526_indep),
  n.ahead = n.forecasts, actual = Y_full_cum,
  title = "Chained (continually reinitialised) vs independently fitted 2025-26 model"
)
```

# Extensions: a Diwali festival regressor

Firecracker use around Diwali falls within the October-November window
where forecast accuracy has been weakest above. We can construct a Diwali 
dummy and add it as a covariate. We use a 5 day span to capture effects 
from the days surrounding Diwali rather than just on-the-day effects.

## Constructing the Diwali dummy

```{r diwali-dates}
diwali_dates <- as.Date(c("2023-11-12", "2024-11-01", "2025-10-20", "2026-11-08"))
names(diwali_dates) <- c("2023", "2024", "2025", "2026")
diwali_dates
```

```{r diwali-dummy}
diwali_window <- 2L  # days either side of Diwali itself

pm$diwali_dummy <- 0L
for (d in diwali_dates) {
  in_window <- pm$date >= (d - diwali_window) & pm$date <= (d + diwali_window)
  pm$diwali_dummy[in_window] <- 1L
}

table(pm$diwali_dummy)
```

```{r diwali-plot, fig.cap="Diwali windows against daily PM2.5, 2023-2026."}
ggplot(pm, aes(x = date, y = pm25_mean)) +
  geom_line(color = "grey40", linewidth = 0.3, na.rm = TRUE) +
  geom_point(data = pm[pm$diwali_dummy == 1L, ], color = "darkorange", size = 1.2, na.rm = TRUE) +
  labs(title = "Diwali windows (orange) against daily PM2.5",
       x = NULL, y = expression(paste("PM2.5 (", mu, "g/", m^3, ")"))) +
  theme(plot.title = element_text(face = "bold"))
```

## Estimating with the Diwali regressor

```{r diwali-xpred-setup}
w2425 <- surge_data[["2024-25"]]
w2425$diwali_dummy <- pm$diwali_dummy[match(w2425$date, pm$date)]

diwali_weather_cols <- c(weather_cols, "diwali_dummy")
diwali_xreg_xts <- xts::xts(w2425[, diwali_weather_cols], order.by = w2425$date)
diwali_xreg_idx_full <- xts_to_idx(
  diwali_xreg_xts,
  start.pos = idx_to_pos(cal2425, w2425$date[1])
)$series

diwali_xreg_est <- get_timeframe(diwali_xreg_idx_full, 1, oct1_pos)
```

```{r diwali-xpred-fit}
model_diwali <- tsgc::SSModelDynamicGompertz(
  Y        = Y2425,
  xpred    = diwali_xreg_est,
  q        = q.default,
  sea.period = 7,
  start    = 1,
  end      = oct1_pos,
  calendar = cal2425
)
res_diwali <- tsgc::estimate(model_diwali)
summary(res_diwali)
tsgc::print_model_diagnostics(res_diwali)
```

Diwali fell on 1 November in 2024. Estimating from a 1-14 October origin
leaves the dummy at zero throughout the estimation window, since Diwali
falls entirely in the forecast horizon rather than the estimation
sample; the dummy's coefficient is then not identified by the data at
all and is carried forward purely from KFAS's diffuse initialisation.
This is visible above as `xregdiwali_dummy` estimating at exactly 0
with a 0 standard error in `res_diwali`, and produces the "model is
degenerate" warnings in `KFS()`.

To let the model actually learn the Diwali effect from data, the
estimation window is moved to 10 November -- a few days after the
1 November +/-2-day Diwali window (which runs 30 Oct-3 Nov) closes, so
those non-zero dummy rows sit inside the estimation sample rather than
the forecast horizon. The 14-day forecast horizon from this origin
(11-24 Nov) then tests the post-Diwali decay rather than asking the
model to predict the festival spike itself from an unidentified
coefficient.

```{r diwali-forecast-origin}
nov10_pos <- idx_to_pos(cal2425, as.Date("2024-11-10"))
stopifnot(!is.na(idx_values(Y2425[nov10_pos])))

model_baseline_nov10 <- tsgc::SSModelDynamicGompertz(
  Y = Y2425, q = q.default, sea.period = 7,
  start = 1, end = nov10_pos, calendar = cal2425
)
res_baseline_nov10 <- tsgc::estimate(model_baseline_nov10)

diwali_xreg_est_nov10 <- get_timeframe(diwali_xreg_idx_full, 1, nov10_pos)
table(diwali_xreg_est_nov10$data[, "diwali_dummy"])

model_diwali_nov10 <- tsgc::SSModelDynamicGompertz(
  Y = Y2425, xpred = diwali_xreg_est_nov10, q = q.default, sea.period = 7,
  start = 1, end = nov10_pos, calendar = cal2425
)
res_diwali_nov10 <- tsgc::estimate(model_diwali_nov10)
summary(res_diwali_nov10)

diwali_xreg_future <- get_timeframe(diwali_xreg_idx_full, nov10_pos + 1, nov10_pos + n.forecasts)
res_diwali_nov10$xpred.new <- diwali_xreg_future
```

```{r diwali-forecast-plot, fig.cap="14-day forecast from 10 Nov 2024 (post-Diwali decay): baseline vs weather+Diwali."}
p <- tsgc::plot_forecast(
  res_diwali_nov10, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = nov10_pos - 30,
  title = "Forecast of daily PM2.5 increment\nweather + Diwali dummy, from 10 Nov 2024",
  series.name = "PM2.5"
)
print(p)
```

```{r diwali-holdout-compare, fig.cap="Holdout accuracy, 10 Nov 2024 origin: baseline vs weather+Diwali."}
p <- tsgc::plot_holdout(
  res_baseline_nov10, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: baseline, 10 Nov 2024 origin (Diwali in estimation window)",
  series.name = "PM2.5"
)
print(p)

p <- tsgc::plot_holdout(
  res_diwali_nov10, Y = Y2425, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: weather + Diwali dummy, 10 Nov 2024 origin",
  series.name = "PM2.5"
)
print(p)

diwali_accuracy_compare <- rbind(
  data.frame(model = "baseline", as.data.frame(res_baseline_nov10$mapes(n.ahead = n.forecasts, Y = Y2425))),
  data.frame(model = "weather_plus_diwali", as.data.frame(res_diwali_nov10$mapes(n.ahead = n.forecasts, Y = Y2425)))
)
diwali_accuracy_compare
```

```{r diwali-compare-plot, fig.cap="Baseline vs weather+Diwali forecast, 10 Nov 2024 origin."}
tsgc::plot_compare_forecast(
  list(baseline = res_baseline_nov10, weather_plus_diwali = res_diwali_nov10),
  n.ahead = n.forecasts, actual = Y2425,
  title = "Baseline vs weather + Diwali dummy (2024-25, 10 Nov origin, post-Diwali)"
)
```

# Extension: cumulative pollution over the whole record

The final extension is to consider the entire dataset as a whole rather than 
as three individual surges. Indeed, COVID cases in the UK never reached a 
proper saturation level, but the model worked quite well.

## Building the whole-record cumulative series

```{r full-record-cumulative}
pm_full <- pm[order(pm$date), ]
pm_full$cum_pm25_full <- cumsum_na(pm_full$pm25_mean)

full_xts <- xts::xts(pm_full$cum_pm25_full, order.by = pm_full$date)
full_conv <- xts_to_idx(full_xts)
Y_full_nosurge <- full_conv$series
cal_full_nosurge <- full_conv$calendar

range(pm_full$date)
tail(pm_full$cum_pm25_full, 1)
```

```{r full-record-plot, fig.cap="Cumulative PM2.5 across the entire observed record, no surge resets."}
ggplot(pm_full, aes(x = date, y = cum_pm25_full)) +
  geom_line(color = "firebrick", linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Cumulative PM2.5, entire record (no surge resets)",
       x = NULL, y = "Cumulative PM2.5") +
  theme(plot.title = element_text(face = "bold"))
```

## Baseline dynamic Gompertz on the full record

A forecast origin is chosen in the most recent complete winter
(1 October 2025, i.e. early in the 2025-26 season) so there is enough
prior history -- almost three full winters -- for the model to have
something to learn from, while still leaving a genuine holdout period.

```{r full-record-baseline-fit}
oct1_2025_pos_full <- idx_to_pos(cal_full_nosurge, as.Date("2025-10-01"))
stopifnot(!is.na(idx_values(Y_full_nosurge[oct1_2025_pos_full])))

model_full_nosurge <- tsgc::SSModelDynamicGompertz(
  Y = Y_full_nosurge, q = q.default, sea.period = 7,
  start = 1, end = oct1_2025_pos_full, calendar = cal_full_nosurge
)
res_full_nosurge <- tsgc::estimate(model_full_nosurge)
summary(res_full_nosurge)
tsgc::print_model_diagnostics(res_full_nosurge)
```

```{r full-record-forecast-plot, fig.cap="Forecast of daily PM2.5, whole-record cumulative model, 14 days from 1 Oct 2025."}
p <- tsgc::plot_forecast(
  res_full_nosurge, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = oct1_2025_pos_full - 60,
  title = "Forecast of daily PM2.5 increment (whole-record cumulative model)\n14 days from 1 Oct 2025",
  series.name = "PM2.5"
)
print(p)
```

```{r full-record-holdout-plot, fig.cap="Holdout accuracy, whole-record cumulative model, 14 days from 1 Oct 2025."}
p <- tsgc::plot_holdout(
  res_full_nosurge, Y = Y_full_nosurge, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: whole-record cumulative model, 14 days from 1 Oct 2025",
  series.name = "PM2.5"
)
print(p)

res_full_nosurge$mapes(n.ahead = n.forecasts, Y = Y_full_nosurge)
```

## Comparison: whole-record model vs the surge-window baseline

The surge-window model (Model 1) is only ever estimated within a single
season, so it can't be compared on the same forecast origin directly.
Instead, we reuse the rolling-origin method, fitting a fresh surge-window 
model from an initial origin on 1 October 2025.

```{r full-record-vs-surge-setup}
w2526_cmp <- surge_data[["2025-26"]]
oct1_2025_pos_surge <- idx_to_pos(surge_idx[["2025-26"]]$calendar, as.Date("2025-10-01"))

model_surge_2526_cmp <- tsgc::SSModelDynamicGompertz(
  Y = surge_idx[["2025-26"]]$series, q = q.default, sea.period = 7,
  start = 1, end = oct1_2025_pos_surge, calendar = surge_idx[["2025-26"]]$calendar
)
res_surge_2526_cmp <- tsgc::estimate(model_surge_2526_cmp)

surge_vs_full_compare <- rbind(
  data.frame(model = "surge_window_2025_26",
             as.data.frame(res_surge_2526_cmp$mapes(n.ahead = n.forecasts, Y = surge_idx[["2025-26"]]$series))),
  data.frame(model = "whole_record_no_surge",
             as.data.frame(res_full_nosurge$mapes(n.ahead = n.forecasts, Y = Y_full_nosurge)))
)
surge_vs_full_compare
```

## Leading-indicator model on the full record: boundary-layer height and PM2.5

We can repeat the whole-dataset exercise with the leading-indicator model, 
using the BLH as the lead series.

```{r full-record-li-transform}
blh_raw_full  <- zoo::na.approx(pm_full$boundary_layer_height_m_mean, na.rm = FALSE)
pm25_raw_full <- zoo::na.approx(pm_full$pm25_mean, na.rm = FALSE)

inv_blh_full <- 1 / blh_raw_full
cum_inv_blh_full <- cumsum_na(inv_blh_full)
cum_pm25_li_full <- cumsum_na(pm25_raw_full)

li_df_full <- data.frame(
  date = pm_full$date,
  lead = cum_inv_blh_full,
  targ = cum_pm25_li_full
)

li_xts_full <- xts::xts(li_df_full[, c("lead", "targ")], order.by = li_df_full$date)
li_conv_full <- xts_to_idx(li_xts_full)
Y_li_full <- li_conv_full$series
cal_li_full <- li_conv_full$calendar
```

```{r full-record-li-preview-plot, fig.cap="Cumulative inverse boundary-layer height (lead) vs cumulative PM2.5 (target), entire record."}
li_plot_df_full <- data.frame(
  date        = li_df_full$date,
  lead_series = li_df_full$lead,
  targ_series = li_df_full$targ
)

lead_range_full <- range(li_plot_df_full$lead_series, na.rm = TRUE)
targ_range_full <- range(li_plot_df_full$targ_series, na.rm = TRUE)
scale_factor_full <- diff(targ_range_full) / diff(lead_range_full)
offset_full <- targ_range_full[1] - lead_range_full[1] * scale_factor_full

p <- ggplot(li_plot_df_full, aes(x = date)) +
  geom_line(aes(y = lead_series * scale_factor_full + offset_full, color = "Cum. inverse BLH"),
            linewidth = 0.6, na.rm = TRUE) +
  geom_line(aes(y = targ_series, color = "Cum. PM2.5"),
            linewidth = 0.6, na.rm = TRUE) +
  scale_y_continuous(
    name = "Cumulative PM2.5",
    sec.axis = sec_axis(~ (. - offset_full) / scale_factor_full, name = "Cumulative inverse BLH")
  ) +
  scale_color_manual(values = c("Cum. inverse BLH" = "steelblue", "Cum. PM2.5" = "firebrick")) +
  labs(title = "Cumulative inverse BLH (lead) vs cumulative PM2.5 (target), entire record",
       x = NULL, color = NULL) +
  theme(plot.title = element_text(face = "bold"),
        axis.title.y.left = element_text(color = "firebrick"),
        axis.title.y.right = element_text(color = "steelblue"))
print(p)
```

```{r full-record-li-fit}
nov10_pos_full <- idx_to_pos(cal_li_full, as.Date("2025-11-10"))
stopifnot(!is.na(idx_values(Y_li_full[nov10_pos_full])))

model_li_full <- SSModelLeadingIndicator$new(
  Y          = Y_li_full,
  n.lag      = n.lag.li,
  LeadIndCol = 1,
  sea.period = 7,
  q          = q.default,
  start      = 1,
  end        = nov10_pos_full,
  calendar   = cal_li_full
)
res_li_full <- estimate(model_li_full)
summary(res_li_full)
tsgc::print_model_diagnostics(res_li_full)
```

```{r full-record-li-forecast-plot, fig.cap="Daily PM2.5 forecast, whole-record leading-indicator model, 14 days from 10 Nov 2025."}
p <- tsgc::plot_forecast(
  res_li_full, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  plt.start = nov10_pos_full - 60,
  title = "Forecast of daily PM2.5 increment (whole-record leading-indicator model)\n14 days from 10 Nov 2025",
  series.name = "PM2.5"
)
print(p)
```

```{r full-record-li-holdout-plot, fig.cap="Holdout accuracy, whole-record leading-indicator model, 14 days from 10 Nov 2025."}
p <- tsgc::plot_holdout(
  res_li_full, Y = Y_li_full, n.ahead = n.forecasts, confidence.level = CONF_LEVEL,
  title = "Accuracy: whole-record leading-indicator model, 14 days from 10 Nov 2025",
  series.name = "PM2.5"
)
print(p)

res_li_full$mapes(n.ahead = n.forecasts, Y = Y_li_full)
```

## Comparison: whole-record baseline vs whole-record leading-indicator

Finally we can compare the two models on the whole dataset.

```{r full-record-li-compare, fig.cap="Whole-record baseline vs BLH leading-indicator forecast, 10 Nov 2025 origin."}
model_baseline_full_nov10 <- tsgc::SSModelDynamicGompertz(
  Y = Y_full_nosurge, q = q.default, sea.period = 7,
  start = 1, end = nov10_pos_full, calendar = cal_full_nosurge
)
res_baseline_full_nov10 <- tsgc::estimate(model_baseline_full_nov10)

tsgc::plot_compare_forecast(
  list(baseline = res_baseline_full_nov10, leading_indicator_blh = res_li_full),
  n.ahead = n.forecasts,
  title = "Whole-record baseline vs leading-indicator (BLH) forecast, 10 Nov 2025 origin"
)
```
