---
title: "Forecasting epidemic trajectories: Time Series Growth Curves package `tsgc`"
author: "Michael Ashby, Andrew Harvey, Paul Kattuman, Edwin Tang, Craig Thamotheram"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: book.bib
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Forecasting epidemic trajectories: Time Series Growth Curves package `tsgc`}
  %\VignetteEngine{knitr::rmarkdown}
abstract: "This paper documents the Time Series Growth Curves (`tsgc`) package for `R`, which is designed for forecasting epidemic trajectories, including the detection of new waves and turning points. The package implements dynamic Gompertz state-space methods, together with a leading-indicator extension that uses a related series moving ahead of the variable of interest to improve short-horizon forecasts, with lag selection supported by rolling-origin cross-validation. The state-space framework applies equally to daily, weekly, monthly, quarterly, or annual data. Estimation and forecasting are defined purely in terms of integer-position (`idx_series`) data; calendar time is reintroduced only as an optional, purely cosmetic layer for plotting. The package is illustrated with COVID-19 confirmed cases, hospitalisation data, a cross-country leading-indicator comparison, and non-epidemic applications such as the diffusion of new products and the adoption of online services. The models are estimated with Kalman-filter techniques and are suitable for short-horizon forecasting of variables whose cumulative form is subject to an unknown saturation level."
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
# NOTE: warnings are shown (not globally suppressed), so that "rows
# removed" and similar notices are visible and can be understood and
# handled locally rather than hidden.
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.width = 7,
                      fig.height = 7)

library(tsgc)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(zoo)
library(xts)
theme_set(theme_economist_white(gray_bg = FALSE, base_size = 16))
```

Naming a specific `tsgc` commit in prose is not the same as verifying
that commit against what is actually installed when this vignette is
rendered. The chunk below records the installed `tsgc` version (and,
where available, the exact commit/remote SHA `install_github()`
recorded at install time), so results in this vignette can be tied to
a specific build rather than "whatever `tsgc` happened to be installed
on the day it was rendered." Set `EXPECTED_TSGC_COMMIT` below to pin
against a specific commit; leave it `NULL` to only report what is
installed without checking it against an expectation.

```{r tsgc-version-check}
EXPECTED_TSGC_COMMIT <- NULL  # e.g. "a1b2c3d4..." to pin/verify a specific commit

`%||%` <- function(x, y) if (is.null(x)) y else x
tsgc_desc <- tryCatch(utils::packageDescription("tsgc"), error = function(e) NULL)
tsgc_version <- if (!is.null(tsgc_desc)) tsgc_desc$Version else NA_character_
tsgc_commit <- if (!is.null(tsgc_desc)) {
  tsgc_desc$GithubSHA1 %||% tsgc_desc$RemoteSha %||% NA_character_
} else NA_character_

tsgc_version
tsgc_commit

if (!is.null(EXPECTED_TSGC_COMMIT) &&
    (is.na(tsgc_commit) || !identical(tsgc_commit, EXPECTED_TSGC_COMMIT))) {
  warning("Installed tsgc commit does not match EXPECTED_TSGC_COMMIT; ",
          "results below may not match the pinned replication exactly.")
}
```

For a fully pinned re-render, generate an `renv.lock` alongside this
vignette (`renv::snapshot()`, once `renv` is initialised for the
project) so every dependency version - not just `tsgc` - is fixed, and
capture `sessionInfo()` for the rendering session:

```{r session-info, eval = FALSE}
# Not evaluated by default (eval = FALSE) since this writes to disk;
# set eval = TRUE for a run whose purpose is to produce a pinned,
# archivable record of the rendering environment.
writeLines(capture.output(sessionInfo()), con = "sessionInfo.txt")
if (requireNamespace("renv", quietly = TRUE)) {
  renv::snapshot(prompt = FALSE)
}
```

# Introduction

Outbreaks of infectious diseases with epidemic potential require
real-time responses by public health authorities. Accurate real-time
forecasting of the trajectory of the epidemic over the near future is of
great value in this regard.

The `R` package, `tsgc`, is intended for use in monitoring and
forecasting the progress of an epidemic, including the detection of *new
waves* and *turning points*. It develops and implements time series
growth curve methods first reported in @Harvey2020 (hereinafter referred
to as HK). HK develop a class of time series models for predicting
future values of a variable which, when cumulated, is subject to an
unknown saturation level. In a single wave of an epidemic, as more and
more people get infected, the pool of susceptible individuals dwindles.
This results in the decline of new infections, and the cumulative number
of infections approaches its saturation level. The model can take
account of deviations relative to this canonical trajectory due to
changes in social behavior and policy. Models in this family are
relevant for many other disciplines, such as marketing (when estimating
the demand for new products). While attention here is focused on the
spread of epidemics and the applications used for illustration relate to
coronavirus, this package is designed with a view to wider
applicability.

Given the number of different modeling approaches for epidemics, there
are many notable packages that can be used for monitoring epidemics. For
the most part, these seek to model explicitly the mechanism by which the
disease spreads through the population. For example, `EpiModel`
[@Jenness2018], `EpiEstim` [@cori_etal], `epinowcast`
[@sam_abbott_pietro_monticone_2021] to name a few, can be categorized as
belonging to the class of *mechanistic* models in the language of
philosophy of science, in that they require structural knowledge of the
disease spread mechanism in order to obtain predictions.

In contrast, the empirical approach implemented in `tsgc` falls into the
class of models described as *phenomenological*. Although it is
motivated by the archetypal pattern in the dynamics of disease spread,
it does not rely on structural assumptions derived from epidemiological
theory. There are advantages to not requiring assumptions about values
of parameters relating to, inter alia, disease infectiousness, disease
severity, or contact patterns, which are difficult to pin down with
sufficient precision, especially in real-time during an epidemic. Our
approach makes minimal assumptions and merely requires past observations
of the epidemic variable of interest, to which we apply time-series
methods to provide predictions over short future time horizons. The
model can be estimated quickly and straightforwardly, and subjected to
standard diagnostic tests. A statistical model of this type is a useful
complement to mechanistic models that attempt to describe the epidemic
in terms of underlying processes.

Section 2 sets out the state space formulation of the dynamic Gompertz
growth curve and the way nowcasts and forecasts are obtained from
predictive recursions. It is then shown how these numbers translate into
estimates of the instantaneous reproduction number $R_t$. Section 3
explains how multiple waves can be accommodated by reinitializing the
series at the start of new waves. The start of a new wave is not obvious
in real time but a rule for triggering reinitialization that works well
in practice is presented. Section 4 describes the functionality of
`tsgc`, including the leading-indicator extension. Section 5 sets out a
full working example of the package, beginning with COVID-19 data for
Gauteng province in South Africa and continuing with the
leading-indicator model, a comparison against the dynamic Gompertz
model, lag selection by rolling cross-validation, the forecast accuracy
measures used to evaluate the package's models, and the package's
applicability to data of non-daily frequency and non-epidemological
nature. Section 6 concludes.

# Theory

## Gompertz curve

Our model is based on the sigmoidal growth curve pattern that
characterizes epidemics. We start by assuming that the cumulative number
of cases follows a Gompertz curve, which is a parsimonious model for the
canonical sigmoid shape of cumulative case numbers in a one-wave
epidemic. Over the course of a wave, the number of new infected cases
increases up to a peak before declining to zero as the pool of
susceptible individuals declines. Specifically, if the cumulative number
of cases at time $t$, $\mu(t)$, follows a Gompertz curve, we can write
$$\mu(t) = \bar{\mu}\exp\{\gamma_0 e^{\gamma t}\},$$ where $\bar{\mu}$
is the unknown saturation level for the cumulative number of cases,
$\gamma_0 < 0$ is a parameter related to $\mu(0)$ and $\gamma < 0$ is
the growth rate parameter. Defining $\dot{\mu}(t) \equiv d\mu(t)/dt$ and
$g(t) \equiv {\dot\mu(t)}/{\mu(t)}$, it is straightforward to show that

\begin{align}
\ln g(t) &= \delta + \gamma t, & (1)
\end{align}where $\delta = \ln \gamma_0 \gamma$.

The observational model needs to be specified in discrete, rather than
continuous, time. This is straightforward. Let $Y_t$ be the observed
cumulative number of cases on day $t$ and $y_t = Y_t-Y_{t-1}$ be the
number of daily new cases.[^1] We can then define the growth rate of
$Y_t$ as $g_t = y_t/Y_{t-1}$ and replace $\ln g(t)$ with
$\ln y_{t}-\ln Y_{t-1}$.

[^1]: Of course, while $y_t$ denotes daily new cases here, it could
    equally denote weekly sales of a new product, etc. None of the
    analysis here is dependent on the data frequency.

## Dynamic Gompertz model

The deterministic trend implied by (1) is too inflexible for practical
time-series modeling of an epidemic. Replacing it with a stochastic
trend allows the model to adapt to changes in dynamics during the course
of the epidemic. We call this stochastic-trend counterpart of (1) the
dynamic Gompertz model. It is a local linear trend model specified as

\begin{align}
\ln g_{t}&=\delta_{t}+\varepsilon_{t}, \;\;\;\;\;\varepsilon_{t}\sim
NID(0,\sigma_{\varepsilon }^{2}),\;t=2,...,T, & (2)
\end{align} where $\ln g_{t}=\ln y_{t}-\ln Y_{t-1}$ and

\begin{align}
\delta_{t} &=\delta_{t-1}+\gamma_{t-1}, & (3)\\
\gamma_{t} &=\gamma_{t-1}+\zeta_{t},\;\;\; \zeta_{t}\sim
NID(0,\sigma_{\zeta }^{2}), & (4)
\end{align} where the disturbances $\varepsilon_{t}$ and $\zeta_{t}$
are mutually independent, and $NID(0,\sigma^2)$ denotes normally and
independently distributed with mean zero and variance $\sigma^2$. Note
that the larger the signal-to-noise ratio,
$q_{\zeta }=\sigma_{\zeta }^{2}/\sigma_{\varepsilon }^{2}$, the faster
the estimate of the slope parameter, $\gamma_t$, which can be
interpreted as the growth rate of the growth rate of cumulative cases,
changes in response to new observations. Conversely, a lower
signal-to-noise ratio induces more smoothness to the estimates. When
$\sigma_{\zeta }^{2}=0$, the trend is deterministic as in (1).

## State space form and estimation

It is convenient to write the dynamic Gompertz model in general state
space form: \begin{align*}
\ln g_t &= Z\alpha_t + \varepsilon_t & \varepsilon_{t} \sim NID(0,\sigma^2_{\varepsilon})\\
\alpha_{t+1} &= T\alpha_{t} + R \eta_t & \eta_{t} \sim NID(0,Q)
\end{align*} with \begin{equation*}
\alpha_t = \left(\delta_t, \gamma_t\right)', \; Z = \left(1, 0\right), \; \eta_t = \left(0, \zeta_t\right)', \;
T = \begin{pmatrix}
1 & 1 \\
0 & 1
\end{pmatrix}, \;
R = \begin{pmatrix}
0 & 0 \\
0 & 1
\end{pmatrix}, \;
Q =
\begin{pmatrix}
0 & 0 \\
0 & \sigma^2_{\zeta}
\end{pmatrix}.
\end{equation*} This model can be estimated using techniques based on
the Kalman filter once a prior is specified. The prior is

\begin{equation*}
(\delta_1, \gamma_1)'
\sim N(a_1,P_1),
\end{equation*}

where $a_1$ is a $2\times 1$ vector of prior means and $P_1$ a
$2\times 2$ prior variance matrix. We use a diffuse prior due to the
absence of prior information about the epidemic when the model is first
estimated: i.e., we set $a_1 = (0,0)'$, $P_1=\kappa I$, and let
$\kappa \to \infty.$ Model estimation, including implementation of the
diffuse prior, is carried out using the `KFAS` package @Helske2017.

The Kalman filter outputs estimates of the state vector
$(\delta_{t},\gamma_{t})^{\prime }.$ The estimates at time $t$
conditional on information up to and including time $t$ are denoted
$(\hat\delta_{t \mid t},\hat\gamma_{t \mid t})^{\prime }$ and given by
the contemporaneous filter; the predictive filter estimates the state at
time $t+1$ from the same information set, outputting
$(\hat\delta_{t+1 \mid t},\hat\gamma_{t+1 \mid t})^{\prime }.$

It may be useful to review past movements of the state vector
$(\delta_{t},\gamma_{t})^{\prime }.$ This can be done using the smoothed
estimates $(\hat\delta_t,\hat\gamma_t)^{\prime}$, which denotes the
estimates of the state vector at time $t$ based on all $T$ observations
in the series.

Estimation of the unknown variance parameters ($\sigma^2_{\varepsilon}$
and $\sigma^2_{\zeta}$) is by maximum likelihood (ML) and is carried out
using `KFAS` following the procedure described in @Helske2017. We retain
the option of either estimating the signal-to-noise ratio $q_{\zeta}$,
or of fixing it at a plausible value. In practice, for coronavirus
applications, we set the value of $q_{\zeta}$ based on experience and
judgment, reducing the number of parameters to be estimated by one.
Tests for normality and residual serial correlation are based on the
standardized innovations, that is one-step ahead prediction errors,
$v_{t}=\ln g_{t}-\delta_{t \mid t-1},$ $t=3,...,T.$

Daily effects, which are generally quite pronounced in the coronavirus
data, can be included in the model as described in the Appendix.

## Forecasts and peak prediction

Forecasts of future observations are obtained from the predictive
recursions

\begin{align}
\nonumber
\widehat{g}_{T+\ell \mid T} &=\exp (\hat\delta_{T\mid T}+\hat\gamma
_{T\mid T}\ell ),\;\;\;\;\ell =1,2,...\\
\widehat{\mu }_{T+\ell \mid T} &=\widehat{\mu }_{T+\ell -1\mid T}(1+
\widehat{g}_{T+\ell \mid T})
\end{align} so that \begin{align}
\widehat{y}_{T+\ell \mid T}&=\widehat{g}_{T+\ell \mid T}\widehat{\mu }
_{T+\ell -1\mid T} = Y_T \exp\hat\delta_{T+\ell|T} \prod_{j=1}^{\ell-1}(1+\exp \hat\delta_{T+j|T}) & (5)
\end{align} and
$\widehat{Y}_{T+\ell \mid T}=\widehat{\mu }_{T+\ell \mid T};$ the
initial value is $\widehat{\mu }_{T\mid T}=Y_{T}.$

We construct state-implied forecast bands for $y_t$ by propagating the
predictive distribution of the latent state vector. In the
one-step-ahead case the implied distribution for $y_{T+\ell}$ is
lognormal, but beyond one step the cumulative structure of the model
means that the forecast path is path-dependent. The bands should
therefore be interpreted as state-implied forecast bands rather than as
calibrated prediction or confidence intervals.[^2]

[^2]: These are not proper prediction or confidence intervals for
    $y_{T+\ell}$. The one-step-ahead predictive distribution of
    $\hat{y}_{T+\ell|T}$ (for $\ell=1$) is lognormal. This is not the
    case more than one step ahead due to the presence of the cumulative
    total in equation (2).

The filtered growth rate $\hat{g}_{y,t\mid t}$ of new cases $y_{t}$, can
be extracted from the continuous-time incidence curve:
$\mu^{\prime}(t) = g(t) \mu(t)$, where $\mu(t)$ is the growth curve and
$g(t)$ is its growth rate. Taking logarithms and differentiating we get

\begin{align}
\hat{g}_{y,t\mid t}&=\hat{g}_{t\mid t}+\hat\gamma_{t\mid t},   & (6)
\end{align}

where $\hat{g}_{t\mid t}=\exp \hat\delta_{t\mid t}.$ The sampling
variability of $\hat{g}_{t\mid t}$ is dominated by that of
$\hat\gamma_{t\mid t}$ (see @Harvey2021a). Therefore when constructing
confidence intervals for $\hat{g}_{t\mid t}$ we treat $\hat{g}_{y,t}$ as
if it has a normal distribution centered on $\hat{g}_{y,t\mid t}$ with
variance $\text{Var}(\hat\gamma_{t\mid t})$.

Even when the nowcast $\hat{g}_{y,T\mid T}$ is positive and daily cases
are growing, there will be a saturation level for the cumulative total,
$Y_{t},$ so long as $\hat\gamma_{T\mid T}$ is negative. The nowcasts of
$y_{t}$ peak when $\hat{g}_{y,t\mid t}=0,$ which requires
$\hat\gamma_{t\mid t}$ to be sufficiently negative to outweigh
$\hat{g}_{t\mid t},$ which is, of course, always positive. This can be
seen from the expression for the growth rate of daily cases:
\begin{align}
\hat{g}_{y,T\mid T}&=\exp \hat\delta_{T\mid T}+\hat\gamma_{T\mid T}=\hat{g}_{T\mid T}+\hat\gamma_{T\mid T}. & (7)
\end{align}

When $\hat\gamma_{T\mid T}$ is negative, there is a flattening of the
curve and a signaling of an upcoming peak in the trend of $y_{t}.$ As
shown in [HK, p10], the peak in the trend is predicted to be
$\ell _ {T}$ days ahead where[^3]

[^3]: Note the change in sign of $\gamma$ as compared with HK.

$$\ell_{T}=\frac{\ln (-\hat\gamma_{T\mid T})-\hat\delta_{T\mid T}}{\hat\gamma_{T\mid T}}=\frac{\ln (-\hat\gamma_{T\mid T}/\hat{g}_{T\mid T})}{\hat\gamma_{T\mid T}},\;\;\;\;
-\hat{g}_{T\mid T}<\hat\gamma_{T\mid T}<0.
$$

The generation of forecasts is demonstrated in Section 4.

## Reproduction Number $R_t$

The path of the epidemic is best tracked by nowcasts and forecasts of
$g_{y,t}$, the growth rate of $y_t$, which are constructed by HK from
the filtered estimates in the state space model, (2), (3) and (4).
@WallingaLipsitch2007 describe how the estimates of $g_{y,t}$ can be
translated into estimates of the instantaneous reproduction number
$R_{t}.$ @Harvey2021a propose \begin{align}
\widetilde{R}_{t,\tau}&=1+\tau g_{y,t\mid t}\;\;\;\text{or}\;\;\;\widetilde{R}_{\tau,t}^{e} =\exp(\tau g_{y,t\mid t}),\;\;\; & (8)
\end{align} \noindent where $\tau$ is the generation interval -- the
typical number of days between an infected person becoming infected and
them transmitting the disease to someone else. We construct approximate
pointwise uncertainty bands for $\widetilde{R}_{t,\tau}$ and
$\widetilde{R}_{\tau,t}^{e}$ by substituting the upper and lower bounds
of the pointwise uncertainty bands for $g_{y,t}$ into (8). These are
approximate bands for the latent reproduction-number summaries implied
by the fitted state-space model, conditional on the chosen generation
interval $\tau$; they are not Bayesian credible intervals. See
@Harvey2021b for an application.

The estimates of $R_{t}$ can be used for tracking and forecasting the
epidemic. The nowcasts of $y_{t}$ peak when $\hat{g}_{y,t\mid t}=0$,
corresponding to $\widetilde{R}_{t,\tau}=\widetilde{R}_{\tau,t}^{e}=1.$
Based on (7), predictions of $g_{y,t}$ are given by

\begin{align}
\hat{g}_{y,T+\ell\mid T} & =\exp\hat\delta_{T+\ell\mid T}+\hat\gamma_{T+\ell\mid T}=\exp(\hat\delta_{T\mid T}+\hat\gamma_{T\mid T}\ell)+\hat\gamma_{T\mid T},\;\;\;\;\;\ell=1,2,... & (9)
\end{align}

We can then obtain predictions of $R_{t},$ as in (8). If
$\hat\gamma_{T\mid T}$ is zero, the estimated growth of $y_{t}$ is
exponential and it is helpful to characterize it by the doubling time,
$\ln2/\hat{g}_{y,T\mid T} =0.693\exp(-\hat\delta_{T\mid T}).$

When $\exp \hat\delta _{T\mid T}+\hat\gamma _{T\mid T}>0$, the nowcast
$\hat{g}_{y,T\mid T}$ is positive and the estimate of $R_{t}$ given by
(8) is greater than one. So long as $\hat\gamma _{T\mid T}$ is negative,
then as $T\rightarrow \infty ,$
$\widetilde{R}_{\tau ,T+\ell \mid T}^{e}\rightarrow \exp (\tau \hat\gamma _{T\mid T})<1,$
and a saturation level for $Y$ appears on the horizon.

We now turn to case where $\gamma _{t}$ potentially turns positive in a
typically short-lived phase, as a new wave emerges.

# Reinitialization

The coronavirus pandemic was characterized by multiple waves punctuated
by plateaus. At the beginning of a new wave the growth rate of daily
cases, $g_{y,{t}}$, turns positive. The initial surge may be explosive
to the point where the growth is *super exponential*. In this case,
$\gamma _{t}$, the growth rate of $g_t$ (which is the growth rate of
cumulative cases) can also turn positive, with no peak in prospect for
$y_{t}.$ Such a phase can be expected to be transient, with
$\gamma _{t}$ dropping back to zero (exponential growth in infection,
accompanied by an upcoming peak in $y_{t}$), and then falling below zero
(sub-exponential growth in infection).

From the point-of-view of forecasting an epidemic, a peak must be in
prospect even if it can only be expected some way into the future. There
is thus a need for a solution to the problem of the estimated
$\hat\gamma _{t\mid t}$ rising to positive values as it adapts to the
upward surge in $y_{t}$, and remaining positive for any protracted
period. This upward shift in $\hat\gamma _{t\mid t}$ can be averted by
*reinitializing* the $\ln g_{t}$ series at the start of a new wave. This
involves setting the cumulative total of cases $Y_t$ back to zero at, or
around, the start of a new wave and setting $\gamma_t$ to zero so as to
impose exponential growth. From the point-of-view of the relationship
$g_{y_{t}} = g_{t} + \gamma _{t},$ the re-initialization effectively
shifts *surplus* $\gamma_t$ emanating from super-exponential growth,
into $\delta_t$ and therefore into $g_t$ (since $g_t = \exp \delta_t$).
Note that, on the date of the re-initialization, $g_{y,t} = g_t$, since
$\gamma_t=0$, and both $g_{y,t}$ and $g_t$ will be high because a new
wave is taking off.

## Reinitializing the data series

In practice the reinitialization rule is implemented in two stages,
following a real-time trigger condition. Let $T$ denote the current
forecast origin and let $\mathcal{F}_T$ denote the information available
up to and including date $T$. Write $P_{\gamma,T\mid T} = Var(\gamma_T
\mid \mathcal{F}_T)$ for the relevant element of $P_{T\mid T}$, so that
$\sqrt{P_{\gamma,T\mid T}}$ is the conditional standard deviation of the
latent slope $\gamma_T$ given the information available at $T$. Define
the standardised signal
$$z_{\gamma,T} = \frac{\hat\gamma_{T\mid T}}{\sqrt{P_{\gamma,T\mid T}}}.$$
The reinitialization warning is triggered when
$$z_{\gamma,T} > c_{reinit}, \qquad c_{reinit} = 2,$$
that is, when $\hat\gamma_{T\mid T}$ exceeds zero by approximately two
conditional standard deviations. This threshold is an operational rule
of thumb rather than a fixed statistical decision rule, and can be
adjusted depending on the noisiness of the data and the desired
trade-off between early warning and false reinitialization.

Let $d$ denote the first forecast origin at which this trigger condition
holds. The reset date $r$ supplied to `reinit.date` is then defined as
the start of the current positive spell in the smoothed slope estimate
computed on the sample ending at $d$:
$$r = \min\{t \le d : \hat\gamma^S_{u\mid d} > 0 \text{ for all } u=t,\ldots,d\},$$
where $\hat\gamma^S_{u\mid d}$ denotes the smoothed estimate of the slope
at date $u$, using observations only up to and including date $d$. Thus
the trigger date $d$ and the reset date $r$ are distinct: $d$ is the
date on which the warning first fires in real time, whereas $r$ is the
earlier date at which the cumulative series is reinitialized. In the
notation below, $t=r$ denotes the reset date and $r_0$ the date at which
the cumulative series is set to 0. Then:

\begin{align}
\ln g_t &= \ln y_t - \ln Y_{t-1} & t=1, \ldots, r & \notag \\
\ln g_t^r &= \ln y_t - \ln Y_{t-1}^r & t=r+1, \ldots, T & (10)\\
Y_{t}^{r}&=Y_{t-1}^{r}+y_{t}  & t=r,\ldots,T &   (11)
\end{align}

where $Y_{t}^{r}$ denotes the cumulative cases after re-initialization.
We set $Y_{r-1}^{r}=0$, so that the growth rate of cumulative cases is
available from $t=r+1$ onwards. Note that $Y_{t}^r = Y_{t} - Y_{r_0}$.

The gap between the two series becomes apparent by writing

\begin{equation}
\ln g_t^r = \ln g_t +  \ln \frac{Y_{t-1}}{Y_{t-1}^r}
= \ln g_t +  \ln \frac{Y_{t-1}}{Y_{t-1} - Y_{r_0}} \;\;\;\; t=r+1, \ldots, T  \;\;\;\;\;\;\;\; (12)
\end{equation}

In the next section, where we illustrate the working of the program, it
can be seen that in contrast to the original $\ln g_t$ series, which
continues to increase, the reinitialized $\ln g_t$ series begins to
decrease from the reinitialization date. The reinitialization enforces
the canonical Gompertz curve with the log of growth rate of cumulative
cases sloping down.

## Reinitializing the model

We reinitialize the model by specifying the appropriate prior
distribution for the initial states: $\alpha^r_1 \sim N(a^r_1,P^r_1)$
with $a_1^r = (a_{\delta,1}^r,a_{\gamma,1}^r)'$ and $P_1^r$ defined as
follows. Let $\mathcal{F}_t = \ln g_t, \ln g_{t-1}, \ldots, \ln g_1$ and
define $a_t = E(\alpha_t|\mathcal{F}_{t-1})$,
$a_{t|t} = E(\alpha_t|\mathcal{F}_t)$,
$P_t = Var(\alpha_t|\mathcal{F}_{t-1})$,
$P_{t|t} = Var(\alpha_t|\mathcal{F}_t).$ Then,

\begin{align*}
a^r_{\delta,1} &= a_{\delta,r+1} + \ln(Y_r/y_r) \\
a^r_{\gamma,1} &= 0\\
P^r_1 &= P_{r+1},
\end{align*}

where $a_{\delta,r+1}$ and $P_{r+1}$ are obtained from the
non-reinitialized model estimated over $t=1, \ldots, r$ via the usual
Kalman filter recursions. Adding $\ln(Y_r/y_r)$ to $a_{\delta,r+1}$
corrects for the shift down in the log cumulative cases caused by
reinitializing the cumulative case series. Setting $a^r_{\gamma,1} = 0$
ensures the model starts off with exponential, rather than
super-exponential, growth.

We reinitialize the model through the priors in this way rather than
simply re-estimating the model from scratch for two reasons. First, it
allows us to impose a proper (rather than diffuse) prior centered on
zero for $\gamma$, so that the starting point is exponential growth.
Second, it enables us to make use of data from before the
reinitialization date. One needs a reasonable sample size for the
estimated model and forecasts to be reliable, but if a new wave is
taking off, forecasts need to be generated quickly. This was
particularly true with the emergence of the Omicron variant, which
caused an explosive increase in infection over a short period of time.

We do not re-estimate the $\sigma^2_{\varepsilon}$ or $\sigma^2_{\zeta}$
parameter in the reinitialized model. Rather, we use the values
estimated in the original model over $t=1, \ldots, r$. The
one-step-ahead prediction error at $t=r$ is the same in both the
initialized and reinitialized models, but after $t=r$, the prediction
errors diverge.

The reinitialization procedure is very similar in the case where we have
seasonal terms. If we let $\alpha_{s,t}$ be the vector of seasonal
states and maintain an analogous notation to that above, the prior mean
of the seasonal components in the reinitialized model is
$$a^r_{s,1} = a_{s,r+1}.$$ The prior variance of $\alpha^r_1$ remains
$P_{r+1}$ where $P_{r+1}$ is appropriately re-defined to include the
seasonal term, as described in the Appendix.

# Functionality of `tsgc`

The main Gompertz class in `tsgc` is `SSModelDynamicGompertz`, which
implements the models described in (2)-(4), with or without
reinitialization. A model object is constructed with
`SSModelDynamicGompertz$new(...)` or `tsgc::SSModelDynamicGompertz(...)`
and its unknown parameters are estimated with the `estimate()` method,
invoked as `estimate(model)`, which returns an object of the
`FilterResults` class. This class wraps the underlying `KFAS` `KFS`
object, together with additional methods for prediction, and an
optional `idx_calendar` (see below) attached.

The `SSModelDynamicGompertz` needs only a cumulative series `Y` as an
input, supplied as an `idx_series` object -- an integer-position
representation of a time series, consisting of a vector or matrix of
values together with a `start` position. In our application, `Y` is the
cumulative number of new coronavirus cases. The estimation window is
specified directly on the model object via the `start` and `end`
arguments, which are integer positions in `Y` rather than calendar
dates. There is an option to specify the signal-to-noise ratio
$q_{\zeta}$, rather than estimate it, and an option to specify the
model to have a seasonal component through the `sea.period` argument.
The period of a seasonal component is specified through the
`sea.period` option.

`idx_series` positions carry no calendar meaning of their own: they are
plain integers, and the model, its estimation, and its forecasts are
defined purely in terms of them. Real calendar time is reintroduced only
as an optional, purely cosmetic layer for plotting, via an
`idx_calendar` object passed through the `calendar` argument. An
`idx_calendar` records an anchor position and the corresponding
calendar date (or other real-world reference point), together with the
step size and unit between consecutive positions; it is never consulted
during estimation or filtering. Supplying a `calendar` therefore changes
nothing about the fitted model -- only whether its plots are labelled
with dates or with raw integer positions. Two helper functions,
`xts_to_idx()` and `idx_to_pos()`, are provided to ease the transition
from calendar-time data: the former converts an `xts`/`Date`-indexed
series into an `idx_series` plus a matching `idx_calendar` in one step,
and the latter translates a calendar date back into the integer
position `idx_series` expects for `start`/`end`/`reinit.idx`. Both are
demonstrated in the Illustration section below. Every calendar used in
this vignette has a step size that is a single, uniform `amount`/`unit`
pair; `idx_calendar` also supports compound and heterogeneous step
sizes for series that need them, described in the Appendices
"Compound calendar steps" and "Heterogeneous repeating cycles" below.

The `SSModelDynamicGompertz` class allows the model to be estimated for
a new wave without losing information from prior waves. It will accept
the reinitialization position specified by `reinit.idx` (an integer
position, rather than the calendar date `reinit.date` used previously)
or a `FilterResults` object from which it can extract the initial
values. If the user wishes to reinitialize the model without using
prior information (i.e. treat the new wave as an entirely separate
epidemic), a reinitialization position can be specified through the
`reinit.idx` option and `use.presample.info` can be set to `FALSE`.

The `FilterResults` class contains prediction methods which can be
applied to estimated dynamic Gompertz curve models (both reinitialized
and non-reinitialized). `get_growth_y()` returns filtered or smoothed
estimates of the growth rate of new cases ($g_t$), while `get_gy_ci()`
returns the same with uncertainty bands. Forecasts of the incidence
variable (new cases, $y_t$) can be obtained with the `predict_level()`
call, and forecasts of all the states can be obtained with the
`predict_all()` call.

Several functions are available to generate plots of smoothed and
filtered estimates and forecasts. `plot_log_forecast()` plots the
forecast and realised values of $\ln(g_t)$. `plot_gy_components()` and
`plot_gy_ci()` can be used to plot the smoothed or filtered growth rate,
its components, and uncertainty bands, respectively. Forecasts of the
incidence variable ($y_t$) and forecast intervals can be plotted using
`plot_forecast()`, while `plot_holdout()` adds plots of prediction
intervals and of realized outcomes over a holdout period to help
evaluate forecast accuracy. `plot_compare_forecast()` overlays the
forecasts from a list of fitted models -- for example a reinitialized
and non-reinitialized fit, or a dynamic Gompertz model against a
leading-indicator model -- on the same axes for visual comparison.
Finally, the `reinitialise_dataframe()` function can be used to
reinitialise a data series at a given `reinit.idx`.

Every one of these plotting functions accepts an `axis` argument, of
class `idx_axis_opts`, which controls how integer positions are
translated to the x-axis and is described in full in the Appendix
"Controlling the plot x-axis" below. Left at its default (`NULL`), the
x-axis reproduces the previous behaviour: real calendar dates whenever
a suitable `calendar` is attached to the model or results object, and
plain integer positions otherwise.

More details on how to use the methods and functions described are
presented in the following section.

## Leading indicator model

The basic dynamic Gompertz model uses only the cumulative target series.
A useful extension is the leading-indicator model, in which a second
series is used to improve forecasts of an outcome of interest. The
motivation is simple: many indicators move earlier than the target, so
their information can be exploited to improve short-horizon forecasts.
In the epidemic setting, for example, hospital admissions often respond
with a lag to the underlying growth in cases, while in marketing
applications early sales of related products may foreshadow demand for a
new item. The package therefore allows the target series to be modeled
jointly with a lagged auxiliary series rather than treating the target
in isolation.

Let $\ln g_{1,t-k}$ denote the log growth rate of the leading series and
$\ln g_{2,t}$ the log growth rate of the target series. The
leading-indicator specification in `tsgc` is

\begin{align}
\ln g_{1,t-k} &= \delta_t + \psi_t + \varepsilon_{1t}, & t = k+1, \ldots, T+k, & (16)\\
\ln g_{2,t} &= \delta_t + \varepsilon_{2t}, & t = k+1, \ldots, T, & (17)
\end{align}

where $k$ is the user-specified lag, $\delta_t$ is the common trend
shared by both series and evolves exactly as in the dynamic Gompertz
model (equations (3)-(4)), $\psi_t$ is a target-independent deviation
specific to the leading equation, and $\varepsilon_{1t}$ and
$\varepsilon_{2t}$ are mutually independent Gaussian observation
disturbances. The lag $k$ is an alignment parameter: after shifting the
leading series by $k$ periods, the lead and target growth rates are
treated as two noisy measurements of a common latent growth component,
rather than the leading series entering as an exogenous regressor.
Estimation is by maximum likelihood using the Kalman filter, exactly as
for the univariate model, and the resulting object is a
`FilterResultsLI` object, which supports the same forecasting, plotting,
and diagnostic methods as the `FilterResults` class used for the dynamic
Gompertz model.

The role of `n.lag` is therefore to align the information in the leading
series with the target series, in integer positions rather than
calendar time. If the indicator is known to anticipate the target by
several periods, a larger lag is appropriate. The package also allows
the user to supply exogenous covariates through `xpred_lead` and
`xpred_targ` at estimation time, and their future values by setting the
`xpred_lead.new` and `xpred_targ.new` fields directly on the fitted
`FilterResultsLI` object before forecasting, making the model suitable
for settings where both the lead and target series are influenced by
observed external drivers.

The empirical benefit of this extension is that it can improve
short-horizon predictions when the relationship between the two series
is stable. In practice, one usually compares several lag values and
evaluates them using a rolling-origin experiment, because the
appropriate lead-lag relationship is seldom known in advance. Worked
examples of the leading-indicator model, including the England
cases-hospitalisations application, a comparison against the dynamic
Gompertz model, lag selection by rolling cross-validation, and
extensions to other data frequencies, are given in the Illustration
section below.

# Forecast accuracy and rolling-origin validation

Forecast accuracy can be evaluated with a variety of scale-dependent and
scale-free measures. In the notation of the package, the mean absolute
percentage error is
$$\mathrm{MAPE}=\frac{100}{n}\sum_{t=1}^n \left|\frac{y_t-\hat y_t}{y_t}\right|,$$
while the symmetric MAPE is defined as
$$\mathrm{sMAPE}=\frac{100}{n}\sum_{t=1}^n \frac{2|y_t-\hat y_t|}{|y_t|+|\hat y_t|},$$
and the mean absolute error and root mean squared error are given by
$$\mathrm{MAE}=\frac{1}{n}\sum_{t=1}^n |y_t-\hat y_t|, \qquad \mathrm{RMSE}=\sqrt{\frac{1}{n}\sum_{t=1}^n (y_t-\hat y_t)^2}.$$
sMAPE is reported as the primary percentage-based comparison measure in
the empirical illustrations above, since MAPE can become unstable when
observed values are close to zero. The rolling-origin procedure used by
`cross_val()` re-estimates the model at successive evaluation positions,
holds back a fixed forecast horizon, and computes the chosen loss metric
over that out-of-sample window, so that forecast accuracy is assessed
under a design that matches the practical forecasting problem.

# Illustration of the `tsgc` package

In this section we provide a full working example of the `tsgc` package
in `R` which implements the modeling framework for time series growth
curves-based epidemic forecasting.

`tsgc` comes with two example data sets relating to COVID-19: one for
Gauteng province in South Africa (sourced from South Africa's official
coronavirus online news and information
[portal](https://sacoronavirus.co.za/)) and another for England (sourced
from the official UK government
[dashboard](https://ukhsa-dashboard.data.gov.uk/) for data and insights
on coronavirus). In the example that follows, we use the data on
confirmed cases in Gauteng. The data series is in cumulative form and
ships with the package as an `xts` object with a `Date` index, as
follows.

```{r load-gauteng-xts}
data(gauteng, package = "tsgc")
```

The modelling code in this version of the package works with
`idx_series` (integer-position data) rather than directly with `xts`
objects, so that estimation and forecasting are defined purely in terms
of position, independent of calendar frequency or gaps. Calendar time is
reintroduced only for plotting, via an optional `idx_calendar` that
records how integer positions map back to real dates. `xts_to_idx()`
performs this conversion in one step for the common case of a series
with no gaps, returning both the `idx_series` and a matching
`idx_calendar`; `idx_to_pos()` is its companion for translating a
calendar date into the integer position that `start`/`end`/`reinit.idx`
expect. A minimal example of converting a `CSV`-derived data frame to
this representation is given in the Appendix; the pattern below extends
directly to that case.

```{r convert-gauteng}
conv <- xts_to_idx(gauteng)
gauteng_idx <- conv$series
gauteng_cal <- conv$calendar
gauteng_cal$anchor_name <- "first recorded case"
```

New COVID-19 cases reported for Gauteng province and their centered
7-day moving average presented in Figure 1 show a sequence of four waves
over the period between 10 March 2020 and 5 January 2022.

```{r make-fig-1, fig.cap="Figure 1: New Cases and their centered 7-day moving average for Gauteng province in South Africa between 10 March 2020 and 5 January 2022."}
mod_all <- SSModelDynamicGompertz(Y = gauteng_idx, calendar = gauteng_cal)
plot(mod_all, series.name = "daily cases", MA_period = 7)
```

## Setting up the forecasting exercise

We begin by specifying a number of options for the forecasting exercise,
as defined below.

-   `Y` is the data, in the form of an `idx_series` of cumulative
    confirmed cases. In this example the object holding this series is
    called `gauteng_idx`, with `gauteng_cal` the accompanying
    `idx_calendar`.

-   `estimation.pos.start` is the integer position of the first
    observation in the sample to be used for estimating the model,
    obtained here from a calendar date via `idx_to_pos()`. By default,
    if `start` is left unspecified when constructing the model, it is
    the first position in `Y`.

-   `estimation.pos.end` is the integer position of the last observation
    in the sample to be used for estimating the model, again obtained
    via `idx_to_pos()`. By default, if `end` is left unspecified, it is
    the last position in `Y`.

-   `n.forecasts` is the number of days or periods for which forecasts
    are to be made. E.g., if `n.forecasts = 14`, forecasts will be
    generated for up to 14 positions following `estimation.pos.end`.

-   `q` is the signal-to-noise ratio, which controls the smoothness of
    the estimated trend. A lower value will lead to more smoothness. By
    default, we use `q = NULL`, so that it is estimated. Alternatively,
    we can set `q`. For example, in our experience, `q = 0.005` ensures
    a good balance between the smoothness of the trend and the speed
    with which changes in estimates respond to new observations.

-   `confidence.level` sets the coverage used to construct the
    state-implied forecast bands for $\ln(g_t)$, which are then used to
    generate the forecast bands for the forecasts themselves. Here, we
    use 0.68, corresponding to the probability that the forecast lies
    within one standard deviation of the point forecast under the
    (approximating) Normal distribution used to construct the bands --
    see the note on interpretation in the Theory section above.

-   `plt.length` sets a truncation position to enhance the clarity of
    plots, e.g. showing only the last 30 positions of the estimation
    sample. The range for plotting can be set as `plt.length` positions
    up to `estimation.pos.end`.

In this example the data is the cumulative confirmed cases time series
for Gauteng. The start and end positions (`estimation.pos.start` and
`estimation.pos.end`) that define the sample used for estimation are
chosen as appropriate for the exercise, translated from calendar dates
via `idx_to_pos()`. In this example, we begin with the sample period set
from 1 February to 3 May 2021. This marks the beginning of the third
wave in Gauteng as can be seen in Figure 1. The options are specified as
below.

```{r set_options}
Y <- gauteng_idx
cal <- gauteng_cal

estimation.pos.start <- idx_to_pos(cal, "2021-02-01")
estimation.pos.end   <- idx_to_pos(cal, "2021-05-03")
n.forecasts <- 14
q <- 0.005
confidence.level <- 0.68
plt.length <- 30
```

## Estimation

We begin by selecting the data series (`Y`) for the defined sample
period.

The estimation period is specified directly when creating the
`SSModelDynamicGompertz` object using the `start` and `end` arguments
(integer positions in `Y`), so no manual subsetting of the data is
required. Passing `calendar = cal` attaches the `idx_calendar` to the
resulting model and, in turn, to its `FilterResults` output, so that
later plots are automatically labelled with real dates rather than raw
positions.

We then estimate the model using a diffuse prior distribution for the
initial state vector. The signal-to-noise ratio can be left as a free
parameter to be estimated, as in the code below.

```{r model_est_free_q}
model_q <- SSModelDynamicGompertz(Y = Y, sea.period = 7,
                                  start = estimation.pos.start,
                                  end = estimation.pos.end,
                                  calendar = cal)
res_q <- estimate(model_q)
```

In the rest of this example we estimate the model setting the
signal-to-noise ratio at $0.005.$ As mentioned, in our experience this
value strikes a useful balance between the smoothness of the estimate of
the slope parameter $\gamma$, and the speed with which it adapts to new
observations.

```{r model_est_fix_q}
model <- SSModelDynamicGompertz(Y = Y, q = q, sea.period = 7,
                                start = estimation.pos.start,
                                end = estimation.pos.end,
                                calendar = cal)
res <- estimate(model)
```

## Model diagnostics

Before looking at forecasts, it is worth checking that the model
actually fitted sensibly. `summary()` on a `FilterResults`/
`FilterResultsLI` object reports the estimation window, model states
and their standard errors, and the estimated variance parameters
(including the signal-to-noise ratio); `tsgc::print_model_diagnostics()`
supplements this with the fitted log-likelihood, a check for estimated
variances sitting at or near a zero/degenerate boundary, and simple
recursive-residual diagnostics.

```{r model-diagnostics}
summary(res)
tsgc::print_model_diagnostics(res)
```

## Results

We can now plot the forecast of $\ln g_t$ -- the log of the growth rate
of $Y$, the cumulative cases -- which is the transformation of the
series that is taken to the model, and we can compare these forecasts to
the actual $\ln g_t$ series. We do this by passing the output (`res`) of
the estimation step along with an evaluation sample to a plotting
function `tsgc::plot_log_forecast`.

```{r fig-2, fig.cap="Figure 2: Fourteen-day forecast of $ln(g_t)$ from 4 May 2021, for Gauteng province in South Africa."}
tsgc::plot_log_forecast(
  res = res,
  Y = Y,
  n.ahead = n.forecasts,
  plt.start = tail(res$index, 1) - plt.length
)
```

From these results we can recover the forecasts of new cases from 4 May
2021, with their forecast bands.

```{r fig-3, fig.cap = "Figure 3: Fourteen-day forecast of new cases from 4 May 2021 for Gauteng province in South Africa. The shaded region is a state-implied forecast band obtained by propagating uncertainty in the latent log growth rate."}
tsgc::plot_forecast(
  res = res,
  n.ahead = n.forecasts,
  confidence.level = confidence.level,
  plt.start = tail(res$index, 1) - plt.length,
  series.name = "cases"
)
```

To assess accuracy, we plot these forecasts against the actual new
cases, that have been held back from the estimation sample, using the
`plot_holdout` function. The model forecasts are compared with the first
differences of `Y`, the cumulative series, over the forecast window.

```{r fig-4, fig.cap = "Figure 4: Accuracy of the fourteen-day forecast of new cases from 20 April 2021 for Gauteng province in South Africa."}
# Update estimation end position
estimation.pos.end <- idx_to_pos(cal, "2021-04-19")

# Reestimate the model
model <- SSModelDynamicGompertz$new(Y = Y, q = q, sea.period = 7,
                                    start = estimation.pos.start,
                                    end = estimation.pos.end,
                                    calendar = cal)
res_eval <- estimate(model)

tsgc::plot_holdout(
  res = res_eval,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

Figure 4 shows that the forecasts were accurate over the first seven
days, with a mean absolute percentage error (MAPE) of 13.9%. Note that
reported cases were unusually low on 27 April due to the fact that it is
Freedom Day and a public holiday in South Africa. That day aside, over
the six days from 28 April the MAPE was 12.8%. Over the full 14 days of
the forecasts, the MAPE was 27%.

As discussed earlier, the reproduction numbers $R_t$ and their 68%
approximate pointwise uncertainty bands can be calculated using
`estimate_r0()`, which returns an `idx_series` of the fitted $R_t$ values
together with their lower and upper bounds. These can then be plotted
directly with `plot_r0()`, as shown in Figure 5.

```{r calc_rt}
gen_int <- 4
r.t <- estimate_r0(res, gen_int, n.ahead = 7)
r.t
```

```{r fig-5, fig.cap = "Figure 5: Reproduction numbers for the 7-day period to 3 May 2021, for Gauteng province in South Africa."}
plot_r0(res, gen_int = gen_int, n.ahead = 7,
        title = "Reproduction numbers for the 7-day period to 3 May 2021")
```

`gen_int` above (4 days) is an epidemiological assumption fed into
`estimate_r0()`, not something estimated from this data, so the
resulting $R_t$ path is conditional on it. The sensitivity of $R_t$ to
this assumption can be checked directly by sweeping `gen_int` over a
plausible range and comparing the resulting point estimates on the
same dates:

```{r rt_gen_int_sensitivity}
gen_int_grid <- c(3, 4, 5, 6, 7)

r.t.sensitivity <- lapply(gen_int_grid, function(g) {
  r.t.g <- estimate_r0(res, g, n.ahead = 7)
  data.frame(
    Date    = idx_to_date(res$calendar, idx_positions(r.t.g)),
    gen_int = g,
    Rt      = as.matrix(idx_values(r.t.g))[, "fit"]
  )
})
r.t.sensitivity.df <- do.call(rbind, r.t.sensitivity)

last_date <- max(r.t.sensitivity.df$Date)
rt_last <- r.t.sensitivity.df$Rt[r.t.sensitivity.df$Date == last_date]
range(rt_last)
all(rt_last > 1) || all(rt_last < 1)
```

The last line reports whether the above/below-1 conclusion for $R_t$
on the most recent date is the same across the whole `gen_int` grid
(`TRUE`) or changes somewhere within it (`FALSE`); the numeric range
just above it gives the spread of point estimates implied by the
grid. Both should be checked before treating a single-`gen_int` $R_t$
estimate as a robust statement about transmission on that date, rather
than reported alongside the main $R_t$ figure without comment.

```{r fig-5b, fig.cap = "Figure 5b: Sensitivity of Gauteng R_t to the assumed generation interval."}
ggplot2::ggplot(r.t.sensitivity.df,
                 ggplot2::aes(x = Date, y = Rt, color = factor(gen_int))) +
  ggplot2::geom_line(linewidth = 0.6) +
  ggplot2::geom_hline(yintercept = 1, linetype = "solid", color = "black") +
  ggplot2::labs(
    title = "Sensitivity of R_t to the assumed generation interval",
    subtitle = paste0("gen_int swept over {", paste(gen_int_grid, collapse = ", "), "} days; point estimate only (no CI)"),
    x = "Date", y = expression(R[t]), color = "gen_int (days)"
  ) +
  ggplot2::theme_light(base_size = 12)
```

A CSV file (`y-forecast`) is written to the directory specified. The
forecast options specified earlier are retained.

```{r write_results, eval=F}
res.dir <- here::here(here::here(), "results")

tsgc::write_results(
 res = res,
 res.dir = res.dir,
 n.ahead = n.forecasts,
 confidence.level = confidence.level
)
```

## Reinitialization

In all countries, the coronavirus pandemic was characterized by a series
of recurring waves due to a combination of biological, behavioral, and
environmental reasons. In an epidemic, the onset of a new wave is
signalled when the slope parameter $\gamma$, which measures of the
growth rate of the growth rate of new cases, rises above zero for a
sustained period. Such a super-exponential phase of the epidemic in
which the growth rate of new cases is itself increasing over time is
typically short.

This section illustrates the reinitialization procedure which allows us
to apply the model to the new wave as it begins, without jettisoning
information from the wave that has just ended. We extend the estimation
window to 25 June 2021, by which date the third wave is well on course
with its peak within sight (see Figure 1). All other options remain the
same.

```{r change-est-date-end}
estimation.pos.end <- idx_to_pos(cal, "2021-06-25")
```

### Triggering reinitialization

As set out above, reinitialization is triggered at the first forecast
origin $T$ at which the standardised signal $z_{\gamma,T} =
\hat\gamma_{T\mid T}/\sqrt{P_{\gamma,T\mid T}}$ exceeds the threshold
$c_{reinit}=2$, that is, when the estimate of the slope parameter,
$\gamma_t$, breaches two standard errors above zero. Reinitialization is
then backdated to $r$, the start of the current positive spell in the
smoothed slope estimate computed on the sample ending at the trigger
position. In applying this rule there is a choice between the filtered
slope estimate and the smoothed slope estimate. Experimental results
suggest that the greater noise in the filtered estimate of $\gamma_t$
often triggers reinitialization too early. The smoothed estimate is more
reliable, and is what we use to compute $z_{\gamma,T}$ below.

Figure 6 shows that for the third wave in Gauteng, the smoothed slope
estimate exceeded twice its standard error on 1 May 2021, having risen
above zero on 21 April 2021, giving trigger date $d =$ 1 May 2021 and
reset date $r =$ 21 April 2021.

```{r fig-6, fig.cap = "Figure 6: Trigger for reinitialization: reinitialization is triggered when the two-standard-error lower threshold of the smoothed slope (slope - 2*se) crosses zero. Lines shown are k-SE thresholds, not confidence bands around the slope. Reset date selected from a single retrospective episode; not a validated general rule."}
# Re-estimate model
model <- SSModelDynamicGompertz$new(Y = Y, q = q, sea.period = 7,
                                    start = estimation.pos.start,
                                    end = estimation.pos.end,
                                    calendar = cal)
res <- estimate(model)

# Extract the smoothed slope and its standard deviation as idx_series,
# then convert to a data.frame with real dates for plotting.
# Use the smoothed-state covariance V (matched to smoothed alphahat),
# not output$P (filtered Ptt) -- see accompanying critical assessment.
smoothed.slope.full <- idx_series(res$output$alphahat[, "slope"], start = res$index[1])
V.smoothed <- get_V(res$output)
i.slope <- grep("slope", colnames(res$output$alphahat))
smoothed.P.slope <- idx_series(V.smoothed[i.slope, i.slope, ], start = res$index[1])

common_pos <- idx_positions(smoothed.slope.full)
d2 <- idx_series(
  cbind(
    smoothed.slope = idx_values(smoothed.slope.full),
    sd.smoothed.slope = sqrt(idx_values(smoothed.P.slope)),
    sd.smoothed.slope.1.5 = 1.5 * sqrt(idx_values(smoothed.P.slope)),
    sd.smoothed.slope.2 = 2 * sqrt(idx_values(smoothed.P.slope))
  ),
  start = common_pos[1]
)

d2.mat <- as.matrix(idx_values(d2))
d2.df <- data.frame(
  Date = idx_to_date(cal, idx_positions(d2)),
  smoothed.slope = d2.mat[, "smoothed.slope"],
  sd.smoothed.slope = d2.mat[, "sd.smoothed.slope"],
  sd.smoothed.slope.1.5 = d2.mat[, "sd.smoothed.slope.1.5"],
  sd.smoothed.slope.2 = d2.mat[, "sd.smoothed.slope.2"]
)

d2.df <- d2.df[d2.df$Date >= as.Date("2021-02-15"), ]

# zt = smoothed slope minus its own two-standard-error threshold (the
# lower bound of the two-SE interval). Trigger = sign change in zt
# itself, not a comparison of lagged slope against the current
# threshold (not equivalent once the SE changes over time).
d2.df$zt <- d2.df$smoothed.slope - d2.df$sd.smoothed.slope.2

trigger.df <- d2.df %>%
  mutate(prev_zt = lag(zt)) %>%
  filter(zt > 0 & prev_zt <= 0)
# Triggered when the two-SE lower bound of the smoothed slope crosses zero

reinit_zero.df <- d2.df %>%
  mutate(prev_smoothed.slope = lag(smoothed.slope)) %>%
  filter(Date < min(trigger.df$Date) &
           (smoothed.slope > 0 & prev_smoothed.slope < 0)) %>%
  arrange(desc(Date)) %>%
  slice(1)
# Reinitialisation on April 21. NOTE: this reset date is selected
# retrospectively from a single episode/forecast origin and does not
# validate the reinitialisation rule as a general real-time trigger.

# Create plot
ggplot(data = d2.df,
      aes(x = Date)) +
  geom_line(aes(y = smoothed.slope,
                color = "smoothed.slope"), linewidth=.5) +
  geom_line(aes(y = sd.smoothed.slope,
                color = "sd.smoothed.slope"),
            linetype = "solid",
            linewidth=.25) +
  geom_line(aes(y = sd.smoothed.slope.1.5,
                color = "sd.smoothed.slope.1.5"),
            linetype = "solid",
            linewidth=.25) +
  geom_line(aes(y = sd.smoothed.slope.2,
                color = "sd.smoothed.slope.2"),
            linetype = "solid",
            linewidth=.5) +
  scale_y_continuous(n.breaks = 10) +
  geom_hline(yintercept = 0,
             linetype = "solid",
             color = "black",linewidth = 1)+
  geom_vline(data = trigger.df,
             aes(xintercept = Date),
             linetype = "solid", size = .5,
             color = "black")+
  ylab("Slope") +
  scale_x_date(date_breaks = "10 days") +
  scale_color_manual(name='',
                     values=c('smoothed.slope'='red',
                              'sd.smoothed.slope'='blue',
                              'sd.smoothed.slope.1.5'='green',
                              'sd.smoothed.slope.2'='black'))+
  theme_light(base_size = 11)+
  #theme(legend.position = c(.88, 0.25))+
  theme(legend.title=element_text(size=2),
        legend.text=element_text(size=6))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        plot.title = element_text(face = "bold"))
```

The position for reinitialization is set accordingly, translating the
reset date to an integer position via `idx_to_pos()`.

```{r set_reinit_date}
reinit.pos <- idx_to_pos(cal, "2021-04-21")
```

### Estimating the reinitialized model

A reinitialized `SSModelDynamicGompertz` uses the same arguments as the
non-reinitialized model, with the addition of the `reinit.idx` argument
-- an integer position, in place of the calendar date `reinit.date` used
previously.

```{r reinit_estim}
model <- SSModelDynamicGompertz$new(
  Y = Y, q = q, sea.period = 7,
  start = estimation.pos.start,
  end = estimation.pos.end,
  reinit.idx = reinit.pos,
  calendar = cal
)
res.reinit <- estimate(model)
```

We generate the reinitialized data series by setting cumulative cases to
$0$ at the appropriate point, as discussed in the Theory section, and
extract the evaluation sample from the reinitialized series as below.

Estimating the model with the reinitialized series, the actual and
forecast $\ln(g_t)$ can be plotted as in Figure 7, which is analogous to
Figure 2 for the non-reinitialized series.

```{r fig-7, fig.cap = "Figure 7: Forecast of $ln(g_t)$ after reinitialization."}
tsgc::plot_log_forecast(
    res = res.reinit,
    Y = Y,
    n.ahead = n.forecasts,
    plt.start = tail(res.reinit$index, 1) - plt.length,
    title='Forecast of log growth rate after reinitialization.'
)
```

Comparing forecasts, without reinitialization the 14-day forecast MAPE
is 41.9% (see Figure 8). With reinitialization, it falls to 20.2% (see
Figure 9). The corresponding MAPE figures for 7-day forecasts is 15.2%
without reinitialization and 9.5% with reinitialization.

```{r fig-8, fig.cap = "Figure 8: Forecast accuracy of the model without reinitialization over the hold-out sample period: 14 days from 25 June 2021."}
tsgc::plot_holdout(
  res = res,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

```{r fig-9, fig.cap = "Figure 9: Forecast accuracy of the model with reinitialization over the hold-out sample period: 14 days from 25 June 2021."}
tsgc::plot_holdout(
  res = res.reinit,
  Y = Y,
  n.ahead = 14,
  confidence.level = 0.68,
  series.name = "cases"
)
```

We can also compare the standard (no reinitialization) and reinitalized
forecasts in a single plot:

```{r fig-10, fig.cap = "Figure 10: Comparison of standard (green) and reinitialized (blue) forecasts"}
tsgc::plot_compare_forecast(
  results = list(res.reinit, res),
  n.ahead = 14,
  actual = Y
)
```

Just as for the standard (non-reinitialized) model, the returned
estimation results contained in `res.reinit` are a `FilterResults`
object, and can be written to CSV using the `write_results` function.

## Leading indicator model: England cases and hospitalisations

We now illustrate the leading-indicator model described above. We load
the England data, which contains daily cumulative COVID-19 cases and
hospital admissions, convert it to `idx_series`/`idx_calendar` form, and
specify an estimation window running from 30 April 2021 to 24 July 2021,
with cases as a four-day leading indicator for hospitalisations.

```{r li-setup}
data(england, package = "tsgc")

conv <- xts_to_idx(england[, 1:2])
eng <- conv$series
eng_cal <- conv$calendar

est.start.eng <- idx_to_pos(eng_cal, "2021-04-30")
est.end.eng   <- idx_to_pos(eng_cal, "2021-07-24")
n.lag.eng     <- 4
n.forecasts.eng <- 7
plt.length.eng  <- 14
```

A preliminary plot of the two series on the log scale is useful for
inspecting comovement before formal estimation.

```{r li-plot, fig.cap = "Figure 11: Daily COVID cases and hospitalisations, England."}
mod.eng.plot <- SSModelLeadingIndicator$new(eng, n.lag = n.lag.eng, calendar = eng_cal)
plot(
  mod.eng.plot,
  title = "Daily COVID cases and hospitalisations (England)",
  series.name.lead = "Cases",
  series.name.target = "Hospitalisations",
  take.log = TRUE
)
```

We then estimate the baseline leading-indicator model, with cases as the
lead series (`LeadIndCol = 1`) and weekly seasonality.

```{r li-estimate}
mod.eng <- SSModelLeadingIndicator$new(
  Y = eng,
  n.lag = n.lag.eng,
  LeadIndCol = 1,
  sea.period = 7,
  start = est.start.eng,
  end = est.end.eng,
  calendar = eng_cal
)
res.eng <- estimate(mod.eng)
```

`FilterResultsLI`'s `summary()` prints the same kind of diagnostics as
`FilterResults$summary()` above, but its observation-equation noise `H`
is a $2\times2$ matrix (one variance per equation, for the lead and
target series) rather than a single scalar; it is printed as a matrix
accordingly, and `print_model_diagnostics()` handles a matrix-valued `H`
the same way when checking for boundary/degenerate variances.

```{r li-diagnostics}
summary(res.eng)
tsgc::print_model_diagnostics(res.eng)
```


The fitted object `res.eng` can be used to generate the same forecast
and holdout panels as for the dynamic Gompertz model.

```{r fig-li-log, fig.cap = "Figure 12: Forecast of log growth rate of hospitalisations, England."}
tsgc::plot_log_forecast(
  res.eng,
  Y = eng,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  title = "Forecast of log growth rate of hospitalisations (England)"
)
```

```{r fig-li-forecast, fig.cap = "Figure 13: Forecast of hospitalisations, England."}
tsgc::plot_forecast(
  res.eng,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  series.name = "Hospitalisations",
  title = "Forecast of hospitalisations (England)"
)
```

```{r fig-li-holdout, fig.cap = "Figure 14: Accuracy of the seven-day forecast of hospitalisations, England."}
tsgc::plot_holdout(
  res.eng,
  Y = eng,
  n.ahead = n.forecasts.eng,
  series.name = "Hospitalisations",
  title = "Accuracy: forecast of hospitalisations (England)"
)
```

### Adding regressors and future covariate paths

The same specification can be enriched with exogenous predictors. Here
we add contemporaneous weather variables (wind speed, relative humidity,
temperature, and maximum temperature) to both the lead and target
equations, and then supply their future values before forecasting. Where
the previous version of the package required a dedicated
`supply_xpred.new()` call, the corresponding fields --
`xpred_lead.new` and `xpred_targ.new` -- are now set directly on the
fitted `FilterResultsLI` object.

***Oracle-forecast caveat.*** The future regressor values supplied below
are REALISED (observed after the fact) weather from `england_weather_idx`,
covering the same forecast horizon being predicted -- not an archived
weather forecast that would actually have been available at the
forecast origin. This makes the resulting forecast a conditional
("oracle") exercise, not an operational validation: it shows what the
model would have predicted GIVEN perfect future weather knowledge, not
what it would have predicted using information available at the time.
A genuine operational evaluation would require archived point/ensemble
weather forecasts as of each origin date, which are not supplied here.

```{r li-xpred}
data(england_weather_2021, package = "tsgc")

conv <- xts_to_idx(england_weather_2021[, 1:4],
                    start.pos = idx_to_pos(eng_cal, zoo::index(england_weather_2021)[1]))
england_weather_idx <- conv$series

xpred_lead <- xpred_targ <- england_weather_idx

mod.eng.x <- SSModelLeadingIndicator$new(
  eng,
  n.lag = n.lag.eng,
  xpred_lead = xpred_lead,
  xpred_targ = xpred_targ,
  sea.period = 7,
  start = est.start.eng,
  end = est.end.eng,
  calendar = eng_cal
)
res.eng.x <- estimate(mod.eng.x)

# *** ORACLE-FORECAST CAVEAT ***
# england_weather_idx covers the forecast horizon itself, so this
# supplies REALISED (observed) weather over that horizon, not an
# archived forecast that would have been available at the forecast
# origin. Figures below are conditional/oracle forecasts, not an
# operational validation -- see prose above.
res.eng.x$xpred_lead.new <- england_weather_idx
res.eng.x$xpred_targ.new <- england_weather_idx
```

```{r fig-li-x-forecast, fig.cap = "Figure 15: Forecast of hospitalisations with weather regressors (weather, oracle/realised), England."}
tsgc::plot_forecast(
  res.eng.x,
  n.ahead = n.forecasts.eng,
  plt.start = est.end.eng - plt.length.eng,
  title = "Forecast of hospitalisations\nwith regressors (weather, oracle/realised)",
  series.name = "Hospitalisations"
)
```


In practice, the recommended sequence is to establish a sensible lag and
seasonal specification first, inspect the resulting forecasts, and only
then augment the model with regressors if additional covariates are
available and future paths can be supplied.

## Comparing the dynamic Gompertz and leading-indicator models

We now compare a UK-only dynamic Gompertz model with a leading-indicator
model that uses Italian case growth as a lead for UK case growth, using
daily COVID-19 cases in Italy and the UK from the first-wave period.
Italy experienced the first large European outbreak earlier than the UK,
making it a natural candidate for a short-horizon lead series during the
first wave.

```{r cv-setup}
data(ukitaly, package = "tsgc")

conv <- xts_to_idx(ukitaly)
ukitaly_idx <- conv$series
ukitaly_cal <- conv$calendar

n.forecasts.uk <- 14
plt.length.uk  <- 30
q.uk   <- 0.005

est.start.uk <- idx_to_pos(ukitaly_cal, "2020-02-25")
est.end.uk   <- idx_to_pos(ukitaly_cal, "2020-04-01")
Yuk          <- idx_series(idx_values(ukitaly_idx)[, "UK"], start = ukitaly_idx$start)
```

```{r fig-uk-italy, fig.cap = "Figure 16: Daily COVID cases in UK and Italy."}
mod.ukit.plot <- SSModelLeadingIndicator$new(ukitaly_idx, n.lag = 4, calendar = ukitaly_cal)
plot(
  mod.ukit.plot,
  title = "Daily COVID cases in UK and Italy",
  series.name.lead = "Italy",
  series.name.target = "UK",
  take.log = FALSE
)
```

The UK-only dynamic Gompertz model is estimated on UK daily cases alone,
and the leading-indicator model uses Italy as the lead series with a
14-day lag, motivated by the observed timing of the first-wave evolution
in Italy relative to the UK. Both models are estimated with weekly
seasonality (`sea.period = 7`, the package default for daily data),
specified explicitly here so that the two specifications are directly
comparable; a fuller factorial comparison across seasonal and other
tuning choices is beyond the scope of this vignette.

```{r uk-gompertz}
mod.uk.gomp <- SSModelDynamicGompertz$new(
  Y = Yuk, q = q.uk, sea.period = 7,
  start = est.start.uk, end = est.end.uk,
  calendar = ukitaly_cal
)
res.uk.gomp <- estimate(mod.uk.gomp)
```

```{r uk-lead}
# NOTE: n.lag.uk = 14 here is illustrative (motivated by observed
# first-wave timing) and precedes the systematic lag comparison below,
# which selects its lag from data using disjoint selection/reporting
# folds. This fixed value is not justified by that comparison.
n.lag.uk <- 14
mod.uk.lead <- SSModelLeadingIndicator$new(
  Y = ukitaly_idx, n.lag = n.lag.uk, sea.period = 7,
  start = est.start.uk, end = est.end.uk,
  calendar = ukitaly_cal
)
res.uk.lead <- estimate(mod.uk.lead)
```

```{r fig-uk-compare, fig.cap = "Figure 17: Comparison of dynamic Gompertz and leading-indicator forecasts, UK daily cases."}
tsgc::plot_compare_forecast(
  list(res.uk.gomp, res.uk.lead),
  actual = Yuk,
  n.ahead = n.forecasts.uk
)
```

```{r fig-uk-holdout, fig.cap = "Figure 18: Accuracy of the leading-indicator forecast, UK daily cases."}
tsgc::plot_holdout(
  res.uk.lead, Y = ukitaly_idx, n.ahead = n.forecasts.uk,
  title = "Accuracy: leading-indicator forecast of UK cases",
  series.name = "UK cases"
)
```

In this first-wave window, the leading-indicator model gives
substantially smaller holdout errors than the UK-only dynamic Gompertz
model.

### Selecting the lag by rolling cross-validation

The baseline lag $k=14$ is motivated by the observed first-wave timing.
To avoid selecting the lag purely by inspection, the package provides a
rolling cross-validation utility, `cross_val()`, which evaluates
candidate specifications over a rolling origin and reports a chosen
accuracy criterion. Because `SSModelLeadingIndicator` uses a single
alignment lag per fitted model, this should be read as model selection
over alternative alignments, not as estimation of a distributed-lag
response.

```{r cv-models}
# The CV loop below tests n.lag up to 21. SSModelLeadingIndicator$estimate()
# shifts the leading-indicator columns forward by n.lag positions before
# differencing, which shrinks the range of usable (non-missing) rows by
# roughly n.lag + 1 positions from the start of the window. The shared
# est.start.uk/est.end.uk window (36 days) leaves too little room for the
# SELECTION block once disjointness from REPORTING is also enforced: the
# highest lags (16, 18, 21) then have no valid estimation positions left in
# the SELECTION block and estimate() errors with "start is after end". A
# separate, wider est.end.cv.uk window is used for the CV section only, so
# the SELECTION block has enough margin for n.lag = 21; est.end.uk itself is
# left untouched since other models earlier in this vignette are estimated
# on that shorter window intentionally.
est.end.cv.uk <- idx_to_pos(ukitaly_cal, "2020-04-15")

cv_models <- list()

cv_models[["Naive_last_value"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, q = 0, sea.period = 7,
  start = est.start.uk, end = est.end.cv.uk,
  calendar = ukitaly_cal
)

cv_models[["RW_growth"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, q = q.uk, sea.period = 0,
  start = est.start.uk, end = est.end.cv.uk,
  calendar = ukitaly_cal
)

cv_models[["Vanilla_q"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, q = q.uk, sea.period = 7,
  start = est.start.uk, end = est.end.cv.uk,
  calendar = ukitaly_cal
)

cv_models[["Vanilla_ar1"]] <- SSModelDynamicGompertz$new(
  Y = Yuk, sea.period = 7,
  start = est.start.uk, end = est.end.cv.uk, ar1 = TRUE,
  calendar = ukitaly_cal
)

for (i in c(7, 10, 14, 16, 18, 21)) {
  cv_models[[paste0("Lag", i)]] <- SSModelLeadingIndicator$new(
    Y = ukitaly_idx, sea.period = 7,
    start = est.start.uk, end = est.end.cv.uk, n.lag = i,
    calendar = ukitaly_cal
  )
}
```

`cross_val()` is run twice with **disjoint** sets of forecast origins,
spaced by the forecast horizon itself so evaluation windows do not
overlap: an earlier SELECTION block used only to choose the lag, and a
later REPORTING block used only to score the selected lag against the
benchmarks. Using the same folds for both selection and reporting
would overstate the selected model's accuracy.

```{r cv-run}
# cross_val() steps origins FORWARD from est.end by gap each iteration,
# not backward. cv_models above is now built on the wider est.end.cv.uk
# window (2020-04-15), not est.end.uk (2020-04-01), specifically so the
# SELECTION block has enough usable positions left for n.lag = 21 after
# SSModelLeadingIndicator's lag-shift + differencing (see the comment in
# the cv-models chunk). Both the SELECTION and REPORTING blocks are laid
# out to forward-step within that wider window, with REPORTING's forecast
# horizon landing exactly at est.end.cv.uk and SELECTION's last forecast
# horizon finishing strictly before REPORTING's origin (not merely before
# REPORTING's forecast horizon), so the two blocks share no origin and no
# forecast window - verified by the stopifnot() below, not asserted by
# comment alone.
n.ahead.cv  <- 5
gap.cv      <- n.ahead.cv   # non-overlapping horizons
n.select.cv <- 2
n.report.cv <- 1

report.end.cv <- est.end.cv.uk - (n.report.cv - 1) * gap.cv - n.ahead.cv
# The last SELECTION origin is select.end.cv + (n.select.cv-1)*gap.cv,
# with its forecast horizon reaching + n.ahead.cv further. Requiring
# that to finish strictly before the REPORTING origin (report.end.cv)
# means subtracting n.select.cv * gap.cv (not a single gap.cv) below -
# subtracting only one gap.cv left the last SELECTION fold's forecast
# horizon landing exactly ON report.end.cv (touching, not disjoint).
select.end.cv <- report.end.cv - n.select.cv * gap.cv - n.ahead.cv
stopifnot(select.end.cv > est.start.uk)

# Explicit disjointness check: the last SELECTION fold's forecast
# horizon must end strictly before the REPORTING block's origin.
last_selection_origin      <- select.end.cv + (n.select.cv - 1) * gap.cv
last_selection_horizon_end <- last_selection_origin + n.ahead.cv
stopifnot(last_selection_horizon_end < report.end.cv)

# Feasibility check for the highest lag tested: SSModelLeadingIndicator's
# estimate() needs at least n.lag + 2 usable positions from est.start.uk
# before the SELECTION block's origin (n.lag positions consumed by the
# forward lag-shift, plus at least 1-2 more lost to differencing/NA
# trimming). This turns an opaque "start is after end" error from deep
# inside estimate() into an explicit, early, diagnosable failure if the
# window is ever narrowed again.
max_lag_cv <- 21
stopifnot(select.end.cv - est.start.uk > max_lag_cv + 1)

cv_selection <- cross_val(
  Y = ukitaly_idx,
  model_list = cv_models,
  est.end = select.end.cv,
  criterion = "smape",
  n.ahead = n.ahead.cv,
  n.estimate = n.select.cv,
  gap = gap.cv
)
cv_selection

lag_rows <- cv_selection[grepl("^Lag", cv_selection$Model), ]
lag_rows$mean_smape <- rowMeans(lag_rows[, -1, drop = FALSE])
selected_lag_name <- lag_rows$Model[which.min(lag_rows$mean_smape)]
selected_lag_name

cv_report_models <- cv_models[c(
  "Naive_last_value", "RW_growth", "Vanilla_q", "Vanilla_ar1", selected_lag_name
)]
cv_reporting <- cross_val(
  Y = ukitaly_idx,
  model_list = cv_report_models,
  est.end = report.end.cv,
  criterion = "smape",
  n.ahead = n.ahead.cv,
  n.estimate = n.report.cv,
  gap = gap.cv
)
cv_reporting
```

For each candidate model, `cross_val()` evaluates forecasting
performance over a rolling-origin scheme, re-estimating the model at
each origin and recording the chosen accuracy criterion over the
forecast horizon. `Naive_last_value` and `RW_growth` are naive
benchmarks; the `Vanilla` rows are UK-only dynamic Gompertz benchmarks
(fixed $q_\zeta = 0.005$ and the AR(1) extension respectively); the
remaining rows are Italy-to-UK leading-indicator models at different
lags. `cv_reporting` is the accuracy figure that should be quoted for
the selected lag, since it comes from origins that played no role in
choosing it; `cv_selection` should not itself be read as an accuracy
claim about the winning lag. Note that this comparison still uses only
a handful of origins from one epidemic wave and realised (not
archived) values throughout, so it remains illustrative of the
`cross_val()` API rather than a general validation of any one lag
choice.

## Extensions: non-epidemiological applications and alternative data frequencies

The package is not restricted to daily epidemic data. The same
state-space machinery can be applied to quarterly, monthly, or annual
data when the underlying series is cumulative and exhibits saturation;
what changes with the sampling frequency is only the interpretation of
the integer position, forecast horizon, lag length, and seasonal period,
not the underlying estimation procedure. For example, the quarterly
Nintendo sales data can be used with a seasonal period of 4 and a lagged
leading indicator, while monthly trading-app downloads can be treated in
the same way with a seasonal period of 12. In both cases, `idx_calendar`
carries the quarterly/monthly step size and unit, and `idx_to_pos()`
still translates calendar dates to the integer positions that
`start`/`end`/`n.lag` require.

```{r freq-quarterly}
data(nintendo_sales, package = "tsgc")

y_q_xts <- nintendo_sales[, c("wii", "switch_all")]

nintendo_idx <- idx_series(zoo::coredata(y_q_xts), start = 1L)
nintendo_cal <- idx_calendar(
  anchor = as.Date(zoo::index(y_q_xts)[1]),
  anchor_pos = 1L,
  amount = 1, unit = "quarters",
  posixct = TRUE
)

n.lag.q <- idx_to_pos(nintendo_cal, "2017-01-01") - idx_to_pos(nintendo_cal, "2006-10-01")

mod.q <- SSModelLeadingIndicator$new(
  Y = nintendo_idx,
  sea.period = 4,
  n.lag = n.lag.q,
  start = idx_to_pos(nintendo_cal, "2017-01-01"),
  end = idx_to_pos(nintendo_cal, "2019-10-01"),
  calendar = nintendo_cal
)
res.q <- estimate(mod.q)
```

```{r fig-freq-q, fig.cap = "Figure 19: Log forecasts of quarterly Switch sales, using Wii sales as a leading indicator."}
tsgc::plot_log_forecast(
  res.q, Y = nintendo_idx, n.ahead = 8,
  title = "Log forecasts of Switch sales"
)
```

```{r freq-monthly}
data(etrading_apps, package = "tsgc")

y_m_xts <- etrading_apps[, c("DEGIRO", "AvaTrade")]

etrading_idx <- idx_series(zoo::coredata(y_m_xts), start = 1L)
etrading_cal <- idx_calendar(
  anchor = as.Date(zoo::index(y_m_xts)[1]),
  anchor_pos = 1L,
  amount = 1, unit = "months",
  posixct = TRUE
)

n.lag.m <- idx_to_pos(etrading_cal, "2017-07-01") - idx_to_pos(etrading_cal, "2017-01-01")

mod.m <- SSModelLeadingIndicator$new(
  Y = etrading_idx,
  sea.period = 12,
  n.lag = n.lag.m,
  start = idx_to_pos(etrading_cal, "2017-07-01"),
  end = idx_to_pos(etrading_cal, "2021-02-01"),
  calendar = etrading_cal
)
res.m <- estimate(mod.m)
```

```{r fig-freq-m, fig.cap = "Figure 20: Forecasts of monthly AvaTrade downloads, using DEGIRO downloads as a leading indicator."}
tsgc::plot_forecast(
  res.m, n.ahead = 4,
  title = "AvaTrade monthly downloads",
  series.name = "Downloads"
)
```

Both examples use the same workflow as the daily illustrations above:
select an estimation window, fit the model, choose a forecast horizon,
and evaluate the resulting forecasts. No change to the state-space
specification or estimation procedure is required when moving between
frequencies.

# Conclusions

The `tsgc` package is based on a dynamic Gompertz curve model for the
log of the growth rate of cumulative cases in an epidemic, with seasonal
terms that capture day-of-the-week effects. The estimation is carried
out using `KFAS`, a package for state-space modeling in `R`.

The Kalman filter is used to estimate the state vector at each time
point. The filter is initialized using a diffuse prior for the initial
state vector. We allow the signal-to-noise ratio to either be estimated
or fixed at some value based on experience and judgment. Future
observations are forecast using predictive recursions.

Epidemics are often characterized by multiple waves. A natural problem
in this context is that there is very little data pertinent to the new
wave available in its initial stages. The package employs a
reinitialization method using priors in a way that allows data from
before the beginning of the new wave to be used in estimation. The same
framework can be extended to a leading-indicator formulation for
outcomes that are preceded by an observable correlate, and it can be
applied to data measured at different frequencies once the relevant
seasonal period and lag structure are chosen.

The package is demonstrated using COVID-19 data from South Africa and
England, but it can be used to model and forecast any time series
variable where a growth curve-like trajectory is expected. Examples
might include sales of a new product, innovation adoption or website
traffic. The approach is phenomenological rather than mechanistic: it is
designed for short-horizon forecasting and exploratory analysis rather
than for causal identification or structural epidemiological inference.
The forecast bands are state-implied summaries of the latent-state
uncertainty and should not be interpreted as calibrated prediction or
confidence intervals in the usual sense. In addition, the implied
lead–lag relationship between a leading indicator and a target series
can break down over time, so the model should be monitored and
re-estimated as conditions change.

# Acknowledgements {.unnumbered}

We thank The Cambridge Centre for Health Leadership & Enterprise,
University of Cambridge Judge Business School, and Public Health
England/UK Health Security Agency for generous support. We are indebted
to Thilo Klein and Stefan Scholtes for constructive comments. Andrew
Harvey's work was carried out as part of the University of Cambridge
Keynes Fund project \`Forecasting and Policy in Environmental
Econometrics'.

# References {.unnumbered}

::: {#refs}
:::

# Appendix: Incorporating seasonal terms into the state space model

When we add a seasonal term to the model, the observation equation in
the dynamic Gompertz curve (2) becomes
$$\ln g_{t}=\delta_{t}+\nu_t+\varepsilon_{t},\;\;\;\;\;\varepsilon_{t}\sim
NID(0,\sigma_{\varepsilon }^{2}),\;\;\;\;\;t=s,...,T,$$ where $\nu_t$ is
the seasonal component, $\delta_t$ remains defined by (3), and
$\varepsilon_t$ remains the iid Normal disturbance.

There are two options for specifying the evolution of the seasonal
component. We can either use a trigonometric seasonal or a dummy
variable seasonal. In our application, we use a trigonometric seasonal,
although the two specifications are closely related. Indeed,
@Proietti2000 shows that, under certain conditions, the two approaches
are equivalent.

In the trigonometric seasonal approach, the seasonal pattern is captured
by a set of trigonometric terms at the seasonal frequencies
$\lambda_j = \frac{2 \pi j}{s}$ for $j=1, \ldots s^*$, where
$s^*=\frac{s}{2}$ if $s$, the periodicity of the seasonal effect, is
even, and $s^*=\frac{s-1}{2}$ if $s$ is odd @Durbin2012. Our
applications use daily data and we set $s=7$ to capture day-of-the-week
effects.

Letting $\nu_{j,t}$ be the effect of season $j$ at time $t$, the
seasonal terms evolve according to

\begin{align}
\nu_{t} &= \sum_{j=1}^{s^{*}} \nu_{j,t}, & (13)
\end{align}

where \begin{align}
\nu_{j,t} &= \nu_{j,t-1} \cos \lambda_j + \nu^*_{j,t-1} \sin \lambda_j + \omega_{j,t} & (14)\\
\nu^*_{j,t} &= -\nu_{j,t-1} \sin \lambda_j + \nu^*_{j,t-1} \cos \lambda_j  + \omega^*_{j,t}, \;\;\;\;\; j = 1, \ldots, s^*, & (15)
\end{align} and $\omega_{j,t}$ and $\omega^*_{j,t}$ are mutually
independent, iid $N(0,\sigma^2_{\omega})$ variables.

When reinitializing the model with seasonal terms, $P^r_1$ is a
block-diagonal matrix based on $P_{r+1}$ which sets the covariances
between $(\delta_t,\gamma_t)'$ and
$(\nu_{1,t},\nu_{2,t},\ldots,\nu_{s^*,t})'$ to be zero. The covariance
between $\delta_t$ and $\gamma_t$, as well as the covariances between
$\nu_{1,t},\nu_{2,t},\ldots,\nu_{s^*,t}$, are permitted to be non-zero
and come directly from $P_{r+1}$.

# Appendix: Validating rendered figures

Zero-byte or undecodable figures can pass a ZIP/Pandoc build silently,
since neither checks that a referenced image actually decodes. The
chunk below is intended to be run once after `rmarkdown::render()` on
this vignette (not during normal knitting, since the output files do
not yet exist while the document itself is being produced), and checks
every figure written to the vignette's `_files` figure directory:
existence, non-zero size, successful PNG decode, and a plausible
minimum pixel dimension.

```{r validate-figures, eval = FALSE}
validate_rendered_figures <- function(fig_dir, min_dim = 50) {
  pngs <- list.files(fig_dir, pattern = "\\.png$", full.names = TRUE, recursive = TRUE)
  bad <- character(0)
  for (f in pngs) {
    sz <- file.info(f)$size
    ok <- !is.na(sz) && sz > 0
    if (ok) {
      dims <- tryCatch(dim(png::readPNG(f))[1:2], error = function(e) NULL)
      ok <- !is.null(dims) && all(dims >= min_dim)
    }
    if (!ok) bad <- c(bad, f)
  }
  if (length(bad) > 0) {
    stop("Empty or undecodable figure(s): ", paste(basename(bad), collapse = ", "))
  }
  message("All ", length(pngs), " rendered figures validated.")
}
# validate_rendered_figures("tsgc_vignette_files/figure-html")
```

For models that include external regressors, forecasts are conditional
on future values of those regressors. In this version of the package
those future values are supplied as an `idx_series`, set directly on
the fields of a fitted `FilterResults`/`FilterResultsLI` object
(`xpred.new` for `SSModelDynamicGompertz`, or `xpred_lead.new`/
`xpred_targ.new` for `SSModelLeadingIndicator`), rather than through the
`supply_xpred.new()` helper used previously. If the future regressor
values originate in a CSV file with a `Date` column, the same
`xts_to_idx()` conversion used elsewhere in this vignette applies: read
the CSV, convert the date column, coerce to `xts`, and then convert to
`idx_series` using the model's existing calendar so that the resulting
positions line up correctly. The following example illustrates this
using an inline text block for reproducibility; in practice the data
would normally be read directly from a CSV file.

The examples below assume a Gompertz model has already been fit with a
weather regressor supplied via `xpred`, producing a `FilterResults`
object called `res_weather`. This is fit explicitly here (using its own
pinned estimation window, rather than the ambient `estimation.pos.end`
reassigned above), so the CSV examples that follow are self-contained.

```{r fit-res-weather}
data(gauteng_weather_2021, package = "tsgc")
gauteng_weather_idx <- xts_to_idx(
  gauteng_weather_2021[, c(1, 3)],
  start.pos = idx_to_pos(cal, zoo::index(gauteng_weather_2021)[1])
)$series

weather.pos.start <- estimation.pos.start
weather.pos.end   <- idx_to_pos(cal, "2021-05-03")

gauteng_weather_est <- get_timeframe(
  gauteng_weather_idx, weather.pos.start, weather.pos.end
)

model_weather <- SSModelDynamicGompertz(
  Y = Y, xpred = gauteng_weather_est, q = q,
  start = weather.pos.start, end = weather.pos.end,
  calendar = cal
)
res_weather <- estimate(model_weather)
```

```{r xpred-csv-example, eval = F}
# This inline CSV block demonstrates the expected format of a file
# containing future values of the external regressors.
txt <- "
Date,windspd_mtrs_p_sec,temperature_C
2021-05-04,2.03,14.61
2021-05-05,1.54,15.51
2021-05-06,1.94,16.42
2021-05-07,2.38,15.54
2021-05-08,2.57,14.18
2021-05-09,2.65,13.55
2021-05-10,2.19,14.84
2021-05-11,2.08,15.55
2021-05-12,2.07,15.97
2021-05-13,1.92,15.86
2021-05-14,1.85,15.79
2021-05-15,2.86,15.57
2021-05-16,3.46,16.42
2021-05-17,2.39,13.53
"

# Read the demonstration CSV text into a data frame.
gauteng_weather_future_csv <- read.csv(
  text = txt,
  stringsAsFactors = FALSE
)

gauteng_weather_future_csv$Date <-
  as.Date(gauteng_weather_future_csv$Date)

# Convert the data frame to xts first...
gauteng_weather_future_xts <- xts::xts(
  gauteng_weather_future_csv[, -1],
  order.by = gauteng_weather_future_csv$Date
)

# ...then to an idx_series, anchored on the fitted model's own calendar
# so that the resulting positions line up with res_weather$calendar.
gauteng_weather_future_idx <- xts_to_idx(
  gauteng_weather_future_xts,
  start.pos = idx_to_pos(res_weather$calendar,
                         zoo::index(gauteng_weather_future_xts)[1])
)$series

# Supply the future regressors directly on the fitted results object.
res_weather$xpred.new <- gauteng_weather_future_idx
```

In practice, users would normally read the future regressor values
directly from a CSV file rather than embedding the data in the code. The
CSV file should have the same structure as shown above, with a `Date`
column followed by one column for each external regressor.

```{r xpred-csv-file, eval = F}
# Read future regressors from a CSV file.
gauteng_weather_future_csv <- read.csv(
  "gauteng_weather_future.csv",
  stringsAsFactors = FALSE
)

gauteng_weather_future_csv$Date <-
  as.Date(gauteng_weather_future_csv$Date)

# Convert to xts, then to idx_series on the model's calendar.
gauteng_weather_future_xts <- xts::xts(
  gauteng_weather_future_csv[, -1],
  order.by = gauteng_weather_future_csv$Date
)
gauteng_weather_future_idx <- xts_to_idx(
  gauteng_weather_future_xts,
  start.pos = idx_to_pos(res_weather$calendar,
                         zoo::index(gauteng_weather_future_xts)[1])
)$series

# Supply the future regressors directly on the fitted results object.
res_weather$xpred.new <- gauteng_weather_future_idx
```

Once the future regressors have been supplied, subsequent forecasts from
the fitted model are conditional on these external regressor values.

While this example is specifically for exogenous predictors supplied to
a `SSModelDynamicGompertz` fit, the same `xts_to_idx()`/field-assignment
pattern extends to `xpred_lead.new` and `xpred_targ.new` on a
`SSModelLeadingIndicator` fit, as demonstrated for the England
cases-hospitalisations example above.

# Appendix: Controlling the plot x-axis with `idx_axis_opts()`

Every plotting function in this section so far has left its `axis`
argument unset, which defaults to `mode = "auto"`: real calendar dates
if the calendar has `posixct = TRUE` (as in almost every example
above), otherwise plain integer positions. `idx_axis_opts()` lets you
override this per plot, and applies uniformly across every
`plot.*`/`plot_*` function in the package. To keep the comparison
clear, this section reuses a single plot and simply varies `axis`.

```{r axis-showcase-setup}
# res/model/estimation.pos.end are re-used, generically-named objects
# earlier in this vignette: by this point in the document, `res` has
# already been reassigned by the Reinitialization section above (to
# the longer-window, 25-June re-estimate), so it no longer refers to
# the original fixed-q Figure 3 model (estimation ending 3 May 2021).
# Re-fit that original specification explicitly under its own name so
# this section's figures do not silently depend on what the ambient
# `res` object happens to hold by this point in the document.
res.axis.demo <- estimate(
  SSModelDynamicGompertz$new(Y = Y, q = q, sea.period = 7,
                             start = estimation.pos.start,
                             end = idx_to_pos(cal, "2021-05-03"),
                             calendar = cal)
)

showcase_axis_plot <- function(axis = NULL) {
  tsgc::plot_forecast(
    res = res.axis.demo,
    n.ahead = n.forecasts,
    confidence.level = confidence.level,
    plt.start = tail(res.axis.demo$index, 1) - plt.length,
    series.name = "cases",
    axis = axis
  )
}
```

## `mode = "date"` (the default here, since `gauteng_cal$posixct = TRUE`)

Real calendar dates on the x-axis - this is what `axis = NULL` already
produced in Figure 3.

```{r fig-axis-date, fig.cap = "Real calendar dates (mode = \"date\")."}
showcase_axis_plot(idx_axis_opts(mode = "date"))
```

## `mode = "position"`

The raw integer `idx_series` position, ignoring the calendar entirely.

```{r fig-axis-position, fig.cap = "Raw integer positions (mode = \"position\")."}
showcase_axis_plot(idx_axis_opts(mode = "position"))
```

## `mode = "steps"`

Steps from the calendar's anchor position
(`position - anchor_pos`). Since `gauteng_cal` has an `anchor_name`
set (`"first recorded case"`, added earlier when the calendar was
built), the axis label uses it automatically.

```{r fig-axis-steps, fig.cap = "Steps from the anchor (mode = \"steps\")."}
showcase_axis_plot(idx_axis_opts(mode = "steps"))
```

## `mode = "time_since"`

The pattern-weighted calendar offset from the anchor, expressed in
`calendar$unit`s rather than a raw step count - here "days since first
recorded case". For a daily series with no gaps this is numerically
identical to `"steps"`, but it diverges for series with an irregular
`pattern` (e.g. business days) or a non-day `unit` (e.g. quarters,
months), and it also works for `posixct = TRUE` calendars where it can
be more readable than a real date.

```{r fig-axis-time-since, fig.cap = "Time since the anchor, in calendar units (mode = \"time_since\")."}
showcase_axis_plot(idx_axis_opts(mode = "time_since"))
```

## Adding an info box

Setting `info_box = TRUE` (on any of the modes above) appends a
caption summarising the calendar: the anchor (and its name, if set),
the step size/unit, and the pattern. `pattern_n` truncates a long
pattern to its first `n` values; by default the full pattern is
shown.

```{r fig-axis-infobox, fig.cap = "Steps from anchor, with an info box summarising the calendar."}
showcase_axis_plot(idx_axis_opts(mode = "steps", info_box = TRUE))
```

If the info box's pattern were long (e.g. a business-day calendar with
weekends skipped), `pattern_n` keeps the caption readable:

```{r fig-axis-infobox-truncated, fig.cap = "Info box with pattern truncated to its first 3 values via pattern_n."}
showcase_axis_plot(idx_axis_opts(mode = "steps", info_box = TRUE, pattern_n = 3))
```
# Appendix: Compound calendar steps with `idx_step()` and `idx_calendar_step()`

Every calendar used elsewhere in this vignette is built with the
primary `idx_calendar()` constructor, whose step size is a single
`amount`/`unit` pair (e.g. `amount = 1, unit = "days"`, or `amount = 1,
unit = "quarters"`). This covers the common case where the step
between consecutive `idx_series` positions is a whole multiple of one
calendar unit. Some series, however, are genuinely spaced by a
combination of units that a single `amount`/`unit` pair cannot
express - for example, a reporting system that timestamps quarterly
figures a few seconds after the quarter boundary, so that the step is
"one quarter plus three seconds". `idx_step()` builds this kind of
compound step, and `idx_calendar_step()` is the `idx_calendar`
constructor that takes one.

```{r appendix-idx-step}
qtr_step <- idx_step(quarters = 1, seconds = 3)
qtr_step

cal_step <- idx_calendar_step(
  anchor = as.POSIXct("2024-01-01 00:00:03", tz = "UTC"),
  step = qtr_step,
  posixct = TRUE
)
idx_to_date(cal_step, 1:4)
```

`idx_to_pos()` inverts this exactly as it does for the primary
constructor. A compound step has no single `amount` to divide by, so
rather than the closed-form calculation used for `idx_calendar()`
calendars, the position is instead located by a binary search on the
number of steps from the anchor:

```{r appendix-idx-step-inverse}
idx_to_pos(cal_step, idx_to_date(cal_step, 4))
```

Any component of an `idx_step` (`years`, `quarters`, `months`,
`weeks`, `days`, `hours`, `minutes`, `seconds`) left at its default of
`0` is simply omitted from the step, so `idx_step(days = 1)` is
equivalent to the `amount = 1, unit = "days"` pairs used throughout
the main text. `idx_step_add()` is the lower-level function that
applies a given number of whole steps to an anchor date directly, and
is what an `idx_calendar_step()`-built calendar uses internally to
convert positions to dates; it is not usually called directly by
users of the package; it is available for cases where a step needs
to be applied outside of an `idx_calendar` altogether.

# Appendix: Heterogeneous repeating cycles with `multi_step_pattern()` and `idx_calendar_multi_step()`

The `pattern` argument of `idx_calendar()`/`idx_calendar_step()`
handles a repeating cycle of *different multiples* of one step - e.g.
a business-day calendar, where four single-day steps (Mon-Thu) are
followed by a three-day step over the weekend. That still assumes
every step in the cycle is the same *kind* of step, just scaled
differently. Some data instead cycle through genuinely different
kinds of steps - for example, "three days, three days, then one
month, repeating". `multi_step_pattern()` describes this kind of
cycle as a sequence of `idx_step` objects, and
`idx_calendar_multi_step()` is the `idx_calendar` constructor that
uses it.

```{r appendix-multi-step}
cal_multi <- idx_calendar_multi_step(
  anchor = as.Date("2024-01-01"),
  multi_step = multi_step_pattern(
    idx_step(days = 3), idx_step(days = 3), idx_step(months = 1)
  ),
  posixct = TRUE
)
idx_to_date(cal_multi, 1:6)
```

Because the slots in a `multi_step_pattern` can mix calendar-relative
(e.g. months) and fixed-duration (e.g. days) components, there is no
closed-form way to collapse the whole cycle into a single numeric
offset. `idx_to_date()` instead walks one slot at a time from the
anchor position, and `idx_to_pos()`'s inverse walks the same cycle in
reverse:

```{r appendix-multi-step-inverse}
idx_to_pos(cal_multi, "2024-02-07")
```

This makes `idx_calendar_multi_step()` calendars more expensive to
convert than the other two constructors - each position takes time
proportional to its distance from the anchor, rather than constant
time - so it is best reserved for genuinely irregular step sequences
that the `pattern` argument of the other two constructors cannot
express. The `"time_since"` axis mode described in the previous
appendix is also unavailable for this calendar: a `multi_step_pattern`
has no single `amount` to express an offset in, so `mode = "steps"`
or `mode = "date"` should be used instead.

# Appendix: Non-calendar `idx_calendar` anchors and `idx_offset_to_pos()`

Not every `idx_series` needs a genuine calendar interpretation. An
`idx_calendar`'s `anchor` can be any single reference point, not just
a `Date`/`POSIXct` - for example, a plain numeric count of
picoseconds since the start of an experiment. `idx_offset_to_pos()`
is the inverse operation for this case: given a value already
expressed in the anchor's own units, it returns the integer
`idx_series` position that corresponds to it - the counterpart of
`idx_to_pos()`, which instead expects a calendar date.

```{r appendix-offset-to-pos}
cal_ps <- idx_calendar(anchor = 0, anchor_pos = 1L, amount = 2.5,
                       unit = "picoseconds")
idx_offset_to_pos(cal_ps, 12.5)
```

None of the examples elsewhere in this vignette need
`idx_offset_to_pos()`, since every series used above has a genuine
`Date`/`POSIXct` calendar and so uses `idx_to_pos()` instead; it is
included here for completeness, for applications - e.g. raw
instrumentation data with no calendar meaning - where `idx_series`
positions are still worth translating to and from the anchor's own
numeric scale.
